### 2012-01-11 rnaseq_paper.tex
add a latex source of a paper  and related files

% Template for PLoS
% Version 1.0 January 2009
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

\usepackage{multirow}

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{mRNA-Seq Assembly Discovers Many Novel Splice Variants}
}
% Insert Author names, affiliations and corresponding author email.
\\
Likit Preeyanon$^{1}$, 
Jerry B. Dodgson$^{1}$
Hans Cheng$^{2}$, 
C. Titus Brown$^{3,1 \ast}$
\\
\bf{1} Microbiology and Molecular Genetics, Michigan State University, East Lansing, MI, USA.
\\
\bf{2} Avian Disease and Oncology Laboratory, East Lansing, MI, USA.
\\
\bf{3} Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, USA.
\\
$\ast$ E-mail: ctb@msu.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

\section*{Introduction}

% Results and Discussion can be combined.
\section*{Results}

\subsection*{Transcripts from Global and Local Assembly}
\subsubsection*{Different Isoforms Detected by Different k-mers}

Without high quality annotations, it is necessary that we construct gene models containing putative isoforms to identify alternative splicing in our samples.
We used Velvet\cite{Zerbino:2008vu} and Oases assembler to construct transcripts from short reads.
We ran Velvet using default settings except a hash length or k-mer parameter, which has no default values.
To find an appropriate value for hash length for our datasets, we experimented on a wide range of k-mer lengths from 23 to 35.
After comparing transcripts from all k-mers, we found that some isoforms were only detected in some particular k-mers (Fig.~\ref{figure4}).
Therefore, instead of choosing one k-mer, we ran Velvet assembler with multiple k-mers and combined the results to increase sensitivity of isoform detection.

\subsubsection*{Local Assembly Enhances Isoform Detection}

Performing assembly with multiple k-mers is time and memory consuming.
To facilitate this, we partitioned reads into groups based on their alignments to the chicken genome.
We refer to this method as a local assembly and a conventional assembly method as a global assembly in this study (see materials and methods).
The advantages of the local assembly method are that each group of reads can be assembled using small amount of memory and it can be assembled on different computers.
Furthermore, we found that local assembly could identify unique alternative splicing from global assembly.
Therefore, we performed both global and local assembly with multiple k-mers (23-35 and 15-39 respectively) to obtain all possible alternative splicing detected from both methods.
Figure~\ref{figure5} shows an example of isoforms detected with different k-mers in two assembly methods.

Using global and local assembly methods, we obtained a different number of contigs.
Table 1 shows a number of contigs from four datasets and unique contigs from global and local assemblies.
As expected, some sequences from local assembly (0.5--0.7\%) are not detected in global assembly.
Homology search in mouse proteins showed that 4--5.6\% of unique regions from local assembly are genuine protein coding sequences (Table 2).
This suggests that some isoforms will be missing in the gene models based solely on global assembly.
In addition, a significant number of transcripts from global assembly are unique, which indicates that approximately 4.8\% of coding sequences are missing in a current chicken genome.

\subsection*{Predicted Gene Models from Assembly}

All transcripts were assembled by Gimme to construct gene models.
Total of 27,448 multi-exon genes were obtained from global assembly and 22,434 genes from local assembly.
However, the number of genes obtained from global + local assembly is only 32,854.
This indicates that some transcripts from both methods were common or merged together.
As expected, the number of genes and isoforms increased significantly after combining global and local assembly together due to the increased number of splice junctions (Table 3).

\subsection*{Fragmented transcripts}
The number of genes from our pipeline is unexpectedly high due to a large number of two-exon genes. As shown in Table 3, the number of two-exon genes is high (28-34\% in models derived from assembly), compared to the number of annotated two-exon genes in reference genes (6\%).
The majority of two-exon genes are shorter than 500bp as shown in Fig.~\ref{figure7}.
Homology search against mouse proteins (using blastp) showed that 60\% of two-exon genes that matched mouse proteins are shorter than 500bp (Fig.~\ref{figure8}).
The results suggest that these two-exon genes are indeed genuine coding sequences.
However, according to their small size, they are possibly parts of larger transcripts that were not completely assembled.
In our datasets, read coverage is lower at the $5'$ end due to the method used to converse mRNA to cDNA in sequencing.
A bias of read coverage toward $3'$ end and low expression level result in fragmented transcripts as shown in Fig.~\ref{figure9}.
Notably, the number of two-exon genes decreased when reference genes were incorporated to global + local assembly gene models (Table 3).
This indicates that some two-exon genes were merged to reference genes since they are part of larger transcripts.

\subsection*{Characteristics of RNA-Seq Derived Gene Models}
Gene models derived from our pipeline are comparable or, in some cases, more complete than gene models available on UCSC genome browser\cite{ucscGenomeBrowser}.
Fig.~\ref{figure10} shows a comparison of a predicted isoforms with gene models from reference genes, Ensembl[?], N-SCAN[?] and Genscan[?].
Our gene models have an advantage of including extended untranslated regions (UTRs), especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
RNA-Seq-derived gene models will facilitates the detection of variations within UTR regions, which may be involved in regulating the expression of the isoform [?] (Fig.~\ref{figure11}).

\subsection*{Comparison with Cufflinks and Trinity}
To evaluate an efficiency of our pipeline in detecting splice junctions,
we compared the number of splice junctions detected by our pipeline to those from Cufflinks\cite{Trapnell:2010kd} and Trinity\cite{Grabherr:2011jb}.
Cufflinks detected 99,958 (85\%) splice junctions that are also detected by our pipeline (Fig.~\ref{figure17}).
60,849 (37.8\%) are not detected by Cufflinks of which 29,207 (48\%) are supported by at least four spliced reads.

Reads from line 6 uninfected dataset that mapped to a chromosome 1 were assembled using Trinity with default settings.
Trinity found a total of 11,902 splice junctions; whereas, our pipeline detected 12,676 splice junctions.
1,395 splice junctions are not found by Trinity and 621 junctions are not detected by our pipeline.
Figure~\ref{figure19} summarizes a number of splice junctions in chromosome 1 detected by all three methods.
Figure~\ref{figure20} shows two splice variants that are detected by local assembly but not Trinity.

\subsection*{Validation of Gene Models}

The results show that our pipeline can be used to detect alternative splicing and construct gene models from a large set of transcripts from assembly.
However, many steps in our pipeline can introduce errors to the gene models.
We performed several methods to validate the gene models as described below.

\subsubsection*{Reads Mapped to Gene Models}

Single-end reads from the same datasets that we used to build the gene models were aligned to the DNA sequences derived from the gene models.
We anticipated that a large number of reads could align to the gene models.
We found that 66-73\% of reads mapped to the gene models (Table 4).
In addition, we mapped paired-end reads that we obtained later from the same samples to the gene models and found that more than 60\% of reads mapped to the gene models (Table 5).
The results suggest that the gene models are of high quality.

\subsubsection*{Conserved Reading Frame}
Alternative isoforms usually maintain a reading frame to be functional [?].
We hypothesized that errors in our pipeline that introduce false splice junctions will disrupt the reading frame.
Although, in nature, alternative-splicing events can disrupt the reading frame, we expected a significant number of isoforms to be translatable.

Transcripts from RNA-Seq assembly are supposed to contain UTRs and coding sequences. However, due to some factors mentioned previously, we could not obtain complete transcripts for all isoforms.
Thus, predicting the reading frame or coding sequence solely from sequence signals is not feasible.
We used ESTScan\cite{Iseli:1999vd}, a program that is designed to find coding sequences from expressed sequence tags (ESTs) and does not rely on sequence signals to translate isoforms to protein sequences.
To evaluate the result, we used reference mRNAs as a control.
However, since our gene models derived from the chicken genome sequence; therefore, we first aligned reference mRNAs to the genome to obtain DNA sequences to be used in the comparison.

ESTScan successfully translated 47,786 (80\%) isoforms from our gene models and 15,952 and 14,999 (97\%) sequences from reference mRNAs and models built from reference mRNAs.
We evaluated the completeness of coding sequence by comparing the number of nucleotides in each isoform to the number of amino acid encoding nucleotides.
We found that cumulative distributions of ratios from reference mRNAs before and after processed by our pipeline are highly similar.
This indicates that the pipeline did not introduce errors that disrupt the reading frame of the mRNAs.
The result from our gene models shows significant similarity to results from mRNAs, which suggests that a majority of isoforms detected by our pipeline is translatable.

\subsubsection*{Homologous sequences in mouse}
To further examine quality of protein sequences derived from our gene models, we searched for homologous sequences in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
We found 29,903 (52\%) isoforms (7,425 genes) match mouse proteins.
As shown in Fig.~\ref{figure14}, most genes and isoforms have a bit score/length ratio more than 1.0, which indicates a good agreement between chicken and mouse proteins.
The results illustrate that isoforms constructed by our pipeline are translatable and may be functional.
However, genes or isoforms with no match to mouse proteins are not necessarily artifacts.
They can be genes or isoforms that are only found in other vertebrates or novel genes.

\subsubsection*{Reads mapped to splice junctions}
We identified 160,807 splice junctions from the gene models.
To validate the splice junctions, we mapped reads to DNA sequences derived from gene models using Bowtie\cite{Langmead:2009fv}.
Because the mapping algorithm implemented in Bowtie does not consider splice sites, reads mapped across splice junctions confirm that the splice junctions are real.
Of all spliced junctions, 18,004 (11\%) junctions have no spliced reads and 11,815 (6.9\%) junctions have 1--3 spliced reads (see materials and methods).
This demonstrates that the method detects splice junctions with high specificity.
Splice junctions with fewer than four spliced reads, however, may not be considered false junctions because of a bias of read coverage toward $3'$ end in our datasets.
Reads are less abundant at the $5'$ end of transcripts; therefore, a relatively small number of spliced reads mapped to splice junctions near the $5'$ end.

\section*{Discussion}

In this study, we used multiple k-mers and a local assembly method to enhance splice variant detection.
Our new method detects splice variants that are not detected by a conventional assembly method.
We also showed that our method can be used with other programs (Cufflinks, Trinity, etc.) to increase sensitivity of splice variant detection.
It works well with single-ended reads and does not require more memory than the conventional assembly method.
The method, however, relies on a quality of a reference genome and a read mapping program.

We also presented Gimme, a program that assembles transcripts from different approaches based on alignments to a reference genome to construct gene models.
We used the pipeline to construct gene models from chicken mRNAs using global and local assembly to identify novel genes and isoforms in chicken.
Many novel genes and isoforms are found and validation results suggest that a majority of them are genuine and may be of biological importance.
The gene models also markedly improve existing annotations.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

\subsection*{Mapping reads to The Reference Genome and Gene Models}

Single and paired-end reads were mapped to chicken genome by Tophat\cite{Trapnell:2009dp} version 1.0.13 using default parameters without annotations.
All reads were mapped to DNA sequences derived from gene models by Bowtie version 0.12.7\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28, e=70, k=1)}.

\subsection*{Global and Local Assembly}

Reads from each dataset were first assembled separately in global assembly without using a reference genome.
In contrast, reads from each dataset were first mapped to the chicken genome using Tophat.
Then only reads that mapped to the genome were assembled by chromosomes in local assembly (Fig.~\ref{figure1}).
Global and local assembly was performed using Velvet version 1.0.17\cite{Zerbino:2008vu} with default parameters except for hash length (k-mer).
k-mer lengths ranging from 15 to 39 were used in local assembly and from 23 to 35 in global assembly to increase sensitivity of splice junction detection.
Lastly, transcripts from both methods were assembled by Oases version 0.1.18 (unpublished).

\subsection*{Gene Model Construction}

\subsubsection*{Overall Pipeline}

Transcripts of all datasets from local and global assembly were mapped to the chicken genome using GMAP version 2011-03-28.v3\cite{Wu:2005bl} \texttt{(-n 0)}.
Alignments and gaps from GMAP outputs are considered exons and introns respectively.
Optionally, data from other sources (ESTs, RefGenes, etc.) can be incorporated with transcripts from assembly to improve gene models.
All transcripts are then assembled using Gimme, a program that assembles transcripts based on their alignments to the reference genome.
An algorithm for assembling transcripts is described below.

\subsubsection*{Algorithm}

A gene model can be represented as a splice graph composed of exons as nodes and introns as edges.
However, transcripts of the same gene vary in size and structure depending on the expression level and a hash length number used in assembly.
Furthermore, incomplete exons and fragmented transcripts complicate the construction of a splice graph.
In this study, we developed an algorithm that handles incomplete exons and fragmented transcripts and constructs a maximum assembly of gene models.

The algorithm first builds an intron graph using introns as nodes  (Fig.~\ref{figure3}).
Each intron contains exons whose one of their splice sites perfectly match intron boundary.
Exons are considered incomplete and eliminated if they locate at the $3'$ or $5'$ end of the transcripts and they are not the largest exons (exon 3a and 3b in Fig.~\ref{figure3}).
Transcripts were then grouped into the same gene if they have at least one intron or exon in common (Fig.~\ref{figure3}).
Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing paths in the splice graph.
Two sets of putative transcripts can be constructed --- a maximum and a minimum set.
The maximum set contains all putative isoforms, which are derived from all paths in splice graphs.
In contrast, the minimum set contains minimum isoforms that represents all unique edges in splice graphs.

\subsection*{Open Reading Frame and Strand Prediction}

We employed ESTScan version 2.1\cite{Iseli:1999vd} to predict strand and ORFs for our gene models.
The matrix used for building Hidden Markov model was built from chicken reference cDNA sequences using tools from ESTScan.

\subsection*{Finding Unique Sequences Between Datasets}

To identify unique sequences from two datasets, a set of 20-mers is created for both datasets using khmer [?].
Then, 20-mers from a query dataset are compared with 20-mers from the target dataset.
The sequence is considered unique if more than 90\% of 20-mers in the query is unique.
Any unique region shorter than 100bp is ignored.

\subsection*{Sequence Homology Analysis}

Protein sequences translated from each isoform using ESTScan were searched against mouse reference sequences [?] by BLAST 2.2.25+\cite{Tatusova:1999tz}.
A bit score to a length ratio was calculated for each hit that had e-value $\le 10^{-20}$.
Only the highest ratio of all isoforms of a gene was used in the gene-wise analysis, whereas, all ratios were used in the isoform-wise analysis.

\subsection*{Spliced Reads Counting}

Reads from each dataset were mapped to transcripts from the gene models using Bowtie version 0.12.7 using default parameters.
Only reads that mapped across exon junctions from all datasets were counted using Samtools\cite{Li:2009vz} and Pysam [?]

\subsection*{Sequence Assembly using Trinity}
To be added.

\subsection*{Sequence Assembly using Cufflinks}
To be added.

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

Royal Thai Government Scholarship.

{\noindent}USDA funding.

%\section*{References}
% The bibtex filename
\bibliography{refs}

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure1.png}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
Transcripts are aligned to a chicken genome by GMAP. Gimme then constructs gene models based on alignments of transcripts.
}
\label{figure1}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure2.png}
\end{center}
\caption{
{\bf Local Assembly Pipeline.}
Reads are first mapped to a chicken genome.
Then only mapped reads are assembled by Velvet and Oases.
}
\label{figure2}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure3a.png}
\includegraphics[width=5in]{figure3b.png}
\end{center}
\caption{
{\bf Intron and exon graphs.}
a. Each intron connects to exons whose splice junctions match it boundary.
Some exons are excluded from the final gene model if they are incomplete.
b. Introns sharing at least one exon are grouped together.
Then an exon graph is made using exons as nodes.
}
\label{figure3}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure4.png}
\end{center}
\caption{
{\bf Different isoforms are detected by different k-mer lengths.}
}
\label{figure4}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure5.png}
\end{center}
\caption{
{\bf Global and local detect different isoforms with the same k-mers.} 
}
\label{figure5}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure6.png}
\end{center}
\caption{
{\bf Isoform length distribution.} 
}
\label{figure6}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure7.png}
\end{center}
\caption{
{\bf Isoform length distribution of two-exon transcripts.} 
}
\label{figure7}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure8.png}
\end{center}
\caption{
{\bf Cumulative distribution of lengths of two-exon transcripts homologous to mouse proteins.}
}
\label{figure8}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure9.png}
\end{center}
\caption{
{\bf Example of fragmented transcripts near $5'$ end of a long transcript.}
}
\label{figure9}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure10.png}
\end{center}
\caption{
{\bf Comparison of RNA-Seq derived gene models and other publicly available gene models from UCSC genome browser.}
}
\label{figure10}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure11.png}
\includegraphics[width=5in]{figure12.png}
\end{center}
\caption{
{\bf Examples of alternative $3'$ and $5'$ UTRs detected in RNA-Seq gene models.}
}
\label{figure11}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure13.png}
\end{center}
\caption{
{\bf Cumulative distribution of the ratio of the number of nucleotides to the number of amino acid encoding nucleotides.}
}
\label{figure13}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure14.png}
\includegraphics[width=5in]{figure15.png}
\end{center}
\caption{
{\bf Histogram of bit score/length ratio of isoforms and genes that match mouse proteins.}
}
\label{figure14}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure16.png}
\end{center}
\caption{
{\bf Cumulative counts of splice junctions with spliced reads.}
}
\label{figure16}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure17.png}
\end{center}
\caption{
{\bf Splice junctions found by Cufflinks and Gimme.}
}
\label{figure17}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{figure19.png}
\end{center}
\caption{
{\bf Number of splice junctions in chromosome 1 detected by local assembly, Cufflinks and Trinity.}
}
\label{figure19}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=7in]{figure20.png}
\end{center}
\caption{
{\bf Example of splice variants detected by local assembly but not Trinity.}
Local assembly detects two exons that are not detected by Trinity.
These two exons are included in a shorter isoform of this gene and are supported by ESTs.
Local assembly also detects another isoform that contains a skipped exon, which is also supported by ESTs.
}
\label{figure20}
\end{figure}

\section*{Tables}
\begin{table}[!ht]
\caption{
\bf{Unique sequences from global and local assembly}}
\begin{tabular}{ccccc}
\hline
Dataset & \multicolumn{2}{c}{Total Sequence} & \multicolumn{2}{c}{Unique Sequence}\\
 & Global & Local & Global & Local\\
\hline
Line 6 uninfected & 486,151 & 628,811 & 17,961 (3.6\%) & 3,563 (0.05\%)\\
Line 6 infected & 389,854 & 572,672 & 18,852 (4.8\%)& 4,107 (0.07\%)\\
Line 7 uninfected & 445,017 & 584,037 & 16,432 (3.6\%) & 3,991 (0.07\%)\\
Line 7 infected & 510,059 & 631,081 & 20,321 (3.9\%)& 3,649 (0.06\%)\\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
%\label{tab:label}

\caption{
\bf{Unique regions from global and local assembly}}
\begin{tabular}{ccccc}
\hline
Dataset & \multicolumn{2}{c}{Unique Region} & \multicolumn{2}{c}{Matched with mouse proteins}\\
 & Global & Local & Global & Local\\
\hline
Line 6 uninfected & 19,084 & 4,685 & 3,113 (16\%) & 194 (4\%)\\
Line 6 infected & 20,003 & 6,177 & 3,730 (19\%)& 224 (3.6\%)\\
Line 7 uninfected & 17,564 & 5,655 & 2,864 (16\%) & 320 (5.6\%)\\
Line 7 infected & 21,384 & 4,894 & 2,832 (13\%)& 238 (4.8\%)\\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
%\label{tab:label}

\caption{
\bf{Number of putative genes and isoforms}}
\begin{tabular}{cccccc}
\hline
Method& Gene & \multicolumn{4}{c}{Isoform} \\ 
& & Maximum & Minimum & Multiexon* & Two-exon* \\
\hline
Global & 27,448 & 65,376 & 42,375 & 27,891 & 14,484 (34\%) \\
Local & 22,434 & 82,285 & 38,876 & 27,724 & 11,152 (28\%) \\
Global + Local & 32,854 & 196,024 & 59,430 & 39,570 & 19,860 (33\%) \\
RefGenes & 4,567 & 5,423 & 4,701 & 4,420 & 281 (6\%) \\
Global + Local + RefGenes & 32,743 & 222,372 & 60,422 & 41,078 & 19,344 (32\%) \\
\hline
%table information
\end{tabular}
\begin{flushleft}\footnotesize \textit{*Minumum set}
\end{flushleft}
\label{tab:label}

\caption{
\bf{Single-end reads mapped to gene models (maximum)}}
\begin{tabular}{cccccc}
\hline
Dataset & Mapped & Unmapped \\
\hline
Line 6 uninfected & 21,967,788 (72.98\%) & 8,134,424 (27.02\%) \\
Line 6 infected & 19,091,229 (66.37\%) & 9,675,355 (33.63\%) \\
Line 7 uninfected & 19,442,853 (70.40\%) & 8,175,936 (29.60\%) \\
Line 7 infected & 21,880,520 (73.69\%) & 7,813,134 (26.31\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
%\label{tab:label}

\caption{
\bf{Paired-end reads mapped to gene models (maximum)}}
\begin{tabular}{cccccc}
\hline
Dataset & Mapped & Unmapped \\
\hline
Line 6 uninfected & 26,548,426 (60.67\%) & 17,213,674 (39.33\%) \\
Line 6 infected & 21,613,615 (60.06\%) & 14,370,321 (39.94\%) \\
Line 7 uninfected & 25,563,801 (59.96\%) & 17,068,932 (40.04\%) \\
Line 7 infected & 26,464,254 (60.55\%) & 17,245,099 (39.45\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
%\label{tab:label}

\end{table}
\end{document}

### 2012-08-15 rnaseq_paper.tex
replaced rnaseq_paper.tex with an updated one.

@@
 {\Large
-\textbf{mRNA-Seq Assembly Discovers Many Novel Splice Variants}
+\textbf{RNA-Seq Assembly Discovers Many Splice Variants}
 }
@@
 \section*{Abstract}
-
+Comparison of spliced reads and reference gene annotations has been successfully used to discover alternative
+splicing in model organisms. However, the method cannot be applied in organisms without high-quality reference genome and
+gene annotations. In this article, we introduce a pipeline, based on \emph{de novo} assembly, that constructs gene models with splice
+variants.
+The pipeline includes a new technique called local assembly that enhances the sensitivity of
+alternative splicing detection.
+We demonstrates that the pipeline detects many novel splice variants in RNA-Seq data from chicken spleens.
+Some of those splice variants; however, only detected by our pipeline and not Cufflinks.
+This indicates that the pipeline can be used to facilitate splice variants detection from RNA-Seq.
 % Please keep the Author Summary between 150 and 200 words
@@
 \section*{Introduction}
+Until recently, studies of alternative splicing had been limited to a small number of genes and isoforms due to high-cost and low-throughput sequencing of expressed sequence tags (ESTs) and full-length cDNA libraries.
+RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been used successfully in many studies to gain unprecedented insight into a complexity of transcriptomes.
+Novel isoforms have been identified based on reads mapped across known splice junctions or putative splice junctions from annotations.
+It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
+This suggests that even in human a large number of splice variants has not been explored.
+
+Sequences from RNA-Seq are very short (~75--250bp for Illumina reads); therefore, it is not feasible to assess gene
+and isoform expression without mapping reads to a reference annotation or construct a full-length transcript from
+\emph{de novo} assembly.
+Several softwares employing different approaches have been developed to reconstruct transcripts from short reads.
+Tophat\cite{Trapnell:2009dp} maps reads to a reference genome to identify exons and splice sites \emph{ab initio}.
+It first searches for known splie sites in a genome and then tries to map reads across splice juntions to identify exon junctions.
+Exon junctions can be used to build gene models by Cufflinks\cite{Trapnell:2010kd}.
+On the other hand, Trinity\cite{Grabherr:2011jb}, has been developed to assemble short reads \emph{de novo} to construct transcripts.
+Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} assembler that are designed to work with RNA-Seq data.
+These programs are comparable at identifying alternative splicing with a slightly different sensitivity and specificity.
+However, Oases with Oases-M has been shown to be superior than other \emph{de novo} assembler at discovering isoforms in human and mouse\cite{Schulz:2012je}.
+
+In this study, we present a pipeline that uses Velvet and Oases assembler to discover alternative splicing in RNA-Seq data from chickens.
+We use a new technique called a local assembly that enhances sensitivity of alternative splicing detection of Oases.
+The results show that the pipeline can detect more isoforms than Oases-M and can detect isoforms not found by
+Cufflinks.
 
@@
 \subsection*{Transcripts from Global and Local Assembly}
-\subsubsection*{Different Isoforms Detected by Different k-mers}
 
-Without high quality annotations, it is necessary that we construct gene models containing putative isoforms to identify alternative splicing in our samples.
-We used Velvet\cite{Zerbino:2008vu} and Oases assembler to construct transcripts from short reads.
-We ran Velvet using default settings except a hash length or k-mer parameter, which has no default values.
-To find an appropriate value for hash length for our datasets, we experimented on a wide range of k-mer lengths from 23 to 35.
-After comparing transcripts from all k-mers, we found that some isoforms were only detected in some particular k-mers (Fig.~\ref{figure4}).
-Therefore, instead of choosing one k-mer, we ran Velvet assembler with multiple k-mers and combined the results to increase sensitivity of isoform detection.
+Without high-quality annotations, it is necessary to construct gene models with putative isoforms to identify alternative splicing in our samples.
+We used Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je} assembler to construct transcripts from short reads.
+We ran Velvet using default settings except a hash length or k-mer length, which has no default values.
+It has been shown that transcripts from a combination of multiple k-mer lengths is superior than a single k-mer
+length\cite{Schulz:2012je}.
+In addition, we found that different k-mer lengths detected different isoforms from the same dataset
+(Fig.~\ref{kmer-variance}).
+For that reason, it is necessary to cover a wide range of k-mers to discover as many isoforms as possible.
+For this study, we selected a range of kmers from $21-31$ since we found that k-mers outside this range are prone
+to introduce assembly errors.
 
-\subsubsection*{Local Assembly Enhances Isoform Detection}
+\subsection*{Local Assembly Enhances Isoform Detection}
 
@@
 We refer to this method as a local assembly and a conventional assembly method as a global assembly in this study (see materials and methods).
-The advantages of the local assembly method are that each group of reads can be assembled using small amount of memory and it can be assembled on different computers.
-Furthermore, we found that local assembly could identify unique alternative splicing from global assembly.
-Therefore, we performed both global and local assembly with multiple k-mers (23-35 and 15-39 respectively) to obtain all possible alternative splicing detected from both methods.
-Figure~\ref{figure5} shows an example of isoforms detected with different k-mers in two assembly methods.
+The advantages of the local assembly method is that each group of reads can be assembled using small amount of memory and it can be assembled on different computers.
+Unexpectedly, we also found that local assembly detected alternative isoforms not found in global assembly.
+As a result, we merged transcripts from both global and local assembly from multiple k-mers to obtain all possible alternative isoforms from our samples.
+Figure~\ref{global_vs_local} shows an example of different isoforms detected from two assembly methods.
 
 Using global and local assembly methods, we obtained a different number of contigs.
-Table 1 shows a number of contigs from four datasets and unique contigs from global and local assemblies.
-As expected, some sequences from local assembly (0.5--0.7\%) are not detected in global assembly.
-Homology search in mouse proteins showed that 4--5.6\% of unique regions from local assembly are genuine protein coding sequences (Table 2).
+Table~\ref{unique_sequences} shows a number of contigs from four datasets and unique contigs from global and local assemblies.
+As expected, some sequences from local assembly (1,199--2,270) are not found in global assembly.
+Homology search in mouse proteins showed that 35.02--75.61\% of unique regions from local assembly match protein
+coding sequences (Table~\ref{unique_sequences_matched_mouse}).
 This suggests that some isoforms will be missing in the gene models based solely on global assembly.
-In addition, a significant number of transcripts from global assembly are unique, which indicates that approximately 4.8\% of coding sequences are missing in a current chicken genome.
+In addition, a significant number of transcripts from global assembly are unique, which indicates that
+approximately at least 3\% of coding sequences are missing in a current chicken genome.
 
-\subsection*{Predicted Gene Models from Assembly}
+\subsection*{Transcripts from Oases-M contain fewer splice junctions}
+Transcripts from multiple k-mer lengths can be merged using Oases-M.
+It has been demonstrated that merged transcripts are more complete than those from any single kmer length\cite{Schulz:2012je}.
+However, sensitivity of Oases-M in detecting alternative isoforms has not been fully evaluated.
+To evaluate this, we compared transcripts from global assembly with k-mer lenght from 21--31 and merged transcripts from Oases-M using k-mer=27. We used alignments of ESTs as a control.
+The results show that Oases-M detect fewer splice junctions than unmerged global assembly.
+A number of total splice junctions and unique splice junctions detected by Oases-M and unmerged assembly is
+shown in Table~\ref{Oases-M}.
+Only 391 unique splice junctions from Oases-M are supported by ESTs whereas unmerged assembly detects 1,752
+unique splice junctions that also found in ESTs.
+This suggests that some genuine splice junctions are missing from merging transcripts from multiple k-mer
+lengths with Oases-M; therefore, we did not use Oases-M in our pipeline.
 
-All transcripts were assembled by Gimme to construct gene models.
-Total of 27,448 multi-exon genes were obtained from global assembly and 22,434 genes from local assembly.
-However, the number of genes obtained from global + local assembly is only 32,854.
-This indicates that some transcripts from both methods were common or merged together.
-As expected, the number of genes and isoforms increased significantly after combining global and local assembly together due to the increased number of splice junctions (Table 3).
+\subsection*{Gene models prediction using Gimme}
 
-\subsection*{Fragmented transcripts}
-The number of genes from our pipeline is unexpectedly high due to a large number of two-exon genes.
-As shown in Table 3, the number of two-exon genes is high (28-34\% in models derived from assembly), compared to the number of annotated two-exon genes in reference genes (6\%).
-The majority of two-exon genes are shorter than 500bp as shown in Fig.~\ref{figure7}.
-Homology search against mouse proteins (using blastp) showed that 60\% of two-exon genes that matched mouse proteins are shorter than 500bp (Fig.~\ref{figure8}).
-The results suggest that these two-exon genes are indeed genuine coding sequences.
-However, according to their small size, they are possibly parts of larger transcripts that were not completely assembled.
-In our datasets, read coverage is lower at the $5'$ end due to the method used to converse mRNA to cDNA in sequencing.
-A bias of read coverage toward $3'$ end and low expression level result in fragmented transcripts as shown in Fig.~\ref{figure9}.
-Notably, the number of two-exon genes decreased when reference genes were incorporated to global + local assembly gene models (Table 3).
-This indicates that some two-exon genes were merged to reference genes since they are part of larger transcripts.
+We developed a program called, Gimme, to constructs a gene model from alignments of transcripts to the genome.
+The program can merge all transcripts from multiple-kmers and predict a structure of full-length isoforms and genes.
+As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
+Total of 14,832 genes were obtained from global assembly and 15,297 genes from local assembly.
+However, the number of genes obtained from global + local assembly is only 15,934.
+The number of genes and isoforms slightly increased after combining transcripts from global and local
+assembly together (Table~\ref{genes_transcripts}). 
+This indicates that some transcripts from both methods were merged together.
 
-\subsection*{Characteristics of RNA-Seq Derived Gene Models}
-Gene models derived from our pipeline are comparable or, in some cases, more complete than gene models available on UCSC genome browser\cite{ucscGenomeBrowser}.
-Fig.~\ref{figure10} shows a comparison of a predicted isoforms with gene models from reference genes, Ensembl[?], N-SCAN[?] and Genscan[?].
-Our gene models have an advantage of including extended untranslated regions (UTRs), especially $3'$ UTR in our datasets.
+Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
 These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
-RNA-Seq-derived gene models will facilitates the detection of variations within UTR regions, which may be involved in regulating the expression of the isoform [?] (Fig.~\ref{figure11}).
+RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).
 
-\subsection*{Comparison with Cufflinks and Trinity}
-To evaluate an efficiency of our pipeline in detecting splice junctions,
-we compared the number of splice junctions detected by our pipeline to those from Cufflinks\cite{Trapnell:2010kd} and Trinity\cite{Grabherr:2011jb}.
-Cufflinks detected 99,958 (85\%) splice junctions that are also detected by our pipeline (Fig.~\ref{figure17}).
-60,849 (37.8\%) are not detected by Cufflinks of which 29,207 (48\%) are supported by at least four spliced reads.
-
-Reads from line 6 uninfected dataset that mapped to a chromosome 1 were assembled using Trinity with default settings.
-Trinity found a total of 11,902 splice junctions; whereas, our pipeline detected 12,676 splice junctions.
-1,395 splice junctions are not found by Trinity and 621 junctions are not detected by our pipeline.
-Figure~\ref{figure19} summarizes a number of splice junctions in chromosome 1 detected by all three methods.
-Figure~\ref{figure20} shows two splice variants that are detected by local assembly but not Trinity.
+\subsection*{An overestimated number of genes and transcripts}
+The number of genes from our pipeline might be overestimated due to a number of fragmented transcripts.
+In our datasets, read coverage is lower at the $5'$ end due to the method used to convert mRNA to cDNA
+in library preparation.
+A bias of read coverage toward $3'$ end and low expression level result in fragmented transcripts as shown in Fig.~\ref{fragmented_transcripts}.
+In addition, overestimation of a number of isoforms is due to spurious splice junctions, which can result in more than 10,000 possible isoforms in one gene.
+This seems to occur in a few genome sequences such as chromosome E64\_random.
+Figure~\ref{isoforms_boxplot} shows that a majority of genes from our gene models have only a few isoforms.
 
@@
 
-Single-end reads from the same datasets that we used to build the gene models were aligned to the DNA sequences derived from the gene models.
+Single-end reads from the same datasets that we used to build the gene models were aligned to cDNA sequences derived from the gene models.
 We anticipated that a large number of reads could align to the gene models.
-We found that 66-73\% of reads mapped to the gene models (Table 4).
-In addition, we mapped paired-end reads that we obtained later from the same samples to the gene models and found that more than 60\% of reads mapped to the gene models (Table 5).
+We found that 63-78\% of reads mapped to the gene models (Table~\ref{single-end_map}).
+In addition, we mapped paired-end reads that we obtained later from the same samples to
+the gene models and found that 63-65\% of reads mapped to the gene models (Table~\ref{paired-end_map}).
 The results suggest that the gene models are of high quality.
 
-\subsubsection*{Conserved Reading Frame}
-Alternative isoforms usually maintain a reading frame to be functional [?].
-We hypothesized that errors in our pipeline that introduce false splice junctions will disrupt the reading frame.
-Although, in nature, alternative-splicing events can disrupt the reading frame, we expected a significant number of isoforms to be translatable.
+\subsubsection*{Reads mapped to splice junctions}
+To validate the splice junctions, we mapped reads to cDNA sequences derived from gene models using Bowtie\cite{Langmead:2009fv}.
+Because the mapping algorithm implemented in Bowtie does not consider splice sites, reads mapped across splice junctions
+confirm that the splice junctions are in transcripts.
+Of 112,089 splice junctions from the gene models, only 1,341 (1.2\%) junctions have no spliced reads and 3,447 (3\%)
+junctions have 1--3 spliced reads (see materials and methods) (Fig.~\ref{cdf_single_splice}.
+The results indicate the method can detect splice junctions with a very high specificity.
+Some splice junctions with fewer than four spliced reads; however, may not be false junctions because of a bias of
+read coverage toward $3'$ end in our datasets as mentioned earlier.
+Reads are less abundant at the $5'$ end of transcripts; consequently, a relatively small number of spliced reads is mapped
+to splice junctions near the $5'$ end.
 
-Transcripts from RNA-Seq assembly are supposed to contain UTRs and coding sequences. However, due to some factors mentioned previously, we could not obtain complete transcripts for all isoforms.
-Thus, predicting the reading frame or coding sequence solely from sequence signals is not feasible.
-We used ESTScan\cite{Iseli:1999vd}, a program that is designed to find coding sequences from expressed sequence tags (ESTs) and does not rely on sequence signals to translate isoforms to protein sequences.
-To evaluate the result, we used reference mRNAs as a control.
-However, since our gene models derived from the chicken genome sequence; therefore, we first aligned reference mRNAs to the genome to obtain DNA sequences to be used in the comparison.
+\subsubsection*{Comparison with ESTs and Genbank mRNAs}
+83,238 (79.7\%) splice junctions from our gene models are supported by ESTs or Genbank mRNAs.
+This illustrates the efficiency of our pipeline in detecting genuine splice junctions.
+We also found that 46,132 junctions that are supported by both Genbank mRNAs and ESTs are also supported by at least four spliced reads.
+Therefore, we can assume that splice junctions with more than three spliced reads are likely to be real.
+Using this criterion, we can conclude that our gene models include 21,872 novel splice junctions that are not in ESTs.
 
-ESTScan successfully translated 47,786 (80\%) isoforms from our gene models and 15,952 and 14,999 (97\%) sequences from reference mRNAs and models built from reference mRNAs.
-We evaluated the completeness of coding sequence by comparing the number of nucleotides in each isoform to the number of amino acid encoding nucleotides.
-We found that cumulative distributions of ratios from reference mRNAs before and after processed by our pipeline are highly similar.
-This indicates that the pipeline did not introduce errors that disrupt the reading frame of the mRNAs.
-The result from our gene models shows significant similarity to results from mRNAs, which suggests that a majority of isoforms detected by our pipeline is translatable.
+\subsubsection*{Comparison with Cufflinks}
+To evaluate an efficiency of our pipeline in detecting splice junctions,
+we compared the number of splice junctions detected by our pipeline to those from Cufflinks\cite{Trapnell:2010kd}, ESTs and
+Ensembl gene models version 64.
+Cufflinks detected 93,756 (83\%) splice junctions that are also detected by our pipeline.
+However, 46\% (8,615) of junctions not detected by Cufflinks are supported by ESTs.
+This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
+genuine splice junctions.
+
+We also aligned transcripts from Cufflinks and the pipeline to Ensembl transcripts to compare genes and transcripts detected by
+both methods.
+We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
+Our pipeline detects 7,765 genes and Cufflinks detects 8,983 genes of total 17,934 Ensembl genes.
+7,049 of those genes are both detectd by our pipeline and Cufflinks.
+In addition, we wanted to compare a number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
+a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
+Therefore, we compare a number of splice junctions detected by both methods within the same gene instead.
+From 7,049 genes, our pipeline detects 1,765 splice junctions, whereas Cufflinks detects 2,516 junctions.
+
+However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
+that are not in Ensembl but are supported by ESTs in those 7,049 genes and found that our pipeline detects 2,039 splice junctions
+and Cufflinks only detects 999 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
+
+To conclude, Cufflinks performs better than our pipeline for detecting both genes and splice junctions in Ensembl gene models.
+However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs.
 
 \subsubsection*{Homologous sequences in mouse}
-To further examine quality of protein sequences derived from our gene models, we searched for homologous sequences in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
-We found 29,903 (52\%) isoforms (7,425 genes) match mouse proteins.
-As shown in Fig.~\ref{figure14}, most genes and isoforms have a bit score/length ratio more than 1.0, which indicates a good agreement between chicken and mouse proteins.
+Splice variants must maintain the reading frame to be translated to functional proteins.
+Based on this assumption, we investigated isoforms in our gene models by searching for homologous protein sequences in
+other organism.
+To obtain a protein sequence of each isoform, we translated a cDNA sequence using ESTScan\cite{Iseli:1999vd},
+which is designed to translate
+protein sequences from incomplete cDNA sequences such as ESTs.
+ESTScan succesfully translated 44,122 (94\%) of all isoforms to protein sequences with 50 or longer amino acids.
+Then, all protein sequences were searched for homology in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
+We found 27,770 (63\%) isoforms from 4,272 genes match mouse proteins.
+As shown in Fig.~\ref{bitscore}, most genes and isoforms have a bit score/length ratio more than 1.0, which
+indicates a good agreement between chicken and mouse proteins.
 The results illustrate that isoforms constructed by our pipeline are translatable and may be functional.
 However, genes or isoforms with no match to mouse proteins are not necessarily artifacts.
-They can be genes or isoforms that are only found in other vertebrates or novel genes.
-
-\subsubsection*{Reads mapped to splice junctions}
-We identified 160,807 splice junctions from the gene models.
-To validate the splice junctions, we mapped reads to DNA sequences derived from gene models using Bowtie\cite{Langmead:2009fv}.
-Because the mapping algorithm implemented in Bowtie does not consider splice sites, reads mapped across splice junctions confirm that the splice junctions are real.
-Of all spliced junctions, 18,004 (11\%) junctions have no spliced reads and 11,815 (6.9\%) junctions have 1--3 spliced reads (see materials and methods).
-This demonstrates that the method detects splice junctions with high specificity.
-Splice junctions with fewer than four spliced reads, however, may not be considered false junctions because of a bias of read coverage toward $3'$ end in our datasets.
-Reads are less abundant at the $5'$ end of transcripts; therefore, a relatively small number of spliced reads mapped to splice junctions near the $5'$ end.
+They can be genes or isoforms that are only found in non-mammal vertebrates or novel genes.
 
@@
 
-In this study, we used multiple k-mers and a local assembly method to enhance splice variant detection.
-Our new method detects splice variants that are not detected by a conventional assembly method.
-We also showed that our method can be used with other programs (Cufflinks, Trinity, etc.) to increase sensitivity of splice variant detection.
-It works well with single-ended reads and does not require more memory than the conventional assembly method.
-The method, however, relies on a quality of a reference genome and a read mapping program.
+Choice of k-mers greatly affects sensitivity of splice variants detection in RNA-Seq assembly using Velvet/Oases.
+Short k-mers increase sensitivity, but also introduce errors from misassembly.
+However, for a study focusing on alternative splicing, it is desirable to detect as many splice variant as possible.
+We have presented the local assembly technique that enhances the ability to detect splice variants of Velvet/Oases from the same
+k-mer.
+The mechanism behind this technique is to be investigated.
+Importantly, understanding of the mechanisms may lead to a filtering technique that can be applied to organisms that lack a reference genome.
 
-We also presented Gimme, a program that assembles transcripts from different approaches based on alignments to a reference genome to construct gene models.
-We used the pipeline to construct gene models from chicken mRNAs using global and local assembly to identify novel genes and isoforms in chicken.
-Many novel genes and isoforms are found and validation results suggest that a majority of them are genuine and may be of biological importance.
-The gene models also markedly improve existing annotations.
+We have also developed a pipeline that assembles reconstructed transcripts from multiple k-mers to build putative gene models.
+Overall, a majority of splice variants detected by our pipeline are also detected by Cufflinks.
+Both our pipeline and Cufflinks also detect some splice variants that are not overlapped.
+This suggests that combination of both \emph{de novo} assembly and reference-based assembly is preferred to study alternative splicing in chicken.
+
+Interestingly, Cufflinks detects more genes and isoforms included in Ensembl annotations than our pipeline.
+However, our pipeline detects many more splice junctions not included in Ensembl but supported by ESTs.
+As observed in previous study by Schulz \emph{et al.}\cite{Schulz:2012je}, Cufflinks also outperforms \emph{de novo} assembly in
+detecting genes and isoforms in Ensembl annotations in mouse.
+We speculate that Cufflinks possesses a mechanism that is tailored to detect Ensembl gene models.
+
+In this study, gene models are built from chicken genome version 2.1 (galGal3), which contains a large number of sequence duplications and
+missassemblies that supposed to be eliminated in the latest version of genome assembly (galGal4).
+Duplications and misassemblies lead to false splice junctions, which in turn produce a large number of splice variants as
+observed in some chromosomes.
+Conversely, this suggests that one might be able to use the pipeline to evaluate the quality of reference genome.
 
@@
 
+\subsection*{Quality trimming of reads}
+Both single- and paired-end reads in this study were trimmed using Condetri version 2.1 with default parameters.
+In addition, the first 10 bases of each reads were trimmed off due to an inconsistency of base-calling as shown in supplementary figure S?.
+
 \subsection*{Mapping reads to The Reference Genome and Gene Models}
 
-Single and paired-end reads were mapped to chicken genome by Tophat\cite{Trapnell:2009dp} version 1.0.13 using default parameters without annotations.
-All reads were mapped to DNA sequences derived from gene models by Bowtie version 0.12.7\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28, e=70, k=1)}.
+Single and paired-end reads were mapped to chicken genome by Tophat2\cite{Trapnell:2009dp} release 2.0.0 using
+default parameters without annotations.
+All reads were mapped to cDNA sequences derived from gene models by Bowtie2\cite{Langmead:2009fv} with
+default parameters \texttt{(n=2, l=28, e=70, k=1)}.
 
@@
 Reads from each dataset were first assembled separately in global assembly without using a reference genome.
-In contrast, reads from each dataset were first mapped to the chicken genome using Tophat.
-Then only reads that mapped to the genome were assembled by chromosomes in local assembly (Fig.~\ref{figure1}).
-Global and local assembly was performed using Velvet version 1.0.17\cite{Zerbino:2008vu} with default parameters except for hash length (k-mer).
-k-mer lengths ranging from 15 to 39 were used in local assembly and from 23 to 35 in global assembly to increase sensitivity of splice junction detection.
-Lastly, transcripts from both methods were assembled by Oases version 0.1.18 (unpublished).
+In contrast, reads from each dataset were first mapped to the chicken genome using Tophat2.
+Then only reads mapped to the genome were assembled by chromosomes in local assembly (Fig.~\ref{local_assembly}).
+Global and local assembly was performed using Velvet version 1.2.03\cite{Zerbino:2008vu}
+with default parameters except for hash length (k-mer).
+A range of k-mer length from 21-31 was used to assemble reads in both global and local assembly.
+Lastly, transcripts from both methods were assembled by Oases version 0.2.06\cite{Schulz:2012je}.
+
+\subsection*{Transcript Reassembly}
+To reduce resources and time to reassemble transcripts, we first trimmed off poly-A tail and remove low
+complexity and short sequences using seqclean\cite{seqclean} with default parameters.
+Then we used cd-hit-est from CD-HIT suite\cite{Li:2006hr} to remove all redundant transcripts.
+A large number of transcripts was removed at this step, which facilitates the reassembly process.
+In order to eliminate some sequence variations between samples.
+We mapped sequences to the chicken genome and then extracted all sequences from the genome.
+All transcripts from each chromosome were then assembled by CAP3\cite{Huang:1999wb}.
 
@@
 
-Transcripts of all datasets from local and global assembly were mapped to the chicken genome using GMAP version 2011-03-28.v3\cite{Wu:2005bl} \texttt{(-n 0)}.
-Alignments and gaps from GMAP outputs are considered exons and introns respectively.
+Figure~\ref{overall_pipeline} depicts an overall gene model construction pipeline.
+Transcripts of all datasets from local and global assembly were mapped to the chicken genome using
+BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna -noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.
+Alignments and gaps from BLAT outputs are considered exons and introns respectively.
 Optionally, data from other sources (ESTs, RefGenes, etc.) can be incorporated with transcripts from assembly to improve gene models.
@@
 An algorithm for assembling transcripts is described below.
+A maximum set of transcripts obtained from Gimme are then reduced to only a minimum set of transcripts that contain
+all splice junctions and untranslated regions (UTRs).
+After that, transcripts that are identical are clustered and removed by CD-HIT version 4.5.6\cite{Li:2006hr}.
+Parameters for clustering are \texttt{-n 10, -r 0 and -c 1.0}.
+Only a representative of each cluster is kept in gene models.
 
@@
 
-The algorithm first builds an intron graph using introns as nodes  (Fig.~\ref{figure3}).
+The algorithm first builds an intron graph using introns as nodes.
 Each intron contains exons whose one of their splice sites perfectly match intron boundary.
-Exons are considered incomplete and eliminated if they locate at the $3'$ or $5'$ end of the transcripts and they are not the largest exons (exon 3a and 3b in Fig.~\ref{figure3}).
-Transcripts were then grouped into the same gene if they have at least one intron or exon in common (Fig.~\ref{figure3}).
+Exons are considered incomplete and eliminated if they locate at the $3'$ or $5'$ end of the transcripts and
+they are not the largest exons (exon 3a and 3b in Fig.~\ref{algorithm}).
+Transcripts were then grouped into the same gene if they have at least one intron or exon in common.
 Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing paths in the splice graph.
-Two sets of putative transcripts can be constructed --- a maximum and a minimum set.
-The maximum set contains all putative isoforms, which are derived from all paths in splice graphs.
-In contrast, the minimum set contains minimum isoforms that represents all unique edges in splice graphs.
+Gimme is open-source and available at \texttt{https://github.com/ged-lab/gimme}.
 
-\subsection*{Open Reading Frame and Strand Prediction}
+\subsection*{Protein sequence translation}
 
-We employed ESTScan version 2.1\cite{Iseli:1999vd} to predict strand and ORFs for our gene models.
+We employed ESTScan version 2.1 to translate protein sequences from our gene models.
 The matrix used for building Hidden Markov model was built from chicken reference cDNA sequences using tools from ESTScan.
 
-\subsection*{Finding Unique Sequences Between Datasets}
+\subsection*{Finding unique sequences between datasets}
 
-To identify unique sequences from two datasets, a set of 20-mers is created for both datasets using khmer [?].
+To identify unique sequences from two datasets, a set of 20-mers is created for both datasets using
+khmer\cite{khmer}.
 Then, 20-mers from a query dataset are compared with 20-mers from the target dataset.
 The sequence is considered unique if more than 90\% of 20-mers in the query is unique.
-Any unique region shorter than 100bp is ignored.
+Any unique region shorter than 100 bp is ignored.
 
-\subsection*{Sequence Homology Analysis}
+\subsection*{Sequence homology analysis}
 
-Protein sequences translated from each isoform using ESTScan were searched against mouse reference sequences [?] by BLAST 2.2.25+\cite{Tatusova:1999tz}.
-A bit score to a length ratio was calculated for each hit that had e-value $\le 10^{-20}$.
-Only the highest ratio of all isoforms of a gene was used in the gene-wise analysis, whereas, all ratios were used in the isoform-wise analysis.
+Protein sequences translated from each isoform using ESTScan were searched against mouse reference proteins by BLAST 2.2.25+\cite{Tatusova:1999tz}.
+A bit score to a length ratio was calculated for each hit that had an e-value $\le 10^{-20}$.
+Only the highest value of all isoforms from each gene was shown in the gene plot; whereas, values of all isoforms were shown in the isoform plot.
 
-\subsection*{Spliced Reads Counting}
+\subsection*{Spliced reads count}
 
-Reads from each dataset were mapped to transcripts from the gene models using Bowtie version 0.12.7 using default parameters.
-Only reads that mapped across exon junctions from all datasets were counted using Samtools\cite{Li:2009vz} and Pysam [?]
+Reads from each dataset were mapped to transcripts from the gene models using Bowtie version 2.0-beta 5 with default parameters.
+Reads mapped across exon junctions from all datasets were counted using Samtools\cite{Li:2009vz} and
+Pysam\cite{pysam}
 
-\subsection*{Sequence Assembly using Trinity}
-To be added.
+\subsection*{Sequence assembly using Cufflinks}
+Reads are mapped to a genome sequence using Tophat2.
+Gene models are built from each datasets by Cufflinks 2.0.0\cite{Trapnell:2010kd}.
+All gene models are then merged together using Cuffmerge.
 
-\subsection*{Sequence Assembly using Cufflinks}
-To be added.
+\subsection*{Expressed sequence tags and Genbank mRNA}
+Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome website on May 9th, 2012.
+The database last updated from GENBANK on April 29th, 2012.
+Sequences were alignd to a chicken genome using BLAT.
 
 % Do NOT remove this, even if you are not including acknowledgments
-\section*{Acknowledgments}
-
-Royal Thai Government Scholarship.
-
-{\noindent}USDA funding.
-
 %\section*{References}
@@
 \bibliography{refs}
-
-\section*{Figure Legends}
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure1.png}
-\end{center}
-\caption{
-{\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
-Transcripts are aligned to a chicken genome by GMAP. Gimme then constructs gene models based on alignments of transcripts.
-}
-\label{figure1}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure2.png}
-\end{center}
-\caption{
-{\bf Local Assembly Pipeline.}
-Reads are first mapped to a chicken genome.
-Then only mapped reads are assembled by Velvet and Oases.
-}
-\label{figure2}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure3a.png}
-\includegraphics[width=5in]{figure3b.png}
-\end{center}
-\caption{
-{\bf Intron and exon graphs.}
-a. Each intron connects to exons whose splice junctions match it boundary.
-Some exons are excluded from the final gene model if they are incomplete.
-b. Introns sharing at least one exon are grouped together.
-Then an exon graph is made using exons as nodes.
-}
-\label{figure3}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure4.png}
-\end{center}
-\caption{
-{\bf Different isoforms are detected by different k-mer lengths.}
-}
-\label{figure4}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure5.png}
-\end{center}
-\caption{
-{\bf Global and local detect different isoforms with the same k-mers.} 
-}
-\label{figure5}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure6.png}
-\end{center}
-\caption{
-{\bf Isoform length distribution.} 
-}
-\label{figure6}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure7.png}
-\end{center}
-\caption{
-{\bf Isoform length distribution of two-exon transcripts.} 
-}
-\label{figure7}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure8.png}
-\end{center}
-\caption{
-{\bf Cumulative distribution of lengths of two-exon transcripts homologous to mouse proteins.}
-}
-\label{figure8}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure9.png}
-\end{center}
-\caption{
-{\bf Example of fragmented transcripts near $5'$ end of a long transcript.}
-}
-\label{figure9}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure10.png}
-\end{center}
-\caption{
-{\bf Comparison of RNA-Seq derived gene models and other publicly available gene models from UCSC genome browser.}
-}
-\label{figure10}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure11.png}
-\includegraphics[width=5in]{figure12.png}
-\end{center}
-\caption{
-{\bf Examples of alternative $3'$ and $5'$ UTRs detected in RNA-Seq gene models.}
-}
-\label{figure11}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure13.png}
-\end{center}
-\caption{
-{\bf Cumulative distribution of the ratio of the number of nucleotides to the number of amino acid encoding nucleotides.}
-}
-\label{figure13}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure14.png}
-\includegraphics[width=5in]{figure15.png}
-\end{center}
-\caption{
-{\bf Histogram of bit score/length ratio of isoforms and genes that match mouse proteins.}
-}
-\label{figure14}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure16.png}
-\end{center}
-\caption{
-{\bf Cumulative counts of splice junctions with spliced reads.}
-}
-\label{figure16}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure17.png}
-\end{center}
-\caption{
-{\bf Splice junctions found by Cufflinks and Gimme.}
-}
-\label{figure17}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{figure19.png}
-\end{center}
-\caption{
-{\bf Number of splice junctions in chromosome 1 detected by local assembly, Cufflinks and Trinity.}
-}
-\label{figure19}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=7in]{figure20.png}
-\end{center}
-\caption{
-{\bf Example of splice variants detected by local assembly but not Trinity.}
-Local assembly detects two exons that are not detected by Trinity.
-These two exons are included in a shorter isoform of this gene and are supported by ESTs.
-Local assembly also detects another isoform that contains a skipped exon, which is also supported by ESTs.
-}
-\label{figure20}
-\end{figure}
 
@@
 \hline
-Line 6 uninfected & 486,151 & 628,811 & 17,961 (3.6\%) & 3,563 (0.05\%)\\
-Line 6 infected & 389,854 & 572,672 & 18,852 (4.8\%)& 4,107 (0.07\%)\\
-Line 7 uninfected & 445,017 & 584,037 & 16,432 (3.6\%) & 3,991 (0.07\%)\\
-Line 7 infected & 510,059 & 631,081 & 20,321 (3.9\%)& 3,649 (0.06\%)\\
+Line 6 uninfected & 338,353 & 335,180 & 10,068 (2.97\%) & 1,258 (0.37\%)\\
+Line 6 infected & 353,485 & 319,920 & 11,639 (3.29\%)& 2,270 (0.70\%)\\
+Line 7 uninfected & 335,876 & 302,443 & 10,649 (3.17\%) & 1,570 (0.51\%)\\
+Line 7 infected & 371,404 & 327,491 & 11,859 (3.19?\%)& 1,199 (0.36\%)\\
 \hline
@@
 %\end{flushleft}
-%\label{tab:label}
+\label{unique_sequences}
 
@@
 \hline
-Line 6 uninfected & 19,084 & 4,685 & 3,113 (16\%) & 194 (4\%)\\
-Line 6 infected & 20,003 & 6,177 & 3,730 (19\%)& 224 (3.6\%)\\
-Line 7 uninfected & 17,564 & 5,655 & 2,864 (16\%) & 320 (5.6\%)\\
-Line 7 infected & 21,384 & 4,894 & 2,832 (13\%)& 238 (4.8\%)\\
+Line 6 uninfected & 2,978 & 117 & 1,956 (52.76\%) & 41 (35.04\%)\\
+Line 6 infected & 3,524 & 236 & 1,991 (56.50\%)& 100 (42.37\%)\\
+Line 7 uninfected & 3,334 & 163 & 1,933 (57.98\%) & 86 (52.76\%)\\
+Line 7 infected & 3,235 & 82 & 1,858 (57.43\%)& 62 (75.61\%)\\
 \hline
@@
 %\end{flushleft}
-%\label{tab:label}
+\label{unique_sequences_matched_mouse}
+
+\caption{
+\bf{Number of total and unique splice junctions}}
+\begin{tabular}{cccccc}
+\hline
+Method& Total & Unique & Supported by ESTs \\ 
+\hline
+Oases-M assembly & 114,336 & 8,510 & 391 (4.59\%) \\
+Unmerged assembly & 116,573 & 10,747 & 1,752 (16.30\%) \\
+\hline
+%table information
+\end{tabular}
+%\begin{flushleft}\footnotesize \textit{*Minumum set}
+%\end{flushleft}
+\label{Oases-M}
 
@@
 \hline
-Method& Gene & \multicolumn{4}{c}{Isoform} \\ 
-& & Maximum & Minimum & Multiexon* & Two-exon* \\
+Method& Gene & Transcript \\ 
 \hline
-Global & 27,448 & 65,376 & 42,375 & 27,891 & 14,484 (34\%) \\
-Local & 22,434 & 82,285 & 38,876 & 27,724 & 11,152 (28\%) \\
-Global + Local & 32,854 & 196,024 & 59,430 & 39,570 & 19,860 (33\%) \\
-RefGenes & 4,567 & 5,423 & 4,701 & 4,420 & 281 (6\%) \\
-Global + Local + RefGenes & 32,743 & 222,372 & 60,422 & 41,078 & 19,344 (32\%) \\
+Global & 14,832 & 32,311 \\
+Local & 15,297 & 23,028 \\
+Global + Local & 15,934 & 46,797 \\
+Cufflinks & 30,235 & 37,967 \\
+Ensembl & 17,934 & 23,392 \\
 \hline
@@
 \end{tabular}
-\begin{flushleft}\footnotesize \textit{*Minumum set}
-\end{flushleft}
-\label{tab:label}
+%\begin{flushleft}\footnotesize \textit{*Minumum set}
+%\end{flushleft}
+\label{genes_transcripts}
 
@@
 \hline
-Line 6 uninfected & 21,967,788 (72.98\%) & 8,134,424 (27.02\%) \\
-Line 6 infected & 19,091,229 (66.37\%) & 9,675,355 (33.63\%) \\
-Line 7 uninfected & 19,442,853 (70.40\%) & 8,175,936 (29.60\%) \\
-Line 7 infected & 21,880,520 (73.69\%) & 7,813,134 (26.31\%) \\
+Line 6 uninfected & 18,375,966 (77.93\%) & 5,203,586 (22.07\%) \\
+Line 6 infected & 17,160,695 (73.18\%) & 6,288,286 (26.82\%) \\
+Line 7 uninfected & 18,130,072 (75.77\%) & 5,795,737 (24.22\%) \\
+Line 7 infected & 19,912,046 (78.21\%) & 5,450,521 (21.49\%) \\
 \hline
@@
 %\end{flushleft}
-%\label{tab:label}
+\label{single-end_map}
 
@@
 \hline
-Line 6 uninfected & 26,548,426 (60.67\%) & 17,213,674 (39.33\%) \\
-Line 6 infected & 21,613,615 (60.06\%) & 14,370,321 (39.94\%) \\
-Line 7 uninfected & 25,563,801 (59.96\%) & 17,068,932 (40.04\%) \\
-Line 7 infected & 26,464,254 (60.55\%) & 17,245,099 (39.45\%) \\
+Line 6 uninfected & 21,598,218 (64.16\%) & 12,065,659 (35.84\%) \\
+Line 6 infected & 15,274,638 (63.89\%) & 15,274,638 (36.11\%) \\
+Line 7 uninfected & 20,961,033 (63.67\%) & 20,961,033 (36.33\%) \\
+Line 7 infected & 22,485,833 (65.22\%) & 22,485,833 (34.78\%) \\
 \hline
@@
 %\end{flushleft}
-%\label{tab:label}
+\label{paired-end_map}
+\end{table}
 
-\end{table}
+\section*{Figure Legends}
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{overall_pipeline.pdf}
+\end{center}
+\caption{
+{\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
+Transcripts are aligned to a chicken genome by BLAT. Gimme then constructs gene models based on alignments of transcripts.
+}
+\label{overall_pipeline}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{local_assembly}
+\end{center}
+\caption{
+{\bf Local Assembly Pipeline.}
+Reads are first mapped to a chicken genome.
+Then only mapped reads are assembled by Velvet and Oases.
+}
+\label{local_assembly}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{algorithm.pdf}
+\end{center}
+\caption{
+{\bf Intron and exon graphs.}
+Each intron connects to exons whose splice junctions match it boundary.
+Some exons are excluded from the final gene model if they are incomplete (exon 3a,b).
+Introns sharing at least one exon are grouped together.
+Then an exon graph is made using exons as nodes.
+}
+\label{algorithm}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{kmers-variance.pdf}
+\end{center}
+\caption{
+{\bf Different isoforms are detected by different k-mer lengths.}
+K-mer=21 detects a skipped exon which is not detected by k-mer=31.
+The skipped exon is also annotated in Ensembl gene models.
+}
+\label{kmer-variance}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{global_vs_local.pdf}
+\end{center}
+\caption{
+{\bf Global and local assembly detect different isoforms with the same k-mers.} 
+}
+\label{global_vs_local}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{isoforms_boxplot.pdf}
+\end{center}
+\caption{
+{\bf Distribution of a number of isoforms in each gene from 1) global and 2) local assembly}
+}
+\label{isoforms_boxplot}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{fragmented_transcripts.pdf}
+\end{center}
+\caption{
+{\bf Example of fragmented transcripts near $5'$ end of a long transcript.}
+}
+\label{fragmented_transcripts}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{model_comparisons.pdf}
+\end{center}
+\caption{
+{\bf Comparison of gene models from our pipeline and other public gene models on UCSC genome browser.}
+}
+\label{model_comparisons}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{long_utr.pdf}
+\end{center}
+\caption{
+{\bf Examples of an extended $3'$ UTR detected in RNA-Seq gene models.}
+}
+\label{long_utr}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{genes_bitscore.pdf}
+\includegraphics[width=5in]{isoforms_bitscore.pdf}
+\end{center}
+\caption{
+{\bf Histogram of bit score/length ratio of isoforms and genes that match mouse proteins.}
+Genes in chromosome E64\_random are not included.
+}
+\label{bitscore}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{cdf_single_splice.pdf}
+\end{center}
+\caption{
+{\bf Cumulative counts of splice junctions with spliced reads.}
+}
+\label{cdf_single_splice}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=7in]{alt_splice_site.pdf}
+\end{center}
+\caption{
+{\bf Unannotated alternative splice site.}
+The pipeline can detect alternative splice site that is not annotated in Ensembl and Cufflinks.
+}
+\label{alt_splice_site}
+\end{figure}
+
 \end{document}

### 2012-08-28 rnaseq_paper.tex
edited some contents, numbers and some figures

@@
 It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
-This suggests that even in human a large number of splice variants has not been explored.
+This suggests that even in human a large number of splice variants have not been explored.
 
@@
 
-Using global and local assembly methods, we obtained a different number of contigs.
-Table~\ref{unique_sequences} shows a number of contigs from four datasets and unique contigs from global and local assemblies.
-As expected, some sequences from local assembly (1,199--2,270) are not found in global assembly.
-Homology search in mouse proteins showed that 35.02--75.61\% of unique regions from local assembly match protein
+Using global and local assembly methods, we obtained the different number of contigs.
+Table~\ref{unique_sequences} shows the number of contigs from four datasets and unique contigs from global and local assemblies.
+As expected, some sequences from local assembly (1,690--2,751) are not found in global assembly.
+Homology search in mouse proteins showed that 32--38\% of long unique regions ($>300$ bp) from local assembly match protein
 coding sequences (Table~\ref{unique_sequences_matched_mouse}).
@@
 In addition, a significant number of transcripts from global assembly are unique, which indicates that
-approximately at least 3\% of coding sequences are missing in a current chicken genome.
+approximately at least 3\% of coding sequences are missing in a current chicken genome (sequences not found in local assembly).
 
@@
 The results show that Oases-M detect fewer splice junctions than unmerged global assembly.
-A number of total splice junctions and unique splice junctions detected by Oases-M and unmerged assembly is
+The number of total splice junctions and unique splice junctions detected by Oases-M and unmerged assembly is
 shown in Table~\ref{Oases-M}.
-Only 391 unique splice junctions from Oases-M are supported by ESTs whereas unmerged assembly detects 1,752
+Only 447 unique splice junctions from Oases-M are supported by ESTs whereas unmerged assembly detects 1,640
 unique splice junctions that also found in ESTs.
@@
 As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
-Total of 14,832 genes were obtained from global assembly and 15,297 genes from local assembly.
-However, the number of genes obtained from global + local assembly is only 15,934.
+Total of 14,703 genes were obtained from global assembly and 14,931 genes from local assembly.
+However, the number of genes obtained from global + local assembly is only 15,411.
 The number of genes and isoforms slightly increased after combining transcripts from global and local
@@
 \subsection*{An overestimated number of genes and transcripts}
-The number of genes from our pipeline might be overestimated due to a number of fragmented transcripts.
+The number of genes from our pipeline might be overestimated due to fragmented transcripts.
 In our datasets, read coverage is lower at the $5'$ end due to the method used to convert mRNA to cDNA
@@
 A bias of read coverage toward $3'$ end and low expression level result in fragmented transcripts as shown in Fig.~\ref{fragmented_transcripts}.
-In addition, overestimation of a number of isoforms is due to spurious splice junctions, which can result in more than 10,000 possible isoforms in one gene.
+Moreover, spurious splice junctions due to duplication in genome sequence can lead to a large number of dubious isoforms.
 This seems to occur in a few genome sequences such as chromosome E64\_random.
 Figure~\ref{isoforms_boxplot} shows that a majority of genes from our gene models have only a few isoforms.
+
+In addition, parameters used in building gene models such as minimum exon/intron size determine how transcripts are merged together, which can increases or decreases the number of isoforms.
+In this study, we adjust the parameters to enhance sensitivity of splice junctions detection that may result in overestimation of the number of isoforms.
 
@@
 We anticipated that a large number of reads could align to the gene models.
-We found that 63-78\% of reads mapped to the gene models (Table~\ref{single-end_map}).
+We found that 73--78\% of reads mapped to the gene models (Table~\ref{single-end_map}).
 In addition, we mapped paired-end reads that we obtained later from the same samples to
-the gene models and found that 63-65\% of reads mapped to the gene models (Table~\ref{paired-end_map}).
+the gene models and found that 77--78\% of reads mapped concordantly to the gene models (Table~\ref{paired-end_map}).
 The results suggest that the gene models are of high quality.
@@
 confirm that the splice junctions are in transcripts.
-Of 112,089 splice junctions from the gene models, only 1,341 (1.2\%) junctions have no spliced reads and 3,447 (3\%)
+Of 104,827 splice junctions from the gene models, only 248 (0.23\%) junctions have no spliced reads and 484 (0.46\%)
 junctions have 1--3 spliced reads (see materials and methods) (Fig.~\ref{cdf_single_splice}.
@@
 \subsubsection*{Comparison with ESTs and Genbank mRNAs}
-83,238 (79.7\%) splice junctions from our gene models are supported by ESTs or Genbank mRNAs.
+82,908 (80.31\%) splice junctions from our gene models are supported by ESTs or Genbank mRNAs.
 This illustrates the efficiency of our pipeline in detecting genuine splice junctions.
-We also found that 46,132 junctions that are supported by both Genbank mRNAs and ESTs are also supported by at least four spliced reads.
+We also found that 82,750 (99.80\%) junctions that are supported by both Genbank mRNAs and ESTs are also supported by at least four spliced reads.
 Therefore, we can assume that splice junctions with more than three spliced reads are likely to be real.
-Using this criterion, we can conclude that our gene models include 21,872 novel splice junctions that are not in ESTs.
+Using this criterion, we can conclude that our gene models include 19,981 novel splice junctions that are not in ESTs and mRNAs.
 
@@
 To evaluate an efficiency of our pipeline in detecting splice junctions,
-we compared the number of splice junctions detected by our pipeline to those from Cufflinks\cite{Trapnell:2010kd}, ESTs and
+we compared the number of splice junctions detected by our pipeline to that from Cufflinks\cite{Trapnell:2010kd}, ESTs and
 Ensembl gene models version 64.
-Cufflinks detected 93,756 (83\%) splice junctions that are also detected by our pipeline.
-However, 46\% (8,615) of junctions not detected by Cufflinks are supported by ESTs.
+Cufflinks detected 89,289 (86.49\%) splice junctions that are also detected by our pipeline.
+However, 37.61\% (5,565) of junctions not detected by Cufflinks are supported by ESTs.
 This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
@@
 We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
-Our pipeline detects 7,765 genes and Cufflinks detects 8,983 genes of total 17,934 Ensembl genes.
-7,049 of those genes are both detectd by our pipeline and Cufflinks.
-In addition, we wanted to compare a number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
+Our pipeline detects 7,046 genes and Cufflinks detects 8,341 genes of total 17,934 Ensembl genes.
+6,572 of those genes are both detectd by our pipeline and Cufflinks.
+In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
 a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
-Therefore, we compare a number of splice junctions detected by both methods within the same gene instead.
-From 7,049 genes, our pipeline detects 1,765 splice junctions, whereas Cufflinks detects 2,516 junctions.
+Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
+From 6,572 genes, our pipeline detects 57,437 splice junctions, whereas Cufflinks detects 62,101 junctions.
 
 However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
-that are not in Ensembl but are supported by ESTs in those 7,049 genes and found that our pipeline detects 2,039 splice junctions
-and Cufflinks only detects 999 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
+that are not in Ensembl but are supported by ESTs in those 6,572 genes and found that our pipeline detects 2,260 splice junctions
+and Cufflinks only detects 1,588 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
 
@@
 However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs.
+
+\subsubsection*{Merged gene models from assembly and Cufflinks}
+As shown above, Cufflinks and assembly method detect non-overlapped splice junctions,
+which suggests that we should to combine gene models from two methods to
+obtain more complete gene models or to detect more splice junctions.
+We obtained a total of 16,445 genes with 53,296 transcripts from Cufflinks + assembly gene models.
+As shown in Table.~\ref{genes_transcripts}, the number of genes from Cufflinks + assembly gene models
+decreases, which indicates that some fragmented transcripts are merged together and form a more complete gene structure (Fig.~\ref{cufflinks_asm}).
 
@@
 protein sequences from incomplete cDNA sequences such as ESTs.
-ESTScan succesfully translated 44,122 (94\%) of all isoforms to protein sequences with 50 or longer amino acids.
+ESTScan succesfully translated 42,459 (94.5\%) of all isoforms to protein sequences with 50 or longer amino acids.
 Then, all protein sequences were searched for homology in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
-We found 27,770 (63\%) isoforms from 4,272 genes match mouse proteins.
+We found 26,931 (63.43\%) isoforms from 4,165 genes match mouse proteins.
 As shown in Fig.~\ref{bitscore}, most genes and isoforms have a bit score/length ratio more than 1.0, which
@@
 Both our pipeline and Cufflinks also detect some splice variants that are not overlapped.
-This suggests that combination of both \emph{de novo} assembly and reference-based assembly is preferred to study alternative splicing in chicken.
+Combination of both methods results in more complete gene models,
+which suggests that combination of \emph{de novo} assembly and reference-based assembly is preferred to study alternative splicing in chicken.
 
@@
 
-In this study, gene models are built from chicken genome version 2.1 (galGal3), which contains a large number of sequence duplications and
+In this study, gene models are built from chicken genome version 2.1 (galGal3),
+which contains a considerable number of sequence duplications and
 missassemblies that supposed to be eliminated in the latest version of genome assembly (galGal4).
@@
 
-\subsection*{Transcript Reassembly}
-To reduce resources and time to reassemble transcripts, we first trimmed off poly-A tail and remove low
-complexity and short sequences using seqclean\cite{seqclean} with default parameters.
-Then we used cd-hit-est from CD-HIT suite\cite{Li:2006hr} to remove all redundant transcripts.
-A large number of transcripts was removed at this step, which facilitates the reassembly process.
-In order to eliminate some sequence variations between samples.
-We mapped sequences to the chicken genome and then extracted all sequences from the genome.
-All transcripts from each chromosome were then assembled by CAP3\cite{Huang:1999wb}.
+A poly-A tail, short transcripts and transcripts with low complexity are removed by
+seqclean\cite{seqclean} with default parameters.
+Redundant transcripts are removed by cd-hit-est from CD-HIT suite\cite{Li:2006hr}.
+A large number of transcripts are removed at this step, which facilitates gene models construction process.
+
+We obtained 334,475 transcripts, of which 295,588 transcripts (88.37\%) mapped to chicken genome.
+Only transcripts mapped to chicken genome are used to build gene models.
 
@@
 \hline
-Line 6 uninfected & 2,978 & 117 & 1,956 (52.76\%) & 41 (35.04\%)\\
-Line 6 infected & 3,524 & 236 & 1,991 (56.50\%)& 100 (42.37\%)\\
-Line 7 uninfected & 3,334 & 163 & 1,933 (57.98\%) & 86 (52.76\%)\\
-Line 7 infected & 3,235 & 82 & 1,858 (57.43\%)& 62 (75.61\%)\\
+Line 6 uninfected & 2,132 & 104 & 1,322 (62.01\%) & 39 (37.50\%)\\
+Line 6 infected & 2,499 & 104 & 1,514 (60.58\%)& 40 (38.46\%)\\
+Line 7 uninfected & 2,633 & 136 & 1,560 (59.25\%) & 52 (38.24\%)\\
+Line 7 infected & 2,409 & 152 & 1,390 (57.70\%)& 50 (32.89\%)\\
 \hline
@@
 \hline
-Oases-M assembly & 114,336 & 8,510 & 391 (4.59\%) \\
-Unmerged assembly & 116,573 & 10,747 & 1,752 (16.30\%) \\
+Oases-M assembly & 114,336 & 8,735 & 447 (5.1\%) \\
+Unmerged assembly & 115,159 & 9,558 & 1,640 (17.15\%) \\
 \hline
@@
 \hline
-Method& Gene & Transcript \\ 
+Method& Gene & Isoform \\ 
 \hline
-Global & 14,832 & 32,311 \\
-Local & 15,297 & 23,028 \\
-Global + Local & 15,934 & 46,797 \\
-Cufflinks & 30,235 & 37,967 \\
+Global & 14,703 & 28,406 \\
+Local & 14,931 & 22,127 \\
+Global + Local & 15,411 & 35,598 \\
+Cufflinks & 25,318 & 34,959 \\
+Global + Local + Cufflinks & 16,445 & 53,296\\
 Ensembl & 17,934 & 23,392 \\
@@
 \hline
-Line 6 uninfected & 18,375,966 (77.93\%) & 5,203,586 (22.07\%) \\
-Line 6 infected & 17,160,695 (73.18\%) & 6,288,286 (26.82\%) \\
-Line 7 uninfected & 18,130,072 (75.77\%) & 5,795,737 (24.22\%) \\
-Line 7 infected & 19,912,046 (78.21\%) & 5,450,521 (21.49\%) \\
+Line 6 uninfected & 18,398,569 (78.03\%) & 5,180,983 (21.97\%) \\
+Line 6 infected & 17,194,894 (73.33\%) & 6,254,087 (26.67\%) \\
+Line 7 uninfected & 18,148,112 (75.85\%) & 5,777,697 (24.15\%) \\
+Line 7 infected & 19,914,624 (78.52\%) & 5,447,943 (21.48\%) \\
 \hline
@@
 \hline
-Line 6 uninfected & 21,598,218 (64.16\%) & 12,065,659 (35.84\%) \\
-Line 6 infected & 15,274,638 (63.89\%) & 15,274,638 (36.11\%) \\
-Line 7 uninfected & 20,961,033 (63.67\%) & 20,961,033 (36.33\%) \\
-Line 7 infected & 22,485,833 (65.22\%) & 22,485,833 (34.78\%) \\
+Line 6 uninfected & 21,641,202 (77.65\%) & 12,022,675 (22.35\%) \\
+Line 6 infected & 18,556,277 (77.61\%) & 5,352,216 (22.39\%) \\
+Line 7 uninfected & 25,670,333 (77.97\%) & 7,250,999 (22.03\%) \\
+Line 7 infected & 26,988,536 (78.27\%) & 7,489,299 (21.73\%) \\
 \hline
@@
 \caption{
-{\bf Distribution of a number of isoforms in each gene from 1) global and 2) local assembly}
+{\bf Distribution of the number of isoforms in each gene from 1) global and 2) local assembly}
 }
@@
 \begin{center}
-\includegraphics[width=5in]{genes_bitscore.pdf}
-\includegraphics[width=5in]{isoforms_bitscore.pdf}
+\includegraphics[width=5in]{bitscore.pdf}
 \end{center}
@@
 {\bf Histogram of bit score/length ratio of isoforms and genes that match mouse proteins.}
-Genes in chromosome E64\_random are not included.
+Genes and isoforms in chromosome E64\_random are not included.
 }
@@
 \caption{
-{\bf Cumulative counts of splice junctions with spliced reads.}
+{\bf Cumulative counts of splice junctions with spliced reads up to 1000 reads.}
 }
@@
 \begin{center}
-\includegraphics[width=7in]{alt_splice_site.pdf}
+\includegraphics[width=5in]{asm-cufflinks.pdf}
+\end{center}
+\caption{
+{\bf Cufflinks + Assembly gene models.}
+Fragmented transcripts in Cufflinks models are merged together after merging with gene models from assembly.
+An isoform from assembly that is not in Cufflinks is also added to the gene model.
+}
+\label{cufflinks_asm}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{alt_splice_site.pdf}
 \end{center}

### 2012-10-10 rnaseq_paper.tex
updated rnaseq_paper.tex

@@
 shown in Table~\ref{Oases-M}.
-Only 447 unique splice junctions from Oases-M are supported by ESTs whereas unmerged assembly detects 1,640
+Only 421 unique splice junctions from Oases-M are supported by ESTs whereas unmerged assembly detects 1,607
 unique splice junctions that also found in ESTs.
@@
 As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
-Total of 14,703 genes were obtained from global assembly and 14,931 genes from local assembly.
-However, the number of genes obtained from global + local assembly is only 15,411.
+Total of 20,439 genes were obtained from global assembly and 21,963 genes from local assembly.
+However, the number of genes obtained from global + local assembly is only 20,992.
 The number of genes and isoforms slightly increased after combining transcripts from global and local
@@
 
-Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
+Most transcripts from RNA-Seq assembly include untranslated regions (UTRs), especially $3'$ UTR in our datasets.
 These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
-RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).
+RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{}.
+Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative $5'$UTRs and extended $3'$UTR respectively.
 
-\subsection*{An overestimated number of genes and transcripts}
+\subsection*{Overestimation of the number of genes and transcripts}
 The number of genes from our pipeline might be overestimated due to fragmented transcripts.
@@
 A bias of read coverage toward $3'$ end and low expression level result in fragmented transcripts as shown in Fig.~\ref{fragmented_transcripts}.
-Moreover, spurious splice junctions due to duplication in genome sequence can lead to a large number of dubious isoforms.
-This seems to occur in a few genome sequences such as chromosome E64\_random.
-Figure~\ref{isoforms_boxplot} shows that a majority of genes from our gene models have only a few isoforms.
-
-In addition, parameters used in building gene models such as minimum exon/intron size determine how transcripts are merged together, which can increases or decreases the number of isoforms.
-In this study, we adjust the parameters to enhance sensitivity of splice junctions detection that may result in overestimation of the number of isoforms.
 
@@
 
-The results show that our pipeline can be used to detect alternative splicing and construct gene models from a large set of transcripts from assembly.
+The results show that our pipeline can be used to detect alternative splicing and construct gene models from a large set of transcripts from \emph{de novo} assembly.
 However, many steps in our pipeline can introduce errors to the gene models.
-We performed several methods to validate the gene models as described below.
+Therefore, we needed to perform several methods to validate the gene models as described below.
 
@@
 We anticipated that a large number of reads could align to the gene models.
-We found that 73--78\% of reads mapped to the gene models (Table~\ref{single-end_map}).
-In addition, we mapped paired-end reads that we obtained later from the same samples to
-the gene models and found that 77--78\% of reads mapped concordantly to the gene models (Table~\ref{paired-end_map}).
+We found that about 80-85\% of reads mapped to the gene models (Table~\ref{single-end_map}).
+We also mapped paired-end reads that we obtained later from the same samples to
+the gene models and found that approximately 85\% of reads mapped concordantly to the gene models (Table~\ref{paired-end_map}).
 The results suggest that the gene models are of high quality.
@@
 confirm that the splice junctions are in transcripts.
-Of 104,827 splice junctions from the gene models, only 248 (0.23\%) junctions have no spliced reads and 484 (0.46\%)
+Of 100,411 splice junctions from the gene models, only 95 (0.09\%) junctions have no spliced reads and 448 (0.45\%)
 junctions have 1--3 spliced reads (see materials and methods) (Fig.~\ref{cdf_single_splice}.
-The results indicate the method can detect splice junctions with a very high specificity.
+The results indicate that the method can detect splice junctions with a very high specificity.
 Some splice junctions with fewer than four spliced reads; however, may not be false junctions because of a bias of
@@
 \subsubsection*{Comparison with ESTs and Genbank mRNAs}
-82,908 (80.31\%) splice junctions from our gene models are supported by ESTs or Genbank mRNAs.
+81,838 (81.5\%) splice junctions from our gene models are supported by ESTs or Genbank mRNAs.
 This illustrates the efficiency of our pipeline in detecting genuine splice junctions.
-We also found that 82,750 (99.80\%) junctions that are supported by both Genbank mRNAs and ESTs are also supported by at least four spliced reads.
+We also found that 81,684 (99.8\%) junctions that are supported by both Genbank mRNAs and ESTs are also supported by at least four spliced reads.
 Therefore, we can assume that splice junctions with more than three spliced reads are likely to be real.
-Using this criterion, we can conclude that our gene models include 19,981 novel splice junctions that are not in ESTs and mRNAs.
+Using this criterion, we can conclude that our gene models include 18,184 novel splice junctions that are not in ESTs and mRNAs.
 
-\subsubsection*{Comparison with Cufflinks}
-To evaluate an efficiency of our pipeline in detecting splice junctions,
+\subsubsection*{Comparison with Cufflinks and Ensembl}
+To evaluate efficiency of our pipeline in detecting splice junctions,
 we compared the number of splice junctions detected by our pipeline to that from Cufflinks\cite{Trapnell:2010kd}, ESTs and
-Ensembl gene models version 64.
-Cufflinks detected 89,289 (86.49\%) splice junctions that are also detected by our pipeline.
-However, 37.61\% (5,565) of junctions not detected by Cufflinks are supported by ESTs.
-This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
-genuine splice junctions.
+Ensembl gene models build 64.
 
-We also aligned transcripts from Cufflinks and the pipeline to Ensembl transcripts to compare genes and transcripts detected by
-both methods.
-We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
-Our pipeline detects 7,046 genes and Cufflinks detects 8,341 genes of total 17,934 Ensembl genes.
-6,572 of those genes are both detectd by our pipeline and Cufflinks.
-In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
-a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
-Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
-From 6,572 genes, our pipeline detects 57,437 splice junctions, whereas Cufflinks detects 62,101 junctions.
+Overall, the pipeline and Cufflinks detect 89,403 and 80,790 splice junctions annotated in Ensembl gene models respectively.
+Both methods detect 78,535 junctions in common and detect some non-overlapped splice junctions.
+Cufflinks detects many splice junctions that are not detected by the pipeline, which only finds 2,255 junctions
+in contrast to 10,868 junctions finds by Cufflinks (Fig.~\ref{chick_venn}).
+It has been shown by a previous study that Cufflinks finds more splice variants than \emph{de novo} assembly
+in human and mouse datasets compared to corresponding Ensembl gene models at all expression levels\cite{Schulz:2012je}.
+The results show that it is also the case for chicken Ensembl gene models.
+% Our results show that Cufflinks also detects more splice junctions than \emph{de novo} assembly in chicken Ensembl gene models.
+However, chicken gene models are relatively incomplete compared to those in human and mouse.
+A large number of splice junctions from ESTs (96,755 junctions) and mRNAs (13,641 junctions) are not included in Ensembl gene models.
+Therefore, we needed to evaluate the efficiency of Cufflinks and the pipeline to detect splice junctions not annotated in Ensembl gene models
+in order to assess the quality of gene models built from RNA-Seq.
+Figure~\ref{chick_est_venn} shows that Cufflinks and our pipeline detect the similar number of unique unannotated splice junctions.
+50\% of unique junctions detected by the pipeline are supported by ESTs or mRNAs, while 37\% of unique junctions from Cufflinks are supported by ESTs or mRNAs.
 
-However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
-that are not in Ensembl but are supported by ESTs in those 6,572 genes and found that our pipeline detects 2,260 splice junctions
-and Cufflinks only detects 1,588 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
+To compare the number of genes detected by Cufflinks and the pipeline, we aligned transcripts from both methods to Ensembl transcripts and
+selected only transcripts that match Ensembl transcripts with more than 90\% identity and more than 80\% sequence coverage.
+The pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of total 17,934 genes in Ensembl models.
 
 To conclude, Cufflinks performs better than our pipeline for detecting both genes and splice junctions in Ensembl gene models.
-However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs.
+However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs or mRNAs.
+Although our pipeline does not detect many splice junctions in Ensembl gene models, it is useful for detecting unannotated junctions that may not be detected by Cufflinks.
+Figure~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
 
-\subsubsection*{Merged gene models from assembly and Cufflinks}
-As shown above, Cufflinks and assembly method detect non-overlapped splice junctions,
-which suggests that we should to combine gene models from two methods to
-obtain more complete gene models or to detect more splice junctions.
-We obtained a total of 16,445 genes with 53,296 transcripts from Cufflinks + assembly gene models.
-As shown in Table.~\ref{genes_transcripts}, the number of genes from Cufflinks + assembly gene models
-decreases, which indicates that some fragmented transcripts are merged together and form a more complete gene structure (Fig.~\ref{cufflinks_asm}).
+In addition, the inefficiency of the pipeline to detect annotated junctions may due to incompleteness of chicken Ensembl gene models.
+As shown in figure~\ref{mus_venn}, Cufflinks and the pipeline detects the same amount of non-overlapped splice junctions in mouse Ensembl build 64.
+We expect the pipeline to detect more junctions in Ensembl gene models if more junctions from chicken ESTs and mRNAs are incorporated.
+
+% The pipeline detected 89,289 (86.49\%) splice junctions that are also detected by our Cufflinks.
+% However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported by ESTs.
+% This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
+% genuine splice junctions.
+
+% In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
+% a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
+% Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
+% From 6,572 genes, our pipeline detects 56,232 splice junctions whereas Cufflinks detects 62,150 junctions.
+
+% However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
+% that are not in Ensembl but are supported by ESTs in those 6,798 genes and found that our pipeline detects 5,160 splice junctions
+% and Cufflinks only detects 5,023 splice junctions.
 
@@
 To obtain a protein sequence of each isoform, we translated a cDNA sequence using ESTScan\cite{Iseli:1999vd},
-which is designed to translate
-protein sequences from incomplete cDNA sequences such as ESTs.
-ESTScan succesfully translated 42,459 (94.5\%) of all isoforms to protein sequences with 50 or longer amino acids.
-Then, all protein sequences were searched for homology in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
-We found 26,931 (63.43\%) isoforms from 4,165 genes match mouse proteins.
+which is designed to translate protein sequences from incomplete cDNA sequences such as ESTs.
+ESTScan succesfully translated 22,488 (81.1\%) of all isoforms to protein sequences with 50 or longer amino acids.
+All protein sequences are searched for homology in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
+We found 17,620 (78.35\%) isoforms from 12,399 genes match mouse proteins.
 As shown in Fig.~\ref{bitscore}, most genes and isoforms have a bit score/length ratio more than 1.0, which
@@
 They can be genes or isoforms that are only found in non-mammal vertebrates or novel genes.
+
+\subsection*{Gene models from Assembly, Cufflinks and mRNAs}
+As shown above, Cufflinks and assembly method detect non-overlapped splice junctions,
+which suggests that we should combine gene models from two methods to obtain more complete gene models or to detect more splice junctions.
+Moreover, we can incorporate transcripts from other sources to build more complete gene models.
+In this study, we demonstrate the process by building gene models from \emph{de novo} assembly, Cufflinks and mRNAs.
+A total of 26,726 genes with 43,074 transcripts are obtained from the gene models (Table.~\ref{genes_transcripts}).
+The number of genes and isoforms from assembly + Cufflinks + mRNAs gene models
+increases significantly due to an addition of unique genes and transcripts from three sources (Fig.~\ref{mrna_cuff_gimme}).
 
@@
 We have also developed a pipeline that assembles reconstructed transcripts from multiple k-mers to build putative gene models.
-Overall, a majority of splice variants detected by our pipeline are also detected by Cufflinks.
-Both our pipeline and Cufflinks also detect some splice variants that are not overlapped.
-Combination of both methods results in more complete gene models,
-which suggests that combination of \emph{de novo} assembly and reference-based assembly is preferred to study alternative splicing in chicken.
+Although a majority of splice junctions detected by our pipeline are also detected by Cufflinks, the pipeline also detect some
+splice junctions not detected by Cufflinks with supporting evidences indicating that they are genuine junctions.
+Combining both gene models from both methods results in more complete gene models that is useful for transcriptome analysis.
+We have also shown that the pipeline can be used to improve existing annotations by combining transcripts from different sources
+to build gene models, which should be useful for non-model organisms that do not possess decent annotations.
 
-Interestingly, Cufflinks detects more genes and isoforms included in Ensembl annotations than our pipeline.
-However, our pipeline detects many more splice junctions not included in Ensembl but supported by ESTs.
-As observed in previous study by Schulz \emph{et al.}\cite{Schulz:2012je}, Cufflinks also outperforms \emph{de novo} assembly in
-detecting genes and isoforms in Ensembl annotations in mouse.
-We speculate that Cufflinks possesses a mechanism that is tailored to detect Ensembl gene models.
-
-In this study, gene models are built from chicken genome version 2.1 (galGal3),
-which contains a considerable number of sequence duplications and
-missassemblies that supposed to be eliminated in the latest version of genome assembly (galGal4).
-Duplications and misassemblies lead to false splice junctions, which in turn produce a large number of splice variants as
-observed in some chromosomes.
-Conversely, this suggests that one might be able to use the pipeline to evaluate the quality of reference genome.
+In addition, the pipeline employs several existing tools that perform different tasks independently.
+Those tools can be replaced by similar tools that more suitable for organisms or datasets.
 
@@
 \section*{Materials and Methods}
+
+\subsection*{Data}
+Mouse RNA-Seq dataset (SRX062280) is downloaded from Short Read Archives (SRA).
+Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from spleen of chicken line 6 and 7.
 
@@
 
-Single and paired-end reads were mapped to chicken genome by Tophat2\cite{Trapnell:2009dp} release 2.0.0 using
-default parameters without annotations.
+Single reads were mapped to chicken genome by Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without annotations.
 All reads were mapped to cDNA sequences derived from gene models by Bowtie2\cite{Langmead:2009fv} with
-default parameters \texttt{(n=2, l=28, e=70, k=1)}.
+default parameters \texttt{(n=2, l=28, e=70, k=20)}.
+Reads from mouse dataset are mapped to mouse genome (mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
 
@@
 all splice junctions and untranslated regions (UTRs).
-After that, transcripts that are identical are clustered and removed by CD-HIT version 4.5.6\cite{Li:2006hr}.
-Parameters for clustering are \texttt{-n 10, -r 0 and -c 1.0}.
+After that, transcripts that are highly similar ($>99\%$) are clustered and removed by CD-HIT version 4.5.6\cite{Li:2006hr}.
+Parameters for clustering are \texttt{-n 10, -r 0 and -c 0.99}.
 Only a representative of each cluster is kept in gene models.
@@
 Transcripts were then grouped into the same gene if they have at least one intron or exon in common.
-Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing paths in the splice graph.
+Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing paths in a splice graph.
+Gimme is able to report both a maximum and a minimum number of isoforms.
+Maximal isoforms are from traversing all simple paths in a splice graph; whereas minimal isoforms are searched by
+a maximum flow algorithm (implemented in Networkx package).
 Gimme is open-source and available at \texttt{https://github.com/ged-lab/gimme}.
@@
 
-Protein sequences translated from each isoform using ESTScan were searched against mouse reference proteins by BLAST 2.2.25+\cite{Tatusova:1999tz}.
-A bit score to a length ratio was calculated for each hit that had an e-value $\le 10^{-20}$.
-Only the highest value of all isoforms from each gene was shown in the gene plot; whereas, values of all isoforms were shown in the isoform plot.
+Protein sequences translated from each isoform using ESTScan were searched against mouse proteins by BLAST 2.2.25+\cite{Tatusova:1999tz}.
+A bit score to a length ratio is calculated for each hit that has an e-value $\le 10^{-20}$.
+The highest value of all isoforms from each gene is shown in the gene plot; whereas, values of all isoforms are shown in the isoform plot.
 
@@
 \hline
-Oases-M assembly & 114,336 & 8,735 & 447 (5.1\%) \\
-Unmerged assembly & 115,159 & 9,558 & 1,640 (17.15\%) \\
+Oases-M assembly & 111,237 & 6,860 & 421 (6.1\%) \\
+Unmerged assembly & 112,708 & 8,295 & 1,607 (19.4\%) \\
 \hline
@@
 \hline
-Global & 14,703 & 28,406 \\
-Local & 14,931 & 22,127 \\
-Global + Local & 15,411 & 35,598 \\
+Global & 20,439 & 25,859 \\
+Local & 21,963 & 25,859 \\
+Global + Local & 20,922 & 27,732 \\
 Cufflinks & 25,318 & 34,959 \\
-Global + Local + Cufflinks & 16,445 & 53,296\\
+Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\
 Ensembl & 17,934 & 23,392 \\
@@
 \hline
-Line 6 uninfected & 18,398,569 (78.03\%) & 5,180,983 (21.97\%) \\
-Line 6 infected & 17,194,894 (73.33\%) & 6,254,087 (26.67\%) \\
-Line 7 uninfected & 18,148,112 (75.85\%) & 5,777,697 (24.15\%) \\
-Line 7 infected & 19,914,624 (78.52\%) & 5,447,943 (21.48\%) \\
+Line 6 uninfected & 20,169,993 (85.54\%) & 3,409,559 (14.46\%) \\
+Line 6 infected & 18,790,831 (80.13\%) & 4,658,150 (19.87\%) \\
+Line 7 uninfected & 19,844,293 (82.94\%) & 4,081,516 (17.06\%) \\
+Line 7 infected & 21,772,102 (85.84\%) & 3,590,465 (14.16\%) \\
 \hline
@@
 \hline
-Line 6 uninfected & 21,641,202 (77.65\%) & 12,022,675 (22.35\%) \\
-Line 6 infected & 18,556,277 (77.61\%) & 5,352,216 (22.39\%) \\
-Line 7 uninfected & 25,670,333 (77.97\%) & 7,250,999 (22.03\%) \\
-Line 7 infected & 26,988,536 (78.27\%) & 7,489,299 (21.73\%) \\
+Line 6 uninfected & 28696112 (85.36\%) & 4967765 (14.76\%) \\
+Line 6 infected & 20514438 (85.93\%) & 3394055 (14.20\%) \\
+Line 7 uninfected & 28159776 (85.65\%) & 4761556 (14.46\%) \\
+Line 7 infected & 29564592 (85.88\%) & 4913243 (14.25\%) \\
 \hline
@@
 {\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
-Transcripts are aligned to a chicken genome by BLAT. Gimme then constructs gene models based on alignments of transcripts.
+Transcripts are aligned to a chicken genome by BLAT\@. Gimme then constructs gene models based on alignments of transcripts.
 }
@@
 \begin{center}
-\includegraphics[width=5in]{isoforms_boxplot.pdf}
-\end{center}
-\caption{
-{\bf Distribution of the number of isoforms in each gene from 1) global and 2) local assembly}
-}
-\label{isoforms_boxplot}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
 \includegraphics[width=5in]{fragmented_transcripts.pdf}
@@
 \caption{
-{\bf Comparison of gene models from our pipeline and other public gene models on UCSC genome browser.}
+{\bf Comparison of gene models from the \emph{de novo} assembly pipeline with reference and Ensembl gene models on UCSC genome browser.}
 }
 \label{model_comparisons}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=6in]{alter_5utr.pdf}
+\end{center}
+\caption{
+{\bf Examples of alternative $5'$ UTRs in RNA-Seq gene models.}
+}
+\label{alter_5utr}
 \end{figure}
@@
 \caption{
-{\bf Examples of an extended $3'$ UTR detected in RNA-Seq gene models.}
+{\bf Examples of an extended $3'$ UTR in RNA-Seq gene models.}
 }
@@
 {\bf Histogram of bit score/length ratio of isoforms and genes that match mouse proteins.}
-Genes and isoforms in chromosome E64\_random are not included.
 }
@@
 \begin{center}
-\includegraphics[width=5in]{asm-cufflinks.pdf}
+\includegraphics[width=5in]{chick_venn.pdf}
 \end{center}
 \caption{
-{\bf Cufflinks + Assembly gene models.}
-Fragmented transcripts in Cufflinks models are merged together after merging with gene models from assembly.
-An isoform from assembly that is not in Cufflinks is also added to the gene model.
+{\bf Splice sites in chicken Ensembl gene models detected by Cufflinks and the \emph{de novo} assembly pipeline.}
+Cufflinks detects many annotated isoforms that are not detected by the pipeline.
+The figure also shows that both methods detect a large number of unannotated splice junctions, which suggests
+that those junctions may be genuine.
 }
-\label{cufflinks_asm}
+\label{chick_venn}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{chick_est_venn.pdf}
+\end{center}
+\caption{
+{\bf Splice sites in Chicken ESTs detected by Cufflinks and the \emph{de novo} assembly pipeline.}
+In contrast to Ensembl gene models, the pipeline and Cufflinks detects the similar number of splice junctions
+from ESTs and mRNAs. This indicates that the pipeline is as efficient as Cufflinks at detecting
+non-annotated splice junctions.
+}
+\label{chick_est_venn}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{mus_venn.pdf}
+\end{center}
+\caption{
+{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de novo} assembly pipeline.}
+In contrast to chicken datasets, Cufflinks and the pipeline detects the similar amount of non-overlapped annotated junctions.
+This may due to a higher quality of gene annotations in mouse.
+}
+\label{mus_venn}
 \end{figure}
@@
 {\bf Unannotated alternative splice site.}
-The pipeline can detect alternative splice site that is not annotated in Ensembl and Cufflinks.
+The pipeline detects alternative splices site not annotated in Ensembl and Cufflinks but are supported by ESTs.
 }
@@
 
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{mrna_cuff_gimme.pdf}
+\end{center}
+\caption{
+{\bf mRNAs + Cufflinks + Assembly gene models.}
+Gimme can combine transcripts from different sources to build gene models.
+In this figure, the final gene model includes several isoforms not annotated in the reference gene model.
+}
+\label{mrna_cuff_gimme}
+\end{figure}
+
 \end{document}

### 2012-10-12 rnaseq_paper.tex
updated CTB friday

--- 2012-08-28 rnaseq_paper.tex
@@
 
-\subsection*{Transcripts from Global and Local Assembly}
-
-Without high-quality annotations, it is necessary to construct gene models with putative isoforms to identify alternative splicing in our samples.
-We used Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je} assembler to construct transcripts from short reads.
-We ran Velvet using default settings except a hash length or k-mer length, which has no default values.
-It has been shown that transcripts from a combination of multiple k-mer lengths is superior than a single k-mer
-length\cite{Schulz:2012je}.
-In addition, we found that different k-mer lengths detected different isoforms from the same dataset
-(Fig.~\ref{kmer-variance}).
-For that reason, it is necessary to cover a wide range of k-mers to discover as many isoforms as possible.
-For this study, we selected a range of kmers from $21-31$ since we found that k-mers outside this range are prone
-to introduce assembly errors.
-
 \subsection*{Local Assembly Enhances Isoform Detection}
 
-Performing assembly with multiple k-mers is time and memory consuming.
-To facilitate this, we partitioned reads into groups based on their alignments to the chicken genome.
-We refer to this method as a local assembly and a conventional assembly method as a global assembly in this study (see materials and methods).
-The advantages of the local assembly method is that each group of reads can be assembled using small amount of memory and it can be assembled on different computers.
-Unexpectedly, we also found that local assembly detected alternative isoforms not found in global assembly.
-As a result, we merged transcripts from both global and local assembly from multiple k-mers to obtain all possible alternative isoforms from our samples.
-Figure~\ref{global_vs_local} shows an example of different isoforms detected from two assembly methods.
+We used the Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je}
+assemblers to construct transcript fragments from four entire Illumina
+GAII mRNAseq data sets sequenced from chicken spleen (see Methods and
+Materials).  In the assembly, we used multiple distinct k-mer values
+for Velvet because different isoforms are recovered for different $k$
+\cite{Schulz:2012je}.  We chose k-mers between 21 and 31 for this
+study, and recovered between 340,000 and 370,000 unique sequences from
+each data set (see Table~\ref{unique_sequences}, and Materials and
+Methods).  These unique sequences represented an unknown number of
+true genes, due to fragmentation from low coverage and incomplete
+assembly.
 
-Using global and local assembly methods, we obtained the different number of contigs.
-Table~\ref{unique_sequences} shows the number of contigs from four datasets and unique contigs from global and local assemblies.
-As expected, some sequences from local assembly (1,690--2,751) are not found in global assembly.
-Homology search in mouse proteins showed that 32--38\% of long unique regions ($>300$ bp) from local assembly match protein
-coding sequences (Table~\ref{unique_sequences_matched_mouse}).
-This suggests that some isoforms will be missing in the gene models based solely on global assembly.
-In addition, a significant number of transcripts from global assembly are unique, which indicates that
-approximately at least 3\% of coding sequences are missing in a current chicken genome (sequences not found in local assembly).
+We next used Tophat to align mRNAseq reads to the genome and partition
+reads by chromosome; we then assembled the partitioned reads with
+Velvet and Oases using the same range of parameters as the global
+assembly, above.  While the local assemblies were considerably more
+computationally efficient, they lacked several thousand unique regions
+that were present in the global assembly
+(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
+the incomplete nature of the current chicken genome (cite).
+Interestingly, over a hundred regions were present {\em only} in the
+local assemblies, suggesting that the local assemblies might be
+recovering additional exons.  Significant numbers of unique regions
+from both global and local assemblies showed homology to the mouse
+genome, indicating that at least some of these unique sequences
+represented real sequence content.  Figure~\ref{global_vs_local} shows
+an example of different isoforms detected by the two assembly methods.
 
-\subsection*{Transcripts from Oases-M contain fewer splice junctions}
-Transcripts from multiple k-mer lengths can be merged using Oases-M.
-It has been demonstrated that merged transcripts are more complete than those from any single kmer length\cite{Schulz:2012je}.
-However, sensitivity of Oases-M in detecting alternative isoforms has not been fully evaluated.
-To evaluate this, we compared transcripts from global assembly with k-mer lenght from 21--31 and merged transcripts from Oases-M using k-mer=27. We used alignments of ESTs as a control.
-The results show that Oases-M detect fewer splice junctions than unmerged global assembly.
-The number of total splice junctions and unique splice junctions detected by Oases-M and unmerged assembly is
-shown in Table~\ref{Oases-M}.
-Only 447 unique splice junctions from Oases-M are supported by ESTs whereas unmerged assembly detects 1,640
-unique splice junctions that also found in ESTs.
-This suggests that some genuine splice junctions are missing from merging transcripts from multiple k-mer
-lengths with Oases-M; therefore, we did not use Oases-M in our pipeline.
+% @CTB how many reads are discarded in the process of mapping?
+% @CTB what does local assembly actually do to the DBG??
+% @CTB what happens if you ask how many unique regions there are not
+%   represented in the final gene models?
 
-\subsection*{Gene models prediction using Gimme}
+\subsection*{Oases-M discards splice variants}
 
-We developed a program called, Gimme, to constructs a gene model from alignments of transcripts to the genome.
+To construct a more comprehensive gene set containing all of the
+assembled contigs, we tried using Oases-M to merge the assemblies from
+multiple $k$ values \cite{Schulz:2012je}.  While it has been
+demonstrated that merged transcripts from multiple-k assemblies
+contain more isoforms than those from any single $k$, the sensitivity
+of Oases-M for recovering splice variants has not been fully
+evaluated.  We merged transcripts from our global assembly, above,
+with Oases-M using a k-mer size of 27, and then compared them with the
+unmerged transcripts.  We then cross-validated using publicly
+available ESTs for an independent data set.  The results show that
+Oases-M and the unmerged assembly share about 95\% of the predicted,
+splice junctions, with about 5\% disjoint.  Of these 5\%,
+approximately 6\% of the Oases-M-specific splice junctions are
+independently supported by ESTs, while 19\% of the unmerged splice
+junctions are supported by ESTs.  This suggests that Oases-M probably
+discards a number of real splice variants, although the unmerged
+assembly is also probably missing some found by Oases-M.
+
+% @CTB why would Oases-M create ``good'' splice variants from the unmerged assembly?
+% @CTB does the math for table three work out here?
+% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a general merging tool?
+
+\subsection*{Exon graphs can be used to reconstruct putative splice variants}
+
+We developed an exon graph approach to construct gene models from
+alignments of transcripts against the genome.  Our approach, called
+Gimme, uses exon graphs to merge transcripts and gene models based on
+overlapping exons (see Materials and Methods).  We used Gimme to
+obtain 20,439 gene models containing 25,859 isoforms from our global
+assembly; 21,963 gene models containing 25,859 XX from our local
+assembly; and 20,922 gene models containing 27,732 isoforms from the
+merged global and local assemblies (Table~\ref{genes_transcripts}).
+
+(more here - what? @CTB)
+
+\subsection*{Validation of Gene Models}
+
+The results show that our pipeline predicts many gene models and
+isoforms after assembly.  We validated these gene models in several
+different ways.
+
+\subsubsection*{The gene models include most reads}
+
+We used bowtie to map single-end reads from the source datasets to the
+transcripts.  More than 80\% of the original reads could be aligned to
+the transcripts, demonstrating that we did not lose a significant
+amount of information during the merge process
+(Table~\ref{single-end_map}).  More importantly, we also mapped
+paired-end reads from technical replicates to the same gene models,
+and found that more than 85\% of the paired-ends mapped concordantly
+the gene models (Table~\ref{paired-end_map}).  Most of the reads that
+did not map were either highly erroneous or contained low-complexity
+artifactual sequence that probably came from the reverse transcription
+procedure (Cite).
+% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
+% paper?  How about a citation that suggests that there's a lot of crap
+% in mRNAseq?
+Thus the merged gene models produced by Gimme represent the significant
+majority of the assemblable data.
+
+\subsubsection*{Almost all splice junctions have high coverage}
+
+To validate the splice junctions reconstructed by the Gimme pipeline,
+we used Bowtie to map mRNAseq reads to the transcript sequences
+derived from gene models Bowtie\cite{Langmead:2009fv}.  Because the
+Velvet/Oases assembly pipeline does not make use of the reference
+genome, reads that map across a splice junction constitute independent
+verification of a splice junction's presence in a transcript.
+
+Of 104,827 splice junctions from the gene models, 248 (0.23\%)
+junctions have no spliced reads and only 484 (0.46\%) junctions
+have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
+More than 99\% of our predicted splice junctions have a coverage
+of 4 or higher in our combined mRNAseq data sets, suggesting that they are
+truly present within the data set.  Note that lower coverage splice junctions
+may still be correct, but may be from low-expressed isoforms or 
+regions of the gene that are low coverage due to sample preparation issues.
+
+\subsubsection{Most splice junctions are independently supported}
+
+Of the 104,827 splice junctions in our gene models, 82,908 (80.31\%)
+are supported by ESTs or mRNAs from Genbank.  This is especially
+surprising since our mRNAseq data is from spleen, and few of the
+publicly available ESTs or mRNAs are from spleen.  Note that this
+cross-validation suggests that the 19,981 novel splice junctions {\em not}
+seen in publicly available ESTs and mRNAs are also likely to be real
+splice junctions from spleen.
+% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?
+
+\subsubsection*{Our pipeline improves on existing reference-based approaches}
+
+We next compared the Gimme gene models to those produced by Cufflinks,
+another reference-based approach to building gene models from mRNAseq
+data\cite{Trapnell:2010kd}.  We also compared both methods to the ENSEMBL gene annotations,
+which are produced by a pipeline that incorporates de novo gene
+prediction and homology-based approaches as well as expression data.
+
+Cufflinks finds 109,641 splice junctions, and Gimme finds 100,411
+splice junctions.  88,289 of them are in common.  (Venn diagram.)
+
+Both Cufflinks and Gimme find approximately 40-50\% of the genes and
+80-90\% of the splice junctions present in the ENSEMBL gene models
+for chicken.  Cufflinks performs about 10\% better in both cases
+than Gimme, demonstrating that Cufflinks has higher sensitivity
+in recovering ENSEMBL gene models.
+
+The ENSEMBL pipeline does not, however, include a large number of
+splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
+Gimme each recover about 10\% of these, with more than 2/3 of these
+recovered by both.  This indicates that both Gimme and Cufflinks
+are equally adept at recovering novel splice junctions.
+% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
+% gene models is false...
+% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?
+
+When we apply Gimme and Cufflinks to a publicly available mouse
+mRNAseq data set, Gimme and Cufflinks recover approximately the same
+number of splice junctions already known from ENSEMBL.  However, Gimme
+recovers a substantial number of additional splice variants beyond
+Cufflinks and ENSEMBL both.
+
+% @CTB is the mouse data set from the trinity paper?
+% @CTB are most of these splice junctions > 3 coverage?
+
+\subsubsection*{Gimme can iteratively merge sets of gene models}
+
+% @CTB may want to distinguish between Gimme and the assembly method.
+
+As shown above, Cufflinks and the assembly method detect a number of
+distinct but equally valid splice junctions, which suggests that we
+could obtain greater sensitivity by merging both sets of predictions.
+We therefore used Gimme to merge the Cufflinks and assembly gene
+models Table.~\ref{genes_transcripts}.  This resulted in a decreased
+number of total genes, suggesting that some fragmented genes were
+merged together to form more complete gene structures (e.g. see
+Fig.~\ref{cufflinks_asm}).
+% @CTB what about splice junctions etc??
+
+\subsubsection*{Validating chicken sequences by using mouse homologs}
+
+To validate our predicted isoforms, we extracted coding sequences from
+our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
+successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
+protein sequences with 50 or more amino acids.  We then searched for
+homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
+of our isoforms from 12,399 genes match mouse proteins.
+Most of these matches have a bit-score/length ratio greater than 1,
+which indicates a good agreement between chicken and mouse proteins.
+% @CTB why so few, really?
+
+\subsection{Leftover text}
+
 The program can merge all transcripts from multiple-kmers and predict a structure of full-length isoforms and genes.
@@
 
-\subsection*{An overestimated number of genes and transcripts}
-The number of genes from our pipeline might be overestimated due to fragmented transcripts.
-In our datasets, read coverage is lower at the $5'$ end due to the method used to convert mRNA to cDNA
-in library preparation.
-A bias of read coverage toward $3'$ end and low expression level result in fragmented transcripts as shown in Fig.~\ref{fragmented_transcripts}.
-Moreover, spurious splice junctions due to duplication in genome sequence can lead to a large number of dubious isoforms.
-This seems to occur in a few genome sequences such as chromosome E64\_random.
-Figure~\ref{isoforms_boxplot} shows that a majority of genes from our gene models have only a few isoforms.
+\section*{An overestimated number of genes and transcripts}
 
-In addition, parameters used in building gene models such as minimum exon/intron size determine how transcripts are merged together, which can increases or decreases the number of isoforms.
-In this study, we adjust the parameters to enhance sensitivity of splice junctions detection that may result in overestimation of the number of isoforms.
+The number of genes from our pipeline might be overestimated due to
+fragmented transcripts.  In our datasets, read coverage is lower at
+the $5'$ end due to the method used to convert mRNA to cDNA in library
+preparation.  A bias of read coverage toward $3'$ end and low
+expression level result in fragmented transcripts as shown in
+Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
+junctions due to duplication in genome sequence can lead to a large
+number of dubious isoforms.  This seems to occur in a few genome
+sequences such as chromosome E64\_random.
+Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
+gene models have only a few isoforms.
 
-\subsection*{Validation of Gene Models}
+In addition, parameters used in building gene models such as minimum
+exon/intron size determine how transcripts are merged together, which
+can increases or decreases the number of isoforms.  In this study, we
+adjust the parameters to enhance sensitivity of splice junctions
+detection that may result in overestimation of the number of isoforms.
 
-The results show that our pipeline can be used to detect alternative splicing and construct gene models from a large set of transcripts from assembly.
-However, many steps in our pipeline can introduce errors to the gene models.
-We performed several methods to validate the gene models as described below.
+\subsubsection{Cufflinks}
 
-\subsubsection*{Reads Mapped to Gene Models}
-
-Single-end reads from the same datasets that we used to build the gene models were aligned to cDNA sequences derived from the gene models.
-We anticipated that a large number of reads could align to the gene models.
-We found that 73--78\% of reads mapped to the gene models (Table~\ref{single-end_map}).
-In addition, we mapped paired-end reads that we obtained later from the same samples to
-the gene models and found that 77--78\% of reads mapped concordantly to the gene models (Table~\ref{paired-end_map}).
-The results suggest that the gene models are of high quality.
-
-\subsubsection*{Reads mapped to splice junctions}
-To validate the splice junctions, we mapped reads to cDNA sequences derived from gene models using Bowtie\cite{Langmead:2009fv}.
-Because the mapping algorithm implemented in Bowtie does not consider splice sites, reads mapped across splice junctions
-confirm that the splice junctions are in transcripts.
-Of 104,827 splice junctions from the gene models, only 248 (0.23\%) junctions have no spliced reads and 484 (0.46\%)
-junctions have 1--3 spliced reads (see materials and methods) (Fig.~\ref{cdf_single_splice}.
-The results indicate the method can detect splice junctions with a very high specificity.
-Some splice junctions with fewer than four spliced reads; however, may not be false junctions because of a bias of
-read coverage toward $3'$ end in our datasets as mentioned earlier.
-Reads are less abundant at the $5'$ end of transcripts; consequently, a relatively small number of spliced reads is mapped
-to splice junctions near the $5'$ end.
-
-\subsubsection*{Comparison with ESTs and Genbank mRNAs}
-82,908 (80.31\%) splice junctions from our gene models are supported by ESTs or Genbank mRNAs.
-This illustrates the efficiency of our pipeline in detecting genuine splice junctions.
-We also found that 82,750 (99.80\%) junctions that are supported by both Genbank mRNAs and ESTs are also supported by at least four spliced reads.
-Therefore, we can assume that splice junctions with more than three spliced reads are likely to be real.
-Using this criterion, we can conclude that our gene models include 19,981 novel splice junctions that are not in ESTs and mRNAs.
-
-\subsubsection*{Comparison with Cufflinks}
 To evaluate an efficiency of our pipeline in detecting splice junctions,
-we compared the number of splice junctions detected by our pipeline to that from Cufflinks\cite{Trapnell:2010kd}, ESTs and
+we compared the number of splice junctions detected by our pipeline to that from , ESTs and
 Ensembl gene models version 64.
@@
 
-\subsubsection*{Merged gene models from assembly and Cufflinks}
-As shown above, Cufflinks and assembly method detect non-overlapped splice junctions,
-which suggests that we should to combine gene models from two methods to
-obtain more complete gene models or to detect more splice junctions.
-We obtained a total of 16,445 genes with 53,296 transcripts from Cufflinks + assembly gene models.
-As shown in Table.~\ref{genes_transcripts}, the number of genes from Cufflinks + assembly gene models
-decreases, which indicates that some fragmented transcripts are merged together and form a more complete gene structure (Fig.~\ref{cufflinks_asm}).
+\subsubsection{isoforms and mouse}
 
-\subsubsection*{Homologous sequences in mouse}
-Splice variants must maintain the reading frame to be translated to functional proteins.
-Based on this assumption, we investigated isoforms in our gene models by searching for homologous protein sequences in
-other organism.
-To obtain a protein sequence of each isoform, we translated a cDNA sequence using ESTScan\cite{Iseli:1999vd},
-which is designed to translate
-protein sequences from incomplete cDNA sequences such as ESTs.
-ESTScan succesfully translated 42,459 (94.5\%) of all isoforms to protein sequences with 50 or longer amino acids.
-Then, all protein sequences were searched for homology in mouse \textit{(Mus musculus)}, which is a related and well-studied animal.
-We found 26,931 (63.43\%) isoforms from 4,165 genes match mouse proteins.
-As shown in Fig.~\ref{bitscore}, most genes and isoforms have a bit score/length ratio more than 1.0, which
-indicates a good agreement between chicken and mouse proteins.
-The results illustrate that isoforms constructed by our pipeline are translatable and may be functional.
-However, genes or isoforms with no match to mouse proteins are not necessarily artifacts.
-They can be genes or isoforms that are only found in non-mammal vertebrates or novel genes.
+The results illustrate that isoforms constructed
+by our pipeline are translatable and may be functional.  However,
+genes or isoforms with no match to mouse proteins are not necessarily
+artifacts.  They can be genes or isoforms that are only found in
+non-mammal vertebrates or novel genes.
 

### 2012-10-15 rnaseq_paper.tex
merged with Titus's update

@@
 recovering additional exons.  Significant numbers of unique regions
-from both global and local assemblies showed homology to the mouse
-genome, indicating that at least some of these unique sequences
+($>300$bp) from both global and local assemblies showed homology to
+the mouse genome, indicating that at least some of these unique sequences
 represented real sequence content.  Figure~\ref{global_vs_local} shows
@@
 overlapping exons (see Materials and Methods).  We used Gimme to
-obtain 20,439 gene models containing 25,859 isoforms from our global
-assembly; 21,963 gene models containing 25,859 XX from our local
+obtain 20,439 gene models containing 26,035 isoforms from our global
+assembly; 21,963 gene models containing 25,859 from our local
 assembly; and 20,922 gene models containing 27,732 isoforms from the
 merged global and local assemblies (Table~\ref{genes_transcripts}).
+The number of genes is slightly decreased after merging transcripts
+from global and local assembly together. 
+This indicates that some transcripts from both methods were merged together.
+Figure~\ref{model_comparisons} shows gene models built by our method with
+other existing gene models.
 
-(more here - what? @CTB)
+Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
+especially $3'$ UTR in our datasets.
+These regions are challenging to predict correctly solely from computational
+methods due to low degree of conservation (cite).
+RNA-Seq-derived gene models are useful for studying variations within UTR regions,
+which may be involved in regulation of isoform expression (cite).
+Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
+$5'$UTRs and extended $3'$UTR respectively.
 
@@
 
-\subsubsection{Most splice junctions are independently supported}
+\subsubsection*{Most splice junctions are independently supported}
 
@@
 could obtain greater sensitivity by merging both sets of predictions.
-We therefore used Gimme to merge the Cufflinks and assembly gene
-models Table.~\ref{genes_transcripts}.  This resulted in a decreased
-number of total genes, suggesting that some fragmented genes were
-merged together to form more complete gene structures (e.g. see
-Fig.~\ref{cufflinks_asm}).
+We therefore used Gimme to merge mRNAs, the Cufflinks and assembly gene
+models Table.~\ref{genes_transcripts}.  This resulted in an increased 
+number of total genes and isoforms (e.g. see Fig.~\ref{mrna_cuff_gimme.pdf}).
+The merged gene models recover 51\% and 66\% of splice junctions from ESTs
+and ENSENBL respectively, which is 10-15\% greater than that from corresponding
+unmerged gene models.
 % @CTB what about splice junctions etc??
@@
 
-\subsection{Leftover text}
+\subsection*{Leftover text}
 
-The program can merge all transcripts from multiple-kmers and predict a structure of full-length isoforms and genes.
-As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
-Total of 14,703 genes were obtained from global assembly and 14,931 genes from local assembly.
-However, the number of genes obtained from global + local assembly is only 15,411.
-The number of genes and isoforms slightly increased after combining transcripts from global and local
-assembly together (Table~\ref{genes_transcripts}). 
-This indicates that some transcripts from both methods were merged together.
+The program can merge all transcripts from multiple-kmers and predict
+a structure of full-length isoforms and genes.
+Total of 20,439 genes were obtained from global assembly and 21,963 genes
+from local assembly.
+However, the number of genes obtained from global + local assembly is only 20,922 genes.
+The number of isoforms slightly increased after combining
+transcripts from global and local assembly together (Table~\ref{genes_transcripts}).
 
-Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
-These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
-RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).
+Most transcripts include a large part of untranslated regions (UTRs),
+especially $3'$ UTR in our datasets.
+These regions are challenging to predict correctly solely from computational
+methods due to low degree of conservation [?].
+RNA-Seq-derived gene models are useful for studying variations within UTR
+regions, which may be involved in regulation of isoform expression\cite{}
 
@@
 junctions due to duplication in genome sequence can lead to a large
-number of dubious isoforms.  This seems to occur in a few genome
-sequences such as chromosome E64\_random.
-Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
-gene models have only a few isoforms.
+number of dubious isoforms.
 
@@
 
-\subsubsection{Cufflinks}
+\subsubsection*{Cufflinks}
 
 To evaluate an efficiency of our pipeline in detecting splice junctions,
-we compared the number of splice junctions detected by our pipeline to that from , ESTs and
-Ensembl gene models version 64.
-Cufflinks detected 89,289 (86.49\%) splice junctions that are also detected by our pipeline.
-However, 37.61\% (5,565) of junctions not detected by Cufflinks are supported by ESTs.
-This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
-genuine splice junctions.
+we compared the number of splice junctions detected by our pipeline to
+that from , ESTs and Ensembl gene models version 64.
+Cufflinks detected 88,289 (87.93\%) splice junctions that are also detected
+by our pipeline.
+However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported
+by ESTs or mRNAs.
+This suggests that the assembly detects some splice junctions that are
+not detected by Cufflinks and they are genuine splice junctions.
 
-We also aligned transcripts from Cufflinks and the pipeline to Ensembl transcripts to compare genes and transcripts detected by
-both methods.
-We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
-Our pipeline detects 7,046 genes and Cufflinks detects 8,341 genes of total 17,934 Ensembl genes.
-6,572 of those genes are both detectd by our pipeline and Cufflinks.
-In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
-a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
-Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
-From 6,572 genes, our pipeline detects 57,437 splice junctions, whereas Cufflinks detects 62,101 junctions.
+We also aligned transcripts from Cufflinks and the pipeline to Ensembl
+transcripts to compare genes and transcripts detected by both methods.
+We selected only transcripts that match with more than 90\% identity and
+cover more than 80\% of an Ensembl transcript.
+Our pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of
+total 17,934 Ensembl genes.
+6,798 of those genes are both detectd by our pipeline and Cufflinks.
+In addition, we wanted to compare the number of isoforms detected by our
+pipeline and Cufflinks within the same gene; however, a transcript can
+match to multiple isoforms in the same gene, especially when it is not complete.
+Therefore, we compare the number of splice junctions detected by both
+methods within the same gene instead.
+From 6,798 genes, our pipeline detects 56,332 splice junctions, whereas
+Cufflinks detects 62,150 junctions.
 
-However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
-that are not in Ensembl but are supported by ESTs in those 6,572 genes and found that our pipeline detects 2,260 splice junctions
-and Cufflinks only detects 1,588 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
+However, because Ensembl gene models do not include all isoforms or
+splice junctions from ESTs, we investigated splice junctions that are
+not in Ensembl but are supported by ESTs in those 6,798 genes and
+found that our pipeline detects 1,997 splice junctions and Cufflinks
+only detects 1,860 splice junctions.
+Figure ~\ref{alt_splice_site} shows an example of splice junctions not
+included in Ensembl gene models, which are detected by our pipeline.
 
-To conclude, Cufflinks performs better than our pipeline for detecting both genes and splice junctions in Ensembl gene models.
-However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs.
+To conclude, Cufflinks performs better than our pipeline for detecting
+both genes and splice junctions in Ensembl gene models.
+However, we found that our pipeline detects more splice junctions
+that are not in Ensembl but are supported by ESTs.
 
@@
 
+\subsection*{Data}
+Mouse RNA-Seq dataset (SRX062280) is downloaded from Short Read Archives (SRA).
+Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from spleen of chicken line 6 and 7.
+
 \subsection*{Mapping reads to The Reference Genome and Gene Models}
@@
 default parameters \texttt{(n=2, l=28, e=70, k=1)}.
+Reads from mouse dataset were mapped to mouse genome (mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
 
@@
 \hline
-Global & 14,703 & 28,406 \\
-Local & 14,931 & 22,127 \\
-Global + Local & 15,411 & 35,598 \\
+Global & 20,439 & 26,035 \\
+Local & 21,963 & 25,859 \\
+Global + Local & 20,922 & 27,732 \\
 Cufflinks & 25,318 & 34,959 \\
-Global + Local + Cufflinks & 16,445 & 53,296\\
+Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\
 Ensembl & 17,934 & 23,392 \\
@@
 \hline
-Line 6 uninfected & 21,641,202 (77.65\%) & 12,022,675 (22.35\%) \\
-Line 6 infected & 18,556,277 (77.61\%) & 5,352,216 (22.39\%) \\
-Line 7 uninfected & 25,670,333 (77.97\%) & 7,250,999 (22.03\%) \\
-Line 7 infected & 26,988,536 (78.27\%) & 7,489,299 (21.73\%) \\
+Line 6 uninfected & 28,696,112 (85.36\%) & 4,967,765 (14.76\%) \\
+Line 6 infected & 20,514,438 (85.93\%) & 3,394,055 (14.20\%) \\
+Line 7 uninfected & 28,159,776 (85.65\%) & 4,761,556 (14.46\%) \\
+Line 7 infected & 29,564,592 (85.88\%) & 4,913,243 (14.25\%) \\
 \hline

### 2012-10-17 rnaseq_paper.tex

@@
 recovering additional exons.  Significant numbers of unique regions
+from both global and local assemblies showed homology to the mouse
+genome, indicating that at least some of these unique sequences
+represented real sequence content.  Figure~\ref{global_vs_local} shows
+an example of different isoforms detected by the two assembly methods.
+
+% @CTB how many reads are discarded in the process of mapping?
+% @CTB what does local assembly actually do to the DBG??
+% @CTB what happens if you ask how many unique regions there are not
+%   represented in the final gene models?
+
+\subsection*{Oases-M discards splice variants}
+
+To construct a more comprehensive gene set containing all of the
+assembled contigs, we tried using Oases-M to merge the assemblies from
+multiple $k$ values \cite{Schulz:2012je}.  While it has been
+demonstrated that merged transcripts from multiple-k assemblies
+contain more isoforms than those from any single $k$, the sensitivity
+of Oases-M for recovering splice variants has not been fully
+evaluated.  We merged transcripts from our global assembly, above,
+with Oases-M using a k-mer size of 27, and then compared them with the
+unmerged transcripts.  We then cross-validated using publicly
+available ESTs for an independent data set.  The results show that
+Oases-M and the unmerged assembly share about 95\% of the predicted,
+splice junctions, with about 5\% disjoint.  Of these 5\%,
+approximately 6\% of the Oases-M-specific splice junctions are
+independently supported by ESTs, while 19\% of the unmerged splice
+junctions are supported by ESTs.  This suggests that Oases-M probably
+discards a number of real splice variants, although the unmerged
+assembly is also probably missing some found by Oases-M.
+
+% @CTB why would Oases-M create ``good'' splice variants from the unmerged assembly?
+% @CTB does the math for table three work out here?
+% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a general merging tool?
+
+\subsection*{Exon graphs can be used to reconstruct putative splice variants}
+
+We developed an exon graph approach to construct gene models from
+alignments of transcripts against the genome.  Our approach, called
+Gimme, uses exon graphs to merge transcripts and gene models based on
+overlapping exons (see Materials and Methods).  We used Gimme to
+obtain 20,439 gene models containing 25,859 isoforms from our global
+assembly; 21,963 gene models containing 25,859 XX from our local
+assembly; and 20,922 gene models containing 27,732 isoforms from the
+merged global and local assemblies (Table~\ref{genes_transcripts}).
+
+(more here - what? @CTB)
+
+\subsection*{Validation of Gene Models}
+
+The results show that our pipeline predicts many gene models and
+isoforms after assembly.  We validated these gene models in several
+different ways.
+
+\subsubsection*{The gene models include most reads}
+
+We used bowtie to map single-end reads from the source datasets to the
+transcripts.  More than 80\% of the original reads could be aligned to
+the transcripts, demonstrating that we did not lose a significant
+amount of information during the merge process
+(Table~\ref{single-end_map}).  More importantly, we also mapped
+paired-end reads from technical replicates to the same gene models,
+and found that more than 85\% of the paired-ends mapped concordantly
+the gene models (Table~\ref{paired-end_map}).  Most of the reads that
+did not map were either highly erroneous or contained low-complexity
+artifactual sequence that probably came from the reverse transcription
+procedure (Cite).
+% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
+% paper?  How about a citation that suggests that there's a lot of crap
+% in mRNAseq?
+Thus the merged gene models produced by Gimme represent the significant
+majority of the assemblable data.
+
+\subsubsection*{Almost all splice junctions have high coverage}
+
+To validate the splice junctions reconstructed by the Gimme pipeline,
+we used Bowtie to map mRNAseq reads to the transcript sequences
+derived from gene models Bowtie\cite{Langmead:2009fv}.  Because the
+Velvet/Oases assembly pipeline does not make use of the reference
+genome, reads that map across a splice junction constitute independent
+verification of a splice junction's presence in a transcript.
+
+Of 104,827 splice junctions from the gene models, 248 (0.23\%)
+junctions have no spliced reads and only 484 (0.46\%) junctions
+have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
+More than 99\% of our predicted splice junctions have a coverage
+of 4 or higher in our combined mRNAseq data sets, suggesting that they are
+truly present within the data set.  Note that lower coverage splice junctions
+may still be correct, but may be from low-expressed isoforms or 
+regions of the gene that are low coverage due to sample preparation issues.
+
+\subsubsection{Most splice junctions are independently supported}
+
+Of the 104,827 splice junctions in our gene models, 82,908 (80.31\%)
+are supported by ESTs or mRNAs from Genbank.  This is especially
+surprising since our mRNAseq data is from spleen, and few of the
+publicly available ESTs or mRNAs are from spleen.  Note that this
+cross-validation suggests that the 19,981 novel splice junctions {\em not}
+seen in publicly available ESTs and mRNAs are also likely to be real
+splice junctions from spleen.
+% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?
+
+\subsubsection*{Our pipeline improves on existing reference-based approaches}
+
+We next compared the Gimme gene models to those produced by Cufflinks,
+another reference-based approach to building gene models from mRNAseq
+data\cite{Trapnell:2010kd}.  We also compared both methods to the ENSEMBL gene annotations,
+which are produced by a pipeline that incorporates de novo gene
+prediction and homology-based approaches as well as expression data.
+
+Cufflinks finds 109,641 splice junctions, and Gimme finds 100,411
+splice junctions.  88,289 of them are in common.  (Venn diagram.)
+
+Both Cufflinks and Gimme find approximately 40-50\% of the genes and
+80-90\% of the splice junctions present in the ENSEMBL gene models
+for chicken.  Cufflinks performs about 10\% better in both cases
+than Gimme, demonstrating that Cufflinks has higher sensitivity
+in recovering ENSEMBL gene models.
+
+The ENSEMBL pipeline does not, however, include a large number of
+splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
+Gimme each recover about 10\% of these, with more than 2/3 of these
+recovered by both.  This indicates that both Gimme and Cufflinks
+are equally adept at recovering novel splice junctions.
+% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
+% gene models is false...
+% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?
+
+When we apply Gimme and Cufflinks to a publicly available mouse
+mRNAseq data set, Gimme and Cufflinks recover approximately the same
+number of splice junctions already known from ENSEMBL.  However, Gimme
+recovers a substantial number of additional splice variants beyond
+Cufflinks and ENSEMBL both.
+
+% @CTB is the mouse data set from the trinity paper?
+% @CTB are most of these splice junctions > 3 coverage?
+
+\subsubsection*{Gimme can iteratively merge sets of gene models}
+
+% @CTB may want to distinguish between Gimme and the assembly method.
+
+As shown above, Cufflinks and the assembly method detect a number of
+distinct but equally valid splice junctions, which suggests that we
+could obtain greater sensitivity by merging both sets of predictions.
+We therefore used Gimme to merge the Cufflinks and assembly gene
+models Table.~\ref{genes_transcripts}.  This resulted in a decreased
+number of total genes, suggesting that some fragmented genes were
+merged together to form more complete gene structures (e.g. see
+Fig.~\ref{cufflinks_asm}).
+% @CTB what about splice junctions etc??
+
+\subsubsection*{Validating chicken sequences by using mouse homologs}
+
+To validate our predicted isoforms, we extracted coding sequences from
+our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
+successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
+protein sequences with 50 or more amino acids.  We then searched for
+homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
+of our isoforms from 12,399 genes match mouse proteins.
+Most of these matches have a bit-score/length ratio greater than 1,
+which indicates a good agreement between chicken and mouse proteins.
+% @CTB why so few, really?
+
+\subsection{Leftover text}
+
+The program can merge all transcripts from multiple-kmers and predict a structure of full-length isoforms and genes.
+As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
+Total of 14,703 genes were obtained from global assembly and 14,931 genes from local assembly.
+However, the number of genes obtained from global + local assembly is only 15,411.
+The number of genes and isoforms slightly increased after combining transcripts from global and local
+assembly together (Table~\ref{genes_transcripts}). 
+This indicates that some transcripts from both methods were merged together.
+
+Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
+These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
+RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).
+
+\section*{An overestimated number of genes and transcripts}
+
+The number of genes from our pipeline might be overestimated due to
+fragmented transcripts.  In our datasets, read coverage is lower at
+the $5'$ end due to the method used to convert mRNA to cDNA in library
+preparation.  A bias of read coverage toward $3'$ end and low
+expression level result in fragmented transcripts as shown in
+Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
+junctions due to duplication in genome sequence can lead to a large
+number of dubious isoforms.  This seems to occur in a few genome
+sequences such as chromosome E64\_random.
+Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
+gene models have only a few isoforms.
+
+In addition, parameters used in building gene models such as minimum
+exon/intron size determine how transcripts are merged together, which
+can increases or decreases the number of isoforms.  In this study, we
+adjust the parameters to enhance sensitivity of splice junctions
+detection that may result in overestimation of the number of isoforms.
+=======
+
+We used the Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je}
+assemblers to construct transcript fragments from four entire Illumina
+GAII mRNAseq data sets sequenced from chicken spleen (see Methods and
+Materials).  In the assembly, we used multiple distinct k-mer values
+for Velvet because different isoforms are recovered for different $k$
+\cite{Schulz:2012je}.  We chose k-mers between 21 and 31 for this
+study, and recovered between 340,000 and 370,000 unique sequences from
+each data set (see Table~\ref{unique_sequences}, and Materials and
+Methods).  These unique sequences represented an unknown number of
+true genes, due to fragmentation from low coverage and incomplete
+assembly.
+
+We next used Tophat to align mRNAseq reads to the genome and partition
+reads by chromosome; we then assembled the partitioned reads with
+Velvet and Oases using the same range of parameters as the global
+assembly, above.  While the local assemblies were considerably more
+computationally efficient, they lacked several thousand unique regions
+that were present in the global assembly
+(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
+the incomplete nature of the current chicken genome (cite).
+Interestingly, over a hundred regions were present {\em only} in the
+local assemblies, suggesting that the local assemblies might be
+recovering additional exons.  Significant numbers of unique regions
 ($>300$bp) from both global and local assemblies showed homology to
@@
 
-\subsection*{Validation of Gene Models}
+\subsubsection{Cufflinks}
 
+To evaluate an efficiency of our pipeline in detecting splice junctions,
+we compared the number of splice junctions detected by our pipeline to that from , ESTs and
+Ensembl gene models version 64.
+Cufflinks detected 89,289 (86.49\%) splice junctions that are also detected by our pipeline.
+However, 37.61\% (5,565) of junctions not detected by Cufflinks are supported by ESTs.
+This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
+genuine splice junctions.
+
+We also aligned transcripts from Cufflinks and the pipeline to Ensembl transcripts to compare genes and transcripts detected by
+both methods.
+We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
+Our pipeline detects 7,046 genes and Cufflinks detects 8,341 genes of total 17,934 Ensembl genes.
+6,572 of those genes are both detectd by our pipeline and Cufflinks.
+In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
+a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
+Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
+From 6,572 genes, our pipeline detects 57,437 splice junctions, whereas Cufflinks detects 62,101 junctions.
+
+However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
+that are not in Ensembl but are supported by ESTs in those 6,572 genes and found that our pipeline detects 2,260 splice junctions
+and Cufflinks only detects 1,588 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.
+
+To conclude, Cufflinks performs better than our pipeline for detecting both genes and splice junctions in Ensembl gene models.
+However, we found that our pipeline detects more splice junctions that are not in Ensembl but are supported by ESTs.
+=======
 The results show that our pipeline predicts many gene models and

### 2012-10-17 rnaseq_paper.tex

--- 2012-10-15 rnaseq_paper.tex
@@
 Materials).  In the assembly, we used multiple distinct k-mer values
-for Velvet because different isoforms are recovered for different $k$
-\cite{Schulz:2012je}.  We chose k-mers between 21 and 31 for this
-study, and recovered between 340,000 and 370,000 unique sequences from
+for Velvet to sensitively recover as many different isoforms as possible
+\cite{Schulz:2012je}.  We chose k-mers between 21 and 31,
+and recovered between 340,000 and 370,000 unique sequences from
 each data set (see Table~\ref{unique_sequences}, and Materials and
@@
 (Table~\ref{unique_sequences_matched_mouse}); this is probably due to
-the incomplete nature of the current chicken genome (cite).
+the incomplete nature of the current chicken genome assembly, which is
+lacking approximately 5\% of its true gene content (cite).
 Interestingly, over a hundred regions were present {\em only} in the
@@
 recovering additional exons.  Significant numbers of unique regions
-($>300$bp) from both global and local assemblies showed homology to
-the mouse genome, indicating that at least some of these unique sequences
+from both global and local assemblies showed homology to the mouse
+genome, indicating that at least some of these unique sequences
 represented real sequence content.  Figure~\ref{global_vs_local} shows
@@
 
+% @CTB DISCSUSS the use of splice junctions for validity; transition
+% from transcripts to splice junctions.
+
 \subsection*{Oases-M discards splice variants}
 
+The above approaches recovered transcript fragments, but not entire genes
+(figure in suppl?)
 To construct a more comprehensive gene set containing all of the
@@
 of Oases-M for recovering splice variants has not been fully
-evaluated.  We merged transcripts from our global assembly, above,
-with Oases-M using a k-mer size of 27, and then compared them with the
+evaluated.
+% @CTB what does this mean? ``not fully evaluated''?
+We merged transcripts from our global assembly, above,
+with Oases-M using a k-mer size of 27, and compared them with the
 unmerged transcripts.  We then cross-validated using publicly
-available ESTs for an independent data set.  The results show that
+available ESTs, which were not used in our assembly.
+The results show that
 Oases-M and the unmerged assembly share about 95\% of the predicted,
-splice junctions, with about 5\% disjoint.  Of these 5\%,
+splice junctions, with 5\% disjoint.  Of these 5\%,
 approximately 6\% of the Oases-M-specific splice junctions are
 independently supported by ESTs, while 19\% of the unmerged splice
-junctions are supported by ESTs.  This suggests that Oases-M probably
+junctions are supported by ESTs.
+% @CTB change from percent to total number, with percent in parantheses
+This suggests that Oases-M probably
 discards a number of real splice variants, although the unmerged
-assembly is also probably missing some found by Oases-M.
+assembly is also missing some found by Oases-M.
 
@@
 We developed an exon graph approach to construct gene models from
-alignments of transcripts against the genome.  Our approach, called
-Gimme, uses exon graphs to merge transcripts and gene models based on
-overlapping exons (see Materials and Methods).  We used Gimme to
-obtain 20,439 gene models containing 26,035 isoforms from our global
+alignments of transcripts against the genome.  Our approach,
+implemented in a software package called Gimme, merges transcripts and
+gene models based on overlapping exons using an exon-graph approach
+(see Materials and Methods; Figure XX).  We used Gimme to obtain
+20,439 gene models containing 25,859 isoforms from our global
 assembly; 21,963 gene models containing 25,859 from our local
@@
 merged global and local assemblies (Table~\ref{genes_transcripts}).
-The number of genes is slightly decreased after merging transcripts
-from global and local assembly together. 
-This indicates that some transcripts from both methods were merged together.
-Figure~\ref{model_comparisons} shows gene models built by our method with
-other existing gene models.
 
-Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
-especially $3'$ UTR in our datasets.
-These regions are challenging to predict correctly solely from computational
-methods due to low degree of conservation (cite).
-RNA-Seq-derived gene models are useful for studying variations within UTR regions,
-which may be involved in regulation of isoform expression (cite).
-Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
-$5'$UTRs and extended $3'$UTR respectively.
+% (more here - what? @CTB)
 
@@
 
-The results show that our pipeline predicts many gene models and
-isoforms after assembly.  We validated these gene models in several
-different ways.
+Our pipeline predicts many gene models and isoforms after assembly.
+We validated these gene models in several different ways.  XX more!
 
@@
 did not map were either highly erroneous or contained low-complexity
-artifactual sequence that probably came from the reverse transcription
-procedure (Cite).
+artifactual sequence that probably originated from sample processing
+and reverse transcription (cite).
 % @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
@@
 To validate the splice junctions reconstructed by the Gimme pipeline,
-we used Bowtie to map mRNAseq reads to the transcript sequences
-derived from gene models Bowtie\cite{Langmead:2009fv}.  Because the
+we used Bowtie to map mRNAseq reads directly to the transcript sequences
+derived from gene models \cite{Langmead:2009fv}.  Because the
 Velvet/Oases assembly pipeline does not make use of the reference
@@
 have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
+% @CTB how do you get junctions with no spliced reads!?
 More than 99\% of our predicted splice junctions have a coverage
 of 4 or higher in our combined mRNAseq data sets, suggesting that they are
-truly present within the data set.  Note that lower coverage splice junctions
-may still be correct, but may be from low-expressed isoforms or 
-regions of the gene that are low coverage due to sample preparation issues.
+real splice junctions.
+% @CTB I'm slightly bugged by this statement.  Are we right?
 
@@
 are supported by ESTs or mRNAs from Genbank.  This is especially
-surprising since our mRNAseq data is from spleen, and few of the
-publicly available ESTs or mRNAs are from spleen.  Note that this
+surprising since our mRNAseq data is from spleen, and most of the
+publicly available ESTs or mRNAs are from other tissues.  Note that this
 cross-validation suggests that the 19,981 novel splice junctions {\em not}
@@
 Gimme each recover about 10\% of these, with more than 2/3 of these
-recovered by both.  This indicates that both Gimme and Cufflinks
-are equally adept at recovering novel splice junctions.
+recovered by both Cufflinks and Gimme.  This indicates that both Gimme
+and Cufflinks are equally adept at recovering novel splice junctions.
 % @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
@@
 % @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?
+
+% @CTB somewhere we might need to define terms: gene/locus,
+% transcript, and splice junction...
 
@@
 could obtain greater sensitivity by merging both sets of predictions.
-We therefore used Gimme to merge mRNAs, the Cufflinks and assembly gene
-models Table.~\ref{genes_transcripts}.  This resulted in an increased 
-number of total genes and isoforms (e.g. see Fig.~\ref{mrna_cuff_gimme.pdf}).
-The merged gene models recover 51\% and 66\% of splice junctions from ESTs
-and ENSENBL respectively, which is 10-15\% greater than that from corresponding
-unmerged gene models.
+We therefore used Gimme to merge the Cufflinks and assembly gene
+models Table.~\ref{genes_transcripts}.  This resulted in a decreased
+number of total genes, suggesting that some fragmented genes were
+merged together to form more complete gene structures (e.g. see
+Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
+66\% of splice junctions from ESTs and ENSENBL respectively, which is
+10-15\% greater than that from corresponding unmerged gene models.
 % @CTB what about splice junctions etc??
@@
 homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
-of our isoforms from 12,399 genes match mouse proteins.
+of our isoforms from 12,399 distinct genes match mouse proteins at a
+bit score of XXX.
 Most of these matches have a bit-score/length ratio greater than 1,
 which indicates a good agreement between chicken and mouse proteins.
-% @CTB why so few, really?
+% @CTB why so few match, really?
+
+% @CTB for discussion: what's surprising about all of this? so many
+% splice variants, etc?  Note we need to go over pickrell paper carefully.
 
@@
 
-Most transcripts include a large part of untranslated regions (UTRs),
-especially $3'$ UTR in our datasets.
-These regions are challenging to predict correctly solely from computational
-methods due to low degree of conservation [?].
-RNA-Seq-derived gene models are useful for studying variations within UTR
-regions, which may be involved in regulation of isoform expression\cite{}
+Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
+These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
+RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).
 
@@
 junctions due to duplication in genome sequence can lead to a large
-number of dubious isoforms.
+number of dubious isoforms.  This seems to occur in a few genome
+sequences such as chromosome E64\_random.
+Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
+gene models have only a few isoforms.
 
@@
 detection that may result in overestimation of the number of isoforms.
+
+Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
+especially $3'$ UTR in our datasets.
+These regions are challenging to predict correctly solely from computational
+methods due to low degree of conservation (cite).
+RNA-Seq-derived gene models are useful for studying variations within UTR regions,
+which may be involved in regulation of isoform expression (cite).
+Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
+$5'$UTRs and extended $3'$UTR respectively.
+
+\subsubsection{Cufflinks}
 
@@
 
-\subsubsection{isoforms and mouse}
+\subsubsection*{Validating chicken sequences by using mouse homologs}
+
+To validate our predicted isoforms, we extracted coding sequences from
+our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
+successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
+protein sequences with 50 or more amino acids.  We then searched for
+homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
+of our isoforms from 12,399 genes match mouse proteins.
+Most of these matches have a bit-score/length ratio greater than 1,
+which indicates a good agreement between chicken and mouse proteins.
+% @CTB why so few, really?
+
+\section*{An overestimated number of genes and transcripts}
+
+The number of genes from our pipeline might be overestimated due to
+fragmented transcripts.  In our datasets, read coverage is lower at
+the $5'$ end due to the method used to convert mRNA to cDNA in library
+preparation.  A bias of read coverage toward $3'$ end and low
+expression level result in fragmented transcripts as shown in
+Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
+junctions due to duplication in genome sequence can lead to a large
+number of dubious isoforms.
+
+In addition, parameters used in building gene models such as minimum
+exon/intron size determine how transcripts are merged together, which
+can increases or decreases the number of isoforms.  In this study, we
+adjust the parameters to enhance sensitivity of splice junctions
+detection that may result in overestimation of the number of isoforms.
+
+\subsubsection*{isoforms and mouse}
 
@@
 \bf{Unique sequences from global and local assembly}}
+% @CTB do we want to make this base pairs instead of #s? Same for table 2.
 \begin{tabular}{ccccc}
@@
 \hline
-Method& Total & Unique & Supported by ESTs \\ 
+Method& Total & Unique & Unique/supported by ESTs \\ 
 \hline

### 2012-10-17 rnaseq_paper.tex
more revisions

@@
 that were present in the global assembly
+% @CTB why are the global ones not included in the final gene models?
 (Table~\ref{unique_sequences_matched_mouse}); this is probably due to
@@
 The results show that
-Oases-M and the unmerged assembly share about 95\% of the predicted,
+Oases-M and the unmerged assembly share about 95\% of the predicted
 splice junctions, with 5\% disjoint.  Of these 5\%,
@@
 We validated these gene models in several different ways.  XX more!
+
+% @CTB discuss: single ended, short reads => difficult to construct
+% gene models.  How are isoforms assembled by oases, anyway!?
 
@@
 another reference-based approach to building gene models from mRNAseq
-data\cite{Trapnell:2010kd}.  We also compared both methods to the ENSEMBL gene annotations,
+data\cite{Trapnell:2010kd}.  We also compared the results from both methods to the ENSEMBL gene annotations,
 which are produced by a pipeline that incorporates de novo gene
@@
 distinct but equally valid splice junctions, which suggests that we
-could obtain greater sensitivity by merging both sets of predictions.
+could obtain greater sensitivity to exon-exon junctions in our gene models by merging both sets of predictions.
 We therefore used Gimme to merge the Cufflinks and assembly gene
@@
 Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
-66\% of splice junctions from ESTs and ENSENBL respectively, which is
+66\% of splice junctions from ESTs and ENSEMBL respectively, which is
 10-15\% greater than that from corresponding unmerged gene models.
@@
 
+% @CTB Note that these results are, frankly, kind of stunning.  Discuss
+% spleen/immune system diversity.
+
 \subsubsection*{Validating chicken sequences by using mouse homologs}
 
-To validate our predicted isoforms, we extracted coding sequences from
-our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
+To validate our predicted isoforms, we extracted putative coding sequences from
+our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan
 successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
@@
 bit score of XXX.
+% @CTB
 Most of these matches have a bit-score/length ratio greater than 1,
@@
 
-\subsection*{Leftover text}
+\section*{Leftover Results text}
 
@@
 
-\section*{An overestimated number of genes and transcripts}
+\subsection*{An overestimated number of genes and transcripts}
 
@@
 $5'$UTRs and extended $3'$UTR respectively.
-
-\subsubsection{Cufflinks}
 
@@
 that are not in Ensembl but are supported by ESTs.
-
-\subsubsection*{Validating chicken sequences by using mouse homologs}
-
-To validate our predicted isoforms, we extracted coding sequences from
-our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
-successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
-protein sequences with 50 or more amino acids.  We then searched for
-homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
-of our isoforms from 12,399 genes match mouse proteins.
-Most of these matches have a bit-score/length ratio greater than 1,
-which indicates a good agreement between chicken and mouse proteins.
-% @CTB why so few, really?
-
-\section*{An overestimated number of genes and transcripts}
-
-The number of genes from our pipeline might be overestimated due to
-fragmented transcripts.  In our datasets, read coverage is lower at
-the $5'$ end due to the method used to convert mRNA to cDNA in library
-preparation.  A bias of read coverage toward $3'$ end and low
-expression level result in fragmented transcripts as shown in
-Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
-junctions due to duplication in genome sequence can lead to a large
-number of dubious isoforms.
-
-In addition, parameters used in building gene models such as minimum
-exon/intron size determine how transcripts are merged together, which
-can increases or decreases the number of isoforms.  In this study, we
-adjust the parameters to enhance sensitivity of splice junctions
-detection that may result in overestimation of the number of isoforms.
 

### 2012-10-23 rnaseq_paper.tex
added some numbers beside percentages in OasesM topic

@@
 The results show that
-Oases-M and the unmerged assembly share about 95\% of the predicted
+Oases-M and the unmerged assembly share about 104,413 (95\%) of the predicted
 splice junctions, with 5\% disjoint.  Of these 5\%,
-approximately 6\% of the Oases-M-specific splice junctions are
-independently supported by ESTs, while 19\% of the unmerged splice
+approximately 421 (6\%) of the Oases-M-specific splice junctions are
+independently supported by ESTs, while 1,607 (19\%) of the unmerged splice
 junctions are supported by ESTs.

### 2012-10-24 rnaseq_paper.tex
edited the introduction

@@
 \section*{Introduction}
-Until recently, studies of alternative splicing had been limited to a small number of genes and isoforms due to high-cost and low-throughput sequencing of expressed sequence tags (ESTs) and full-length cDNA libraries.
-RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been used successfully in many studies to gain unprecedented insight into a complexity of transcriptomes.
-Novel isoforms have been identified based on reads mapped across known splice junctions or putative splice junctions from annotations.
-It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
+Until recently, studies of alternative splicing had been limited to a small
+number of genes and isoforms due to high-cost and low-throughput sequencing
+of expressed sequence tags (ESTs) and full-length cDNA libraries.  
+RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been
+used successfully in many studies to gain unprecedented insight into a
+complexity of transcriptomes.
+It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative
+splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
 This suggests that even in human a large number of splice variants have not been explored.
 
-Sequences from RNA-Seq are very short (~75--250bp for Illumina reads); therefore, it is not feasible to assess gene
-and isoform expression without mapping reads to a reference annotation or construct a full-length transcript from
-\emph{de novo} assembly.
-Several softwares employing different approaches have been developed to reconstruct transcripts from short reads.
-Tophat\cite{Trapnell:2009dp} maps reads to a reference genome to identify exons and splice sites \emph{ab initio}.
-It first searches for known splie sites in a genome and then tries to map reads across splice juntions to identify exon junctions.
-Exon junctions can be used to build gene models by Cufflinks\cite{Trapnell:2010kd}.
-On the other hand, Trinity\cite{Grabherr:2011jb}, has been developed to assemble short reads \emph{de novo} to construct transcripts.
-Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} assembler that are designed to work with RNA-Seq data.
-These programs are comparable at identifying alternative splicing with a slightly different sensitivity and specificity.
-However, Oases with Oases-M has been shown to be superior than other \emph{de novo} assembler at discovering isoforms in human and mouse\cite{Schulz:2012je}.
+Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
+it is not feasible to assess expression isoforms and their structures without mapping reads
+to a reference annotation or reconstructing a full-length transcript from short reads.
 
-In this study, we present a pipeline that uses Velvet and Oases assembler to discover alternative splicing in RNA-Seq data from chickens.
-We use a new technique called a local assembly that enhances sensitivity of alternative splicing detection of Oases.
-The results show that the pipeline can detect more isoforms than Oases-M and can detect isoforms not found by
-Cufflinks.
+Despite the small size of sequencing reads, several studies have detected
+novel splice junctions based on alignment of reads spanning across putative exons.
+To map reads across splice junctions, reads are split into two parts and mapped
+to the genome independently.
+A splice junction is then identified based on alignments of each half of a read
+that fall between two exons at exon-intron boundaries.
+With this approach, Wang \emph{et al.} have identified a large number of
+splice junctions that are not annotated from human cell lines.
+These novel splice sites include both canonical splie sites and non-canonical splice site.
+Approximately, $46\%-75\%$ of canonical splice sites are supported by ESTs.
+Splice junctions have different read coverage, which suggests that both highly
+and lowly expressed isoforms are not annotated.
+
+Pickrell \emph{et al.} have also identified more than 150,000 novel canonical splice junctions
+in lymphoblastoid cells using a similar approach.
+The study has also shown that the number of unannotated splice junctions varies among cells from
+different human tissues, which suggests tissue-specific expression of isoforms.
+A majority of unannotated isoforms found in both studies are from alternative splicing
+of annotated exons a small fraction of isoforms contain exons from intergenic or intronic region.
+
+Several tools have been developed not only to detect novel splice junctions but also to reconstruct
+full-length isoforms from short reads with or without gene annotations.
+Cufflinks\cite{Trapnell:2010kd} relies on splice junctions detected from Tophat\cite{Trapnell:2009dp},
+a read aligner that can align reads across putative exons, to reconstruct a full-length transcripts.
+Cufflinks was used to identified 12,712 novel isoforms, of which 7,395 ($58\%$) contain novel splice
+junctions in mouse myoblast cell lines.
+The mapping-based approach has been useful in both detecting splice junctions and isoforms.
+One downside of this approach is it relies heavily on a reference genome.
+Therefore, organisms lacking a genome or a high-quality one may not find this approach practical.
+This limitation can be circumvent by employing \emph{de novo} assembly approach.
+
+A number of \emph{de novo} assemblers have been used to reconstruct transcripts from RNA-Seq data in
+many studies.
+Trinity\cite{Grabherr:2011jb}, has successfully been used to reconstruct transcripts both yeast and mouse.
+It has also shown that Trinity detects unique sets of splice junctions when compared to annotated
+junctions and junctions detected by Cufflinks.
+This suggests that \emph{de novo} assembly approach is capabale of increasing sensitivity of detecting
+alternative isoforms.
+Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension
+of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} assembler
+that are tuned to work with RNA-Seq data.
+These programs are comparable at reconstructing known and novel alternative isoforms with a slightly
+different sensitivity and specificity.
+However, Oases with Oases-M has been shown to be superior than other \emph{de novo}
+assembler at discovering isoforms in human and mouse\cite{Schulz:2012je}.
+
+In this study, we present a pipeline that uses Velvet and Oases assembler to
+reconstruct alternative isoforms in RNA-Seq data from chickens.
+We use a new technique called a local assembly that enhances sensitivity of
+alternative splicing detection of Oases.
+The results show that the pipeline can detect more isoforms than Oases-M and
+can detect isoforms not found by Cufflinks.
 

### 2012-10-24 rnaseq_paper.tex
edited some text and added citations

@@
 
-Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
-it is not feasible to assess expression isoforms and their structures without mapping reads
-to a reference annotation or reconstructing a full-length transcript from short reads.
+%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
+%it is nokt feasible to assess expression isoforms and their structures without mapping reads
+%to a reference annotation or reconstructing a full-length transcript from short reads.
 
@@
 that fall between two exons at exon-intron boundaries.
-With this approach, Wang \emph{et al.} have identified a large number of
-splice junctions that are not annotated from human cell lines.
-These novel splice sites include both canonical splie sites and non-canonical splice site.
+With this approach, Wang \emph{et al} have identified a large number of
+splice junctions that are not annotated from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.
+These novel splice sites include both canonical splie and non-canonical splice sites.
 Approximately, $46\%-75\%$ of canonical splice sites are supported by ESTs.
-Splice junctions have different read coverage, which suggests that both highly
-and lowly expressed isoforms are not annotated.
+Novel splice junctions have different read coverage suggesting that both high-
+and low-expressed isoforms are not annotated.
 
-Pickrell \emph{et al.} have also identified more than 150,000 novel canonical splice junctions
+Pickrell \emph{et al} have also identified more than 150,000 novel canonical splice junctions
 in lymphoblastoid cells using a similar approach.
 The study has also shown that the number of unannotated splice junctions varies among cells from
-different human tissues, which suggests tissue-specific expression of isoforms.
+different human tissues, which suggests tissue-specific expression of isoforms\cite{Pickrell:2010gt}.
 A majority of unannotated isoforms found in both studies are from alternative splicing
-of annotated exons a small fraction of isoforms contain exons from intergenic or intronic region.
+of annotated exons and a small fraction of isoforms contain exons from intergenic or
+intronic region.
 
@@
 Cufflinks\cite{Trapnell:2010kd} relies on splice junctions detected from Tophat\cite{Trapnell:2009dp},
-a read aligner that can align reads across putative exons, to reconstruct a full-length transcripts.
+a read aligner that can align reads across putative exon junctions, to reconstruct a full-length transcripts.
 Cufflinks was used to identified 12,712 novel isoforms, of which 7,395 ($58\%$) contain novel splice
 junctions in mouse myoblast cell lines.
-The mapping-based approach has been useful in both detecting splice junctions and isoforms.
+Guttman \emph{et al} used Scripture, a tool employing similar mapping-based approach, to reconstruct
+a full-length transcripts from mouse RNA-Seq data and discovered approximately 490 novel alternative
+isoforms in lincRNA loci in any of the three different cell types\cite{Guttman:2010io}.
+
+The genome-guided method has been useful in both detecting splice junctions and isoforms.
 One downside of this approach is it relies heavily on a reference genome.
 Therefore, organisms lacking a genome or a high-quality one may not find this approach practical.
-This limitation can be circumvent by employing \emph{de novo} assembly approach.
+This limitation can be overcome by \emph{de novo} assembly of short reads.
 
@@
 many studies.
-Trinity\cite{Grabherr:2011jb}, has successfully been used to reconstruct transcripts both yeast and mouse.
-It has also shown that Trinity detects unique sets of splice junctions when compared to annotated
-junctions and junctions detected by Cufflinks.
+Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct transcripts from yeast and mouse datasets.
+It has also been shown that Trinity detects unique sets of splice junctions when compared to annotated
+junctions and junctions detected by Cufflinks and Scripture.
 This suggests that \emph{de novo} assembly approach is capabale of increasing sensitivity of detecting
-alternative isoforms.
+alternative isoforms of genome-guided methods.
 Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension
@@
 can detect isoforms not found by Cufflinks.
+We also showed that transcripts reconstructed from \emph{de novo} assembly and mapping-based approach can be
+merged to build more complete predicted gene models.
 

### 2012-10-24 rnaseq_paper.tex
corrected some numbers, edited the intro

--- 2012-10-23 rnaseq_paper.tex
@@
 \section*{Introduction}
-Until recently, studies of alternative splicing had been limited to a small number of genes and isoforms due to high-cost and low-throughput sequencing of expressed sequence tags (ESTs) and full-length cDNA libraries.
-RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been used successfully in many studies to gain unprecedented insight into a complexity of transcriptomes.
-Novel isoforms have been identified based on reads mapped across known splice junctions or putative splice junctions from annotations.
-It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
+Until recently, studies of alternative splicing had been limited to a small
+number of genes and isoforms due to high-cost and low-throughput sequencing
+of expressed sequence tags (ESTs) and full-length cDNA libraries.  
+RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been
+used successfully in many studies to gain unprecedented insight into a
+complexity of transcriptomes.
+It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative
+splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
 This suggests that even in human a large number of splice variants have not been explored.
 
-Sequences from RNA-Seq are very short (~75--250bp for Illumina reads); therefore, it is not feasible to assess gene
-and isoform expression without mapping reads to a reference annotation or construct a full-length transcript from
-\emph{de novo} assembly.
-Several softwares employing different approaches have been developed to reconstruct transcripts from short reads.
-Tophat\cite{Trapnell:2009dp} maps reads to a reference genome to identify exons and splice sites \emph{ab initio}.
-It first searches for known splie sites in a genome and then tries to map reads across splice juntions to identify exon junctions.
-Exon junctions can be used to build gene models by Cufflinks\cite{Trapnell:2010kd}.
-On the other hand, Trinity\cite{Grabherr:2011jb}, has been developed to assemble short reads \emph{de novo} to construct transcripts.
-Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} assembler that are designed to work with RNA-Seq data.
-These programs are comparable at identifying alternative splicing with a slightly different sensitivity and specificity.
-However, Oases with Oases-M has been shown to be superior than other \emph{de novo} assembler at discovering isoforms in human and mouse\cite{Schulz:2012je}.
+%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
+%it is nokt feasible to assess expression isoforms and their structures without mapping reads
+%to a reference annotation or reconstructing a full-length transcript from short reads.
 
-In this study, we present a pipeline that uses Velvet and Oases assembler to discover alternative splicing in RNA-Seq data from chickens.
-We use a new technique called a local assembly that enhances sensitivity of alternative splicing detection of Oases.
-The results show that the pipeline can detect more isoforms than Oases-M and can detect isoforms not found by
-Cufflinks.
+Despite the small size of sequencing reads, several studies have detected
+novel splice junctions based on alignment of reads spanning across putative exon junctions.
+To map reads across exon junctions, reads are split into two parts and each part is mapped
+to the genome independently.
+A splice junction is then identified based on alignments of each half of a read
+that fall between two exons at exon-intron boundaries.
+With this approach, Wang \emph{et al} have identified a large number of
+splice junctions that are not annotated from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.
+These novel splice sites include both canonical and non-canonical splice sites.
+Approximately, $46\%-75\%$ of canonical splice sites are supported by ESTs.
+Novel splice junctions have different levels of read coverage suggesting that both high-
+and low-expressed isoforms are unannotated.
+
+Using a similar approach, Pickrell \emph{et al} identified more than 150,000 novel canonical splice junctions
+in lymphoblastoid cells.
+The study also shows that the number of unannotated splice junctions varies among cells from
+different human tissues, which suggests tissue-specific expression of isoforms\cite{Pickrell:2010gt}.
+A majority of unannotated isoforms found in both studies are from alternative splicing
+of annotated exons. Only a small fraction of isoforms contain exons from intergenic or
+intronic region.
+
+Several tools have been developed not only to detect novel splice junctions but also to reconstruct
+full-length isoforms from short reads without using gene annotations.
+These tools are especially useful for transcriptome analysis of organims with poor gene annotations.
+Cufflinks\cite{Trapnell:2010kd} relies on splice junctions detected from Tophat\cite{Trapnell:2009dp},
+a read aligner that can align reads across putative exon junctions, to reconstruct a full-length transcript.
+Cufflinks identified 12,712 novel isoforms, of which 7,395 ($58\%$) contain novel splice junctions in
+mouse myoblast cell lines.
+Guttman \emph{et al} used Scripture, a tool employing similar mapping-based approach, to reconstruct
+a full-length transcripts from mouse RNA-Seq data and discovered approximately 490 novel alternative
+isoforms in lincRNA loci, which are expressed in any of the three different cell types\cite{Guttman:2010io}.
+
+Although the mapping-based method has been useful in both detecting splice junctions and isoforms,
+it relies heavily on a reference genome.
+Hence, organisms lacking a genome or a high-quality one may not find this approach practical.
+This limitation can be overcome by \emph{de novo} assembly of short reads.
+
+A number of \emph{de novo} assemblers have been used to reconstruct transcripts from RNA-Seq data in
+many studies.
+Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct transcripts from yeast and mouse datasets.
+It was also shown that Trinity detects a unique set of novel splice junctions not detected by Cufflinks
+or Scripture.
+This suggests that \emph{de novo} assembly approach is capabale of increasing sensitivity of detecting
+alternative isoforms of a mapping-based method.
+Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension
+of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assembler
+that are tuned to work with RNA-Seq data.
+These assemblers are comparable at reconstructing existing and novel alternative isoforms with a slightly
+different sensitivity and specificity.
+However, Oases with Oases-M has been shown to be superior than other \emph{de novo}
+assemblers at discovering isoforms in human and mouse\cite{Schulz:2012je}.
+
+In this study, we present a pipeline that uses Velvet and Oases assembler to
+reconstruct alternative isoforms in RNA-Seq data from chickens.
+We use a new technique called a local assembly that enhances sensitivity of
+alternative isoform detection of Oases.
+The results show that the pipeline can detect more isoforms than Oases-M and
+can detect isoforms not found by Cufflinks.
+We also showed that transcripts reconstructed from \emph{de novo} assembly and mapping-based approach can be
+merged to build more complete predicted gene models.
 
@@
 gene models based on overlapping exons using an exon-graph approach
-(see Materials and Methods; Figure XX).  We used Gimme to obtain
+(see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
 20,439 gene models containing 25,859 isoforms from our global
@@
 
-Of 104,827 splice junctions from the gene models, 248 (0.23\%)
-junctions have no spliced reads and only 484 (0.46\%) junctions
+Of 100,411 splice junctions from the gene models, 95 (0.09\%)
+junctions have no spliced reads and only 448 (0.45\%) junctions
 have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
@@
 
-Of the 104,827 splice junctions in our gene models, 82,908 (80.31\%)
+Of the 100,411 splice junctions in our gene models, 81,838 (81.50\%)
 are supported by ESTs or mRNAs from Genbank.  This is especially
@@
 publicly available ESTs or mRNAs are from other tissues.  Note that this
-cross-validation suggests that the 19,981 novel splice junctions {\em not}
+cross-validation suggests that the 18,573 novel splice junctions {\em not}
 seen in publicly available ESTs and mRNAs are also likely to be real
@@
 distinct but equally valid splice junctions, which suggests that we
-could obtain greater sensitivity to exon-exon junctions in our gene models by merging both sets of predictions.
+could obtain greater sensitivity to exon-exon junctions in our gene
+models by merging both sets of predictions.
 We therefore used Gimme to merge the Cufflinks and assembly gene
@@
 merged together to form more complete gene structures (e.g. see
-Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
-66\% of splice junctions from ESTs and ENSEMBL respectively, which is
+Fig.~\ref{cufflinks_asm}). The merged gene models recover 44.19\% and
+58.57\% of splice junctions from ESTs and ENSEMBL respectively, which is
 10-15\% greater than that from corresponding unmerged gene models.
@@
 protein sequences with 50 or more amino acids.  We then searched for
-homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
+homologous sequences in mouse ENSEMBL, and found that 15,399 (68.47\%)
 of our isoforms from 12,399 distinct genes match mouse proteins at a
-bit score of XXX.
+bit score $\ge1.0$.
 % @CTB
-Most of these matches have a bit-score/length ratio greater than 1,
+These matches have a bit-score/length ratio greater than 1,
 which indicates a good agreement between chicken and mouse proteins.
@@
 Cufflinks & 25,318 & 34,959 \\
+Global + Local + Cufflinks & 21,734 & 32,855 \\
 Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\

### 2014-06-22 paper.tex
renamed rnaseq_paper.tex to paper.tex

--- 2012-10-24 rnaseq_paper.tex
@@
 A gene model can be represented as a splice graph composed of exons as nodes and introns as edges.
-However, transcripts of the same gene vary in size and structure depending on the expression level and a hash length number used in assembly.
+However, transcripts of the same gene vary in size and structure depending on the expression level
+and a hash length number used in assembly.
 Furthermore, incomplete exons and fragmented transcripts complicate the construction of a splice graph.
-In this study, we developed an algorithm that handles incomplete exons and fragmented transcripts and constructs a maximum assembly of gene models.
+In this study, we developed an algorithm that handles incomplete exons and fragmented transcripts
+and constructs a maximum assembly of gene models.
 
 The algorithm first builds an intron graph using introns as nodes.
-Each intron contains exons whose one of their splice sites perfectly match intron boundary.
+Each intron contains exons whose one of their splice sites perfectly match intron boundaries.
 Exons are considered incomplete and eliminated if they locate at the $3'$ or $5'$ end of the transcripts and
@@
 Transcripts were then grouped into the same gene if they have at least one intron or exon in common.
-Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing paths in the splice graph.
+Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing
+paths in the splice graph.
 Gimme is open-source and available at \texttt{https://github.com/ged-lab/gimme}.
@@
 \subsection*{Expressed sequence tags and Genbank mRNA}
-Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome website on May 9th, 2012.
-The database last updated from GENBANK on April 29th, 2012.
-Sequences were alignd to a chicken genome using BLAT.
+Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome website.
+The database was updated from GENBANK on 1 January 2014.
+Sequences were aligned to a chicken genome using BLAT.
 

### 2014-07-05 paper.tex
minor changes

@@
 \section*{Abstract}
-Comparison of spliced reads and reference gene annotations has been successfully used to discover alternative
-splicing in model organisms. However, the method cannot be applied in organisms without high-quality reference genome and
-gene annotations. In this article, we introduce a pipeline, based on \emph{de novo} assembly, that constructs gene models with splice
-variants.
-The pipeline includes a new technique called local assembly that enhances the sensitivity of
-alternative splicing detection.
-We demonstrates that the pipeline detects many novel splice variants in RNA-Seq data from chicken spleens.
-Some of those splice variants; however, only detected by our pipeline and not Cufflinks.
-This indicates that the pipeline can be used to facilitate splice variants detection from RNA-Seq.
+Comparison of spliced reads and reference gene annotations has been
+successfully used to discover alternative splicing in model
+organisms. However, the method cannot be applied in organisms without
+high-quality reference genome and gene annotations. In this article,
+we introduce a pipeline, based on \emph{de novo} assembly, that
+constructs gene models with splice variants.  The pipeline uses a
+technique called local assembly that enhances the sensitivity of
+alternative splicing detection.  We demonstrate that the pipeline
+detects many novel splice variants in RNA-Seq data from chicken
+spleen.  Many of these splice variants; however, are only detected by our
+pipeline and not Cufflinks.  This indicates that the pipeline can be
+used to facilitate splice variant detection from RNA-Seq.
 % Please keep the Author Summary between 150 and 200 words
@@
 % Author Summary not valid for PLoS ONE submissions.
+
+% @CTB: aren't some splice variants detected only by Cufflinks and not by
+% gimme, too?
+% @CTB: for me to check, do Tophat and cufflinks only detect canonical splice?
+% @CTB: ask Olga for references to single-cell splice stuff.
+
 \section*{Author Summary}
@@
 \section*{Introduction}
-Until recently, studies of alternative splicing had been limited to a small
-number of genes and isoforms due to high-cost and low-throughput sequencing
-of expressed sequence tags (ESTs) and full-length cDNA libraries.  
-RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been
-used successfully in many studies to gain unprecedented insight into a
-complexity of transcriptomes.
-It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative
-splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
-This suggests that even in human a large number of splice variants have not been explored.
+
+Until recently, studies of alternative splicing have been limited to a
+small number of genes and isoforms due to high-cost and low-throughput
+sequencing of expressed sequence tags (ESTs) and full-length cDNA
+libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
+has been used successfully in many studies to gain unprecedented
+insight into a complexity of transcriptomes.
+
+It has been estimated that, in human, $92-94$\% of multiexon genes
+undergo alternative splicing and different isoforms are expressed in
+different tissues\cite{Wang:2008ea}.  This suggests that even in human
+a large number of splice variants have not been explored.
 
@@
 
-Despite the small size of sequencing reads, several studies have detected
-novel splice junctions based on alignment of reads spanning across putative exon junctions.
-To map reads across exon junctions, reads are split into two parts and each part is mapped
-to the genome independently.
-A splice junction is then identified based on alignments of each half of a read
-that fall between two exons at exon-intron boundaries.
-With this approach, Wang \emph{et al} have identified a large number of
-splice junctions that are not annotated from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.
-These novel splice sites include both canonical and non-canonical splice sites.
-Approximately, $46\%-75\%$ of canonical splice sites are supported by ESTs.
-Novel splice junctions have different levels of read coverage suggesting that both high-
-and low-expressed isoforms are unannotated.
+Despite the small size of sequencing reads, several studies have
+detected novel splice junctions based on alignment of reads spanning
+across putative exon junctions.  To map reads across exon junctions,
+reads are split into two parts and each part is mapped to the genome
+independently (@cite).  A splice junction is then identified based on
+alignments of each half of a read that fall between two exons at
+exon-intron boundaries.  With this approach, Wang \emph{et al} have
+identified a large number of splice junctions that are not annotated
+from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
+splice sites include both canonical and non-canonical splice sites.
+Approximately, $46\%-75\%$ of canonical splice sites are supported by
+ESTs.  Novel splice junctions have different levels of read coverage
+suggesting that both high- and low-expressed isoforms are unannotated.
 
-Using a similar approach, Pickrell \emph{et al} identified more than 150,000 novel canonical splice junctions
-in lymphoblastoid cells.
-The study also shows that the number of unannotated splice junctions varies among cells from
-different human tissues, which suggests tissue-specific expression of isoforms\cite{Pickrell:2010gt}.
-A majority of unannotated isoforms found in both studies are from alternative splicing
-of annotated exons. Only a small fraction of isoforms contain exons from intergenic or
-intronic region.
+Using a similar approach, Pickrell \emph{et al} identified more than
+150,000 novel canonical splice junctions in lymphoblastoid cells.  The
+study also shows that the number of unannotated splice junctions
+varies among cells from different human tissues, which suggests
+tissue-specific expression of isoforms\cite{Pickrell:2010gt}.  A
+majority of unannotated isoforms found in both studies are from
+alternative splicing of annotated exons. Only a small fraction of
+isoforms contain previously unknown exons.
+% @CTB is this last point important?
 
-Several tools have been developed not only to detect novel splice junctions but also to reconstruct
-full-length isoforms from short reads without using gene annotations.
-These tools are especially useful for transcriptome analysis of organims with poor gene annotations.
-Cufflinks\cite{Trapnell:2010kd} relies on splice junctions detected from Tophat\cite{Trapnell:2009dp},
-a read aligner that can align reads across putative exon junctions, to reconstruct a full-length transcript.
-Cufflinks identified 12,712 novel isoforms, of which 7,395 ($58\%$) contain novel splice junctions in
-mouse myoblast cell lines.
-Guttman \emph{et al} used Scripture, a tool employing similar mapping-based approach, to reconstruct
-a full-length transcripts from mouse RNA-Seq data and discovered approximately 490 novel alternative
-isoforms in lincRNA loci, which are expressed in any of the three different cell types\cite{Guttman:2010io}.
+Several tools have been developed not only to detect novel splice
+junctions but also to reconstruct full-length isoforms from short
+reads without using prior gene annotations.  These tools are especially
+useful for transcriptome analysis of organims with incomplete gene
+annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
+junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
+that can align reads across putative exon junctions, to reconstruct a
+full-length transcript.  Cufflinks identified 12,712 novel isoforms,
+of which 7,395 ($58\%$) contain novel splice junctions in mouse
+myoblast cell lines.  Guttman \emph{et al} used Scripture, a tool
+employing similar mapping-based approach, to reconstruct a full-length
+transcripts from mouse RNA-Seq data and discovered approximately 490
+novel alternative isoforms in lincRNA loci, which are expressed in any
+of the three different cell types\cite{Guttman:2010io}.
 
-Although the mapping-based method has been useful in both detecting splice junctions and isoforms,
-it relies heavily on a reference genome.
-Hence, organisms lacking a genome or a high-quality one may not find this approach practical.
-This limitation can be overcome by \emph{de novo} assembly of short reads.
+Although these mapping-based methods have been useful in detecting
+both splice junctions and isoforms, they rely heavily on a reference
+genome.  Hence, it is not necessarily practical to apply these methods
+to organisms lacking a high-quality reference genome.  This limitation
+can be overcome by \emph{de novo} assembly of short reads.
 
-A number of \emph{de novo} assemblers have been used to reconstruct transcripts from RNA-Seq data in
-many studies.
-Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct transcripts from yeast and mouse datasets.
-It was also shown that Trinity detects a unique set of novel splice junctions not detected by Cufflinks
-or Scripture.
-This suggests that \emph{de novo} assembly approach is capabale of increasing sensitivity of detecting
+A number of \emph{de novo} assemblers have been used to reconstruct
+transcripts from RNA-Seq data in many studies.
+Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct
+transcripts from yeast and mouse datasets.  It was also shown that
+Trinity detects a unique set of novel splice junctions not detected by
+Cufflinks or Scripture.  This suggests that a \emph{de novo} assembly
+approach is capabale of increasing sensitivity of detecting
 alternative isoforms of a mapping-based method.
-Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension
-of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assembler
-that are tuned to work with RNA-Seq data.
-These assemblers are comparable at reconstructing existing and novel alternative isoforms with a slightly
-different sensitivity and specificity.
-However, Oases with Oases-M has been shown to be superior than other \emph{de novo}
-assemblers at discovering isoforms in human and mouse\cite{Schulz:2012je}.
+Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
+extensions of the Abyss\cite{Simpson:2009iv} and
+Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
+tuned to work with RNA-Seq data.  These assemblers are comparable at
+reconstructing existing and novel alternative isoforms with a slightly
+different sensitivity and specificity.  However, Oases with Oases-M
+has been shown to be superior than other \emph{de novo} assemblers at
+discovering isoforms in human and mouse\cite{Schulz:2012je}.
 
-In this study, we present a pipeline that uses Velvet and Oases assembler to
-reconstruct alternative isoforms in RNA-Seq data from chickens.
-We use a new technique called a local assembly that enhances sensitivity of
-alternative isoform detection of Oases.
-The results show that the pipeline can detect more isoforms than Oases-M and
-can detect isoforms not found by Cufflinks.
-We also showed that transcripts reconstructed from \emph{de novo} assembly and mapping-based approach can be
-merged to build more complete predicted gene models.
+In this study, we present a pipeline that uses {\em de novo} assembly
+to reconstruct alternative isoforms in RNA-Seq data from chickens.  We
+apply a technique we call ``local assembly'' that enhances the sensitivity
+of alternative isoform detection by Oases.  The results show that the
+pipeline can detect more isoforms than Oases-M and can detect isoforms
+not found by Cufflinks.  We also showed that transcripts reconstructed
+from \emph{de novo} assembly and mapping-based approach can be merged
+to build more complete gene models.
 
@@
 the incomplete nature of the current chicken genome assembly, which is
-lacking approximately 5\% of its true gene content (cite).
+lacking approximately 5\% of its true gene content (@cite).
 Interestingly, over a hundred regions were present {\em only} in the
@@
 
-% @CTB DISCSUSS the use of splice junctions for validity; transition
+% @CTB DISCUSS the use of splice junctions for validity; transition
 % from transcripts to splice junctions.
@@
 
-\subsection*{Exon graphs can be used to reconstruct putative splice variants}
+\subsection*{Exon graphs can reconstruct putative splice variants}
 
-We developed an exon graph approach to construct gene models from
+We used an exon graph approach to construct gene models from
 alignments of transcripts against the genome.  Our approach,
@@
 amount of information during the merge process
-(Table~\ref{single-end_map}).  More importantly, we also mapped
+(Table~\ref{single-end_map}).
+% @CTB compared to what?
+More importantly, we also mapped
 paired-end reads from technical replicates to the same gene models,
@@
 artifactual sequence that probably originated from sample processing
-and reverse transcription (cite).
+and reverse transcription (@cite).
 % @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
@@
 real splice junctions.
-% @CTB I'm slightly bugged by this statement.  Are we right?
+% @CTB I'm slightly bugged by this statement.  Are we right?!
 
@@
 surprising since our mRNAseq data is from spleen, and most of the
-publicly available ESTs or mRNAs are from other tissues.  Note that this
-cross-validation suggests that the 18,573 novel splice junctions {\em not}
-seen in publicly available ESTs and mRNAs are also likely to be real
-splice junctions from spleen.
+publicly available ESTs or mRNAs are from other tissues.  Note that
+this cross-validation suggests that the 18,573 novel splice junctions
+{\em not} seen in publicly available ESTs and mRNAs are also likely to
+be real splice junctions from spleen.
 % @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?
@@
 another reference-based approach to building gene models from mRNAseq
-data\cite{Trapnell:2010kd}.  We also compared the results from both methods to the ENSEMBL gene annotations,
-which are produced by a pipeline that incorporates de novo gene
-prediction and homology-based approaches as well as expression data.
+data\cite{Trapnell:2010kd}.  We also compared the results from both
+methods to the ENSEMBL gene annotations, which are produced by a
+pipeline that incorporates de novo gene prediction and homology-based
+approaches as well as expression data (@cite).
 
@@
 and Cufflinks are equally adept at recovering novel splice junctions.
+
 % @CTB note this suggests that Cufflinks improved recovery for ENSEMBL

### 2014-07-27 paper.tex
updated data, text and citations

@@
 \\
+
 \bf{3} Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, USA.
@@
 \section*{Abstract}
-Comparison of spliced reads and reference gene annotations has been
-successfully used to discover alternative splicing in model
-organisms. However, the method cannot be applied in organisms without
-high-quality reference genome and gene annotations. In this article,
-we introduce a pipeline, based on \emph{de novo} assembly, that
-constructs gene models with splice variants.  The pipeline uses a
-technique called local assembly that enhances the sensitivity of
-alternative splicing detection.  We demonstrate that the pipeline
-detects many novel splice variants in RNA-Seq data from chicken
-spleen.  Many of these splice variants; however, are only detected by our
-pipeline and not Cufflinks.  This indicates that the pipeline can be
-used to facilitate splice variant detection from RNA-Seq.
+Splice variants play an important role in biological systems especially in the
+immune system and the central nervous system. An expression profile of splice
+variants has been shown to be a better signature of some diseases than an
+overall gene expression profile~\cite{zhang2013isoform}. However, for most
+organisms, available gene models such as Ensembl may not include all splice
+variants expressed in RNA-Seq data. Reference-guided (e.g.
+Cufflinks~\cite{Trapnell:2010kd}) and de novo assembly (e.g.
+Velvet/Oases~\cite{Schulz:2012je}) methods are available for building gene
+models from RNA-Seq reads. However, we found that both types of method recovered
+distinct set of splice variants.  In this article, we introduce a pipeline that
+can combined transcripts from {\em de novo} assembly and other gene models to
+constructs gene models that include more splice variants. We also described a
+method called local assembly that we used to enhance the assembly of splice
+variants to recover splice variants not found by Cufflinks and {\em de novo}
+assembly.  Combination of Cufflinks gene models and transcripts from assembly
+include more splice variants, increasing sensitivity of alternative splicing
+study.
 % Please keep the Author Summary between 150 and 200 words
@@
 % gimme, too?
+% @LP: yes
+
 % @CTB: for me to check, do Tophat and cufflinks only detect canonical splice?
+% @LP: the mechanism used for reads >45 bp detects three canonical splice sites
+% ab initio.
+
 % @CTB: ask Olga for references to single-cell splice stuff.
@@
 
-%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
-%it is nokt feasible to assess expression isoforms and their structures without mapping reads
-%to a reference annotation or reconstructing a full-length transcript from short reads.
+%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina
+%reads), it is not feasible to assess expression isoforms and their structures
+%without mapping reads to a reference annotation or reconstructing a full-length
+%transcript from short reads.
 
@@
 
-Using a similar approach, Pickrell \emph{et al} identified more than
+Using a similar approach, Pickrell {\em et al} identified more than
 150,000 novel canonical splice junctions in lymphoblastoid cells.  The
@@
 varies among cells from different human tissues, which suggests
-tissue-specific expression of isoforms\cite{Pickrell:2010gt}.  A
-majority of unannotated isoforms found in both studies are from
-alternative splicing of annotated exons. Only a small fraction of
-isoforms contain previously unknown exons.
+tissue-specific expression of isoforms\cite{Pickrell:2010gt}. 
+% A majority of unannotated isoforms found in both studies are from alternative
+% splicing of annotated exons. Only a small fraction of isoforms contain
+% previously unknown exons.
 % @CTB is this last point important?
@@
 of which 7,395 ($58\%$) contain novel splice junctions in mouse
-myoblast cell lines.  Guttman \emph{et al} used Scripture, a tool
+myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
 employing similar mapping-based approach, to reconstruct a full-length
@@
 to organisms lacking a high-quality reference genome.  This limitation
-can be overcome by \emph{de novo} assembly of short reads.
+can be overcome by {\em de novo} assembly of short reads.
 
-A number of \emph{de novo} assemblers have been used to reconstruct
+A number of {\em de novo} assemblers have been used to reconstruct
 transcripts from RNA-Seq data in many studies.
@@
 Trinity detects a unique set of novel splice junctions not detected by
-Cufflinks or Scripture.  This suggests that a \emph{de novo} assembly
+Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
 approach is capabale of increasing sensitivity of detecting
@@
 different sensitivity and specificity.  However, Oases with Oases-M
-has been shown to be superior than other \emph{de novo} assemblers at
+has been shown to be superior than other {\em de novo} assemblers at
 discovering isoforms in human and mouse\cite{Schulz:2012je}.
@@
 not found by Cufflinks.  We also showed that transcripts reconstructed
-from \emph{de novo} assembly and mapping-based approach can be merged
+from {\em de novo} assembly and mapping-based approach can be merged
 to build more complete gene models.
@@
 \cite{Schulz:2012je}.  We chose k-mers between 21 and 31,
-and recovered between 340,000 and 370,000 unique sequences from
-each data set (see Table~\ref{unique_sequences}, and Materials and
+and recovered between 90,700 and 104,000 unique sequences from
+each data set (see Table~\ref{total_sequences}, and Materials and
 Methods).  These unique sequences represented an unknown number of
@@
 computationally efficient, they lacked several thousand unique regions
-that were present in the global assembly
+that were present in the global assembly (Table~\ref{unique_sequences},
+
 % @CTB why are the global ones not included in the final gene models?
-(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
-the incomplete nature of the current chicken genome assembly, which is
-lacking approximately 5\% of its true gene content (@cite).
-Interestingly, over a hundred regions were present {\em only} in the
-local assemblies, suggesting that the local assemblies might be
-recovering additional exons.  Significant numbers of unique regions
-from both global and local assemblies showed homology to the mouse
-genome, indicating that at least some of these unique sequences
-represented real sequence content.  Figure~\ref{global_vs_local} shows
-an example of different isoforms detected by the two assembly methods.
+% @LP they are included in the final models if they align to the genome.
+
+and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
+incomplete nature of the current chicken genome assembly, which is lacking
+approximately 5\% of its true gene content (@cite).  Interestingly, over a
+hundred regions were present {\em only} in the local assemblies,
+suggesting that the local assemblies might be recovering additional exons.
+Significant numbers of unique regions from both global and local assemblies
+showed homology to the mouse genome, indicating that at least some of these
+unique sequences represented real sequence content.
+Figure~\ref{global_vs_local} shows an example of different isoforms detected by
+the two assembly methods.
 
 % @CTB how many reads are discarded in the process of mapping?
+
 % @CTB what does local assembly actually do to the DBG??
+% @LP I think the difference is that ambiguous reads, especially spliced reads
+% are used for assembly more than once in local assembly because Tophat reports
+% multiple alignments. So, reads mapped to multiple chromosomes get assembled
+% independently. Assembling all mapped reads at once misses some splice
+% junctions when compared to assembling each chromosome separately.
+
 % @CTB what happens if you ask how many unique regions there are not
@@
 evaluated.
+
 % @CTB what does this mean? ``not fully evaluated''?
-We merged transcripts from our global assembly, above,
-with Oases-M using a k-mer size of 27, and compared them with the
-unmerged transcripts.  We then cross-validated using publicly
-available ESTs, which were not used in our assembly.
-The results show that
-Oases-M and the unmerged assembly share about 104,413 (95\%) of the predicted
-splice junctions, with 5\% disjoint.  Of these 5\%,
-approximately 421 (6\%) of the Oases-M-specific splice junctions are
-independently supported by ESTs, while 1,607 (19\%) of the unmerged splice
-junctions are supported by ESTs.
-% @CTB change from percent to total number, with percent in parantheses
-This suggests that Oases-M probably
-discards a number of real splice variants, although the unmerged
+% @LP Oases paper only evaluate completeness of transcripts by their alignments
+% to the ref sequences. I believe the cutoff they used for a good alignment was
+% 80 percent match. I have to reread it again though, but I'm quite positive.
+
+We merged transcripts from our global assembly, above, with Oases-M using a
+k-mer size of 27, and compared them with the unmerged transcripts.  We then
+cross-validated using publicly available ESTs, which were not used in our
+assembly.  The results show that Oases-M and the unmerged assembly share about
+104,413 (94\%) of the predicted splice junctions, with 6\% disjoint.  Of these
+6\%, approximately 420 (6.1\%) of the Oases-M-specific splice junctions are
+independently supported by ESTs, while 1,608 (19.4\%) of the unmerged splice
+junctions are supported by ESTs (Table~\ref{Oases-M}).  This suggests that Oases-M
+probably discards a number of real splice variants, although the unmerged
 assembly is also missing some found by Oases-M.
 
-% @CTB why would Oases-M create ``good'' splice variants from the unmerged assembly?
+% @CTB why would Oases-M create ``good'' splice variants from the unmerged
+% assembly?
+% @LP My guess is OasesM can extend transcripts from all k-mers and longer
+% transcripts may be aligned better to the genome.
+
 % @CTB does the math for table three work out here?
-% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a general merging tool?
+% @LP yes, percent match is calculated from unique region of each type of
+% assembly.
+
+% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a
+% general merging tool?
+% @LP I think it's kind of a general merging tool.
 
@@
 (see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
-20,439 gene models containing 25,859 isoforms from our global
-assembly; 21,963 gene models containing 25,859 from our local
-assembly; and 20,922 gene models containing 27,732 isoforms from the
+21,492 gene models containing 31,418 isoforms from our global
+assembly; 24,928 gene models containing 29,776 from our local
+assembly; and 22,587 gene models containing 34,800 isoforms from the
 merged global and local assemblies (Table~\ref{genes_transcripts}).
-
-% (more here - what? @CTB)
 
@@
 
-Our pipeline predicts many gene models and isoforms after assembly.
-We validated these gene models in several different ways.  XX more!
+Our pipeline predicts many gene models and isoforms after assembly.  We
+validated these gene models in several different ways.
 
@@
 
-We used bowtie to map single-end reads from the source datasets to the
-transcripts.  More than 80\% of the original reads could be aligned to
-the transcripts, demonstrating that we did not lose a significant
-amount of information during the merge process
-(Table~\ref{single-end_map}).
-% @CTB compared to what?
-More importantly, we also mapped
-paired-end reads from technical replicates to the same gene models,
-and found that more than 85\% of the paired-ends mapped concordantly
-the gene models (Table~\ref{paired-end_map}).  Most of the reads that
-did not map were either highly erroneous or contained low-complexity
-artifactual sequence that probably originated from sample processing
-and reverse transcription (@cite).
-% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
-% paper?  How about a citation that suggests that there's a lot of crap
-% in mRNAseq?
-Thus the merged gene models produced by Gimme represent the significant
-majority of the assemblable data.
+We used Bowtie to map single-end reads from the source datasets to the
+transcripts.  More than $75-80$\% of the original reads could be aligned to the
+transcripts, demonstrating that we did not lose a significant amount of
+information during the merge process compared to the number of reads mapped to
+Ensembl gene models (Figure~\ref{mapped-reads}).
+
+More importantly, we also mapped paired-end reads from technical replicates to
+the same gene models, and found that more than 74\% of the paired-ends mapped
+concordantly the gene models.  Most of the reads that did not map were either
+highly erroneous or contained low-complexity artifactual sequence that probably
+originated from sample processing and reverse transcription (@cite).
+
+% @CTB do we have something we can cite for typical mapping?  Maybe Trinity paper?
+% How about a citation that suggests that there's a lot of crap in mRNAseq?
+% @LP I'll search for it.
+
+Thus the merged gene models produced by Gimme represent the significant majority
+of the assemblable data.
 
@@
 
-Of 100,411 splice junctions from the gene models, 95 (0.09\%)
-junctions have no spliced reads and only 448 (0.45\%) junctions
-have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
+Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
+spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
+(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in random chromosomes,
+which may have a great number of genome missassemblies. The number of junctions
+outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
+
 % @CTB how do you get junctions with no spliced reads!?
-More than 99\% of our predicted splice junctions have a coverage
+% @LP our models have many 2,400 more junctions in random chromosomes than
+% cufflinks and 710 of no reads junctions are from random chromosomes. It is
+% likely that those junctions are from missassemblies.
+% @LP Another reason is it has something to do with variation in read
+% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
+% stringency because it has to allow gap alignment and it uses -v mode to map
+% spliced reads. On the other hand, the default mode for bowtie is -n mode,
+% which is the one I used.
+% @LP I used a parameter that set max alignments to 100, only one alignment will
+% be reported if there are more than 100 alignments. Maybe I should let
+% bowtie report all alignments instead.
+
+More than 95\% of our predicted splice junctions have a coverage
 of 4 or higher in our combined mRNAseq data sets, suggesting that they are
 real splice junctions.
+
 % @CTB I'm slightly bugged by this statement.  Are we right?!
+% @LP the reason I use 4 as a cutoff because there are four samples and if at
+% least one read from each sample map to the junction, it could be real.
+% Apparently, it is not always the case, but that is my simple logic.
 
@@
 
-Of the 100,411 splice junctions in our gene models, 81,838 (81.50\%)
-are supported by ESTs or mRNAs from Genbank.  This is especially
-surprising since our mRNAseq data is from spleen, and most of the
-publicly available ESTs or mRNAs are from other tissues.  Note that
-this cross-validation suggests that the 18,573 novel splice junctions
-{\em not} seen in publicly available ESTs and mRNAs are also likely to
-be real splice junctions from spleen.
+Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
+supported by ESTs or mRNAs from Genbank~\ref{chick_est_venn}.  This is
+especially surprising since our mRNAseq data is from spleen, and most of the
+publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
+are found by either Cufflinks or our gene models suggesting a variation in
+sensitivity between two methods. Note that this cross-validation suggests that
+the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
+mRNAs are also likely to be real splice junctions from spleen.
+
 % @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?
+% @LP supported by ESTs or mRNAs. I unioned splice junctions from ESTs and
+% mRNAs.
 
@@
 
-Cufflinks finds 109,641 splice junctions, and Gimme finds 100,411
-splice junctions.  88,289 of them are in common.  (Venn diagram.)
-
-Both Cufflinks and Gimme find approximately 40-50\% of the genes and
-80-90\% of the splice junctions present in the ENSEMBL gene models
-for chicken.  Cufflinks performs about 10\% better in both cases
-than Gimme, demonstrating that Cufflinks has higher sensitivity
-in recovering ENSEMBL gene models.
-
-The ENSEMBL pipeline does not, however, include a large number of
-splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
-Gimme each recover about 10\% of these, with more than 2/3 of these
-recovered by both Cufflinks and Gimme.  This indicates that both Gimme
-and Cufflinks are equally adept at recovering novel splice junctions.
+Cufflinks finds 92,077 splice junctions, and Gimme finds 105,461 splice
+junctions.  80,964 of them are in common (Figure~\ref{chick_venn}). Both
+Cufflinks and Gimme find approximately 40-50\% of the genes and 50-53\% of the
+splice junctions present in the ENSEMBL gene models for chicken.  The ENSEMBL
+pipeline does not, however, include a large number of splice junctions from ESTs
+(97,740) or mRNAs (13,987).  Cufflinks and Gimme each recover about 18\% of
+these, with more than 2/3 of these recovered by both Cufflinks and Gimme.  This
+indicates that both Gimme and Cufflinks are equally adept at recovering novel
+splice junctions.
 
@@
 % gene models is false...
+% @LP it does recover some that not recovered by assembly.
 % @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?
+% @LP a bit higher than our gene models, around 99\%, 179 junctions with no
+% reads and 1010 junctions <4.
 
@@
 
-When we apply Gimme and Cufflinks to a publicly available mouse
-mRNAseq data set, Gimme and Cufflinks recover approximately the same
-number of splice junctions already known from ENSEMBL.  However, Gimme
-recovers a substantial number of additional splice variants beyond
-Cufflinks and ENSEMBL both.
+When we apply Gimme and Cufflinks to a publicly available mouse mRNAseq data
+set, Gimme and Cufflinks recover approximately the same number of splice
+junctions already known from ENSEMBL (Figure ~\ref{mus_venn}).  However, Gimme
+recovers a substantial number of additional splice variants beyond Cufflinks and
+ENSEMBL both.
 
 % @CTB is the mouse data set from the trinity paper?
+% @LP yes
+
 % @CTB are most of these splice junctions > 3 coverage?
+% @LP I'll figure that out later on.
 
@@
 
-As shown above, Cufflinks and the assembly method detect a number of
-distinct but equally valid splice junctions, which suggests that we
-could obtain greater sensitivity to exon-exon junctions in our gene
-models by merging both sets of predictions.
-We therefore used Gimme to merge the Cufflinks and assembly gene
-models Table.~\ref{genes_transcripts}.  This resulted in a decreased
-number of total genes, suggesting that some fragmented genes were
+As shown above, Cufflinks and the Gimme method detect a number of distinct but
+equally valid splice junctions, which suggests that we could obtain greater
+sensitivity to exon-exon junctions in our gene models by merging both sets of
+predictions.  We therefore used Gimme to incorporate the Cufflinks models to
+global and local assembly Table.~\ref{genes_transcripts}.  This resulted in a
+decreased number of total genes, suggesting that some fragmented genes were
 merged together to form more complete gene structures (e.g. see
-Fig.~\ref{cufflinks_asm}). The merged gene models recover 44.19\% and
-58.57\% of splice junctions from ESTs and ENSEMBL respectively, which is
-10-15\% greater than that from corresponding unmerged gene models.
-% @CTB what about splice junctions etc??
+Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
+splice junctions from ESTs and ENSEMBL respectively
+(Figure~\ref{combined_venn}), which is about 10\% greater than that from
+corresponding unmerged gene models.
 
@@
 To validate our predicted isoforms, we extracted putative coding sequences from
-our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan
-successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
-protein sequences with 50 or more amino acids.  We then searched for
-homologous sequences in mouse ENSEMBL, and found that 15,399 (68.47\%)
-of our isoforms from 12,399 distinct genes match mouse proteins at a
-bit score $\ge1.0$.
-% @CTB
-These matches have a bit-score/length ratio greater than 1,
-which indicates a good agreement between chicken and mouse proteins.
-% @CTB why so few match, really?
+our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
+translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
+50 or more amino acids.  We then searched for homologous sequences in mouse
+ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
+genes match mouse proteins at a bit score $\ge1.0$.
+These matches have a bit-score/length ratio greater than 1, which indicates a
+good agreement between chicken and mouse proteins.
 
@@
 
-\section*{Leftover Results text}
-
-The program can merge all transcripts from multiple-kmers and predict
-a structure of full-length isoforms and genes.
-Total of 20,439 genes were obtained from global assembly and 21,963 genes
-from local assembly.
-However, the number of genes obtained from global + local assembly is only 20,922 genes.
-The number of isoforms slightly increased after combining
-transcripts from global and local assembly together (Table~\ref{genes_transcripts}).
-
-Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
-These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
-RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).
-
-\subsection*{An overestimated number of genes and transcripts}
-
-The number of genes from our pipeline might be overestimated due to
-fragmented transcripts.  In our datasets, read coverage is lower at
-the $5'$ end due to the method used to convert mRNA to cDNA in library
-preparation.  A bias of read coverage toward $3'$ end and low
-expression level result in fragmented transcripts as shown in
-Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
-junctions due to duplication in genome sequence can lead to a large
-number of dubious isoforms.  This seems to occur in a few genome
-sequences such as chromosome E64\_random.
-Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
-gene models have only a few isoforms.
-
-In addition, parameters used in building gene models such as minimum
-exon/intron size determine how transcripts are merged together, which
-can increases or decreases the number of isoforms.  In this study, we
-adjust the parameters to enhance sensitivity of splice junctions
-detection that may result in overestimation of the number of isoforms.
-
-Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
-especially $3'$ UTR in our datasets.
-These regions are challenging to predict correctly solely from computational
-methods due to low degree of conservation (cite).
-RNA-Seq-derived gene models are useful for studying variations within UTR regions,
-which may be involved in regulation of isoform expression (cite).
-Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
-$5'$UTRs and extended $3'$UTR respectively.
-
-\subsubsection*{Cufflinks}
-
-To evaluate an efficiency of our pipeline in detecting splice junctions,
-we compared the number of splice junctions detected by our pipeline to
-that from , ESTs and Ensembl gene models version 64.
-Cufflinks detected 88,289 (87.93\%) splice junctions that are also detected
-by our pipeline.
-However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported
-by ESTs or mRNAs.
-This suggests that the assembly detects some splice junctions that are
-not detected by Cufflinks and they are genuine splice junctions.
-
-We also aligned transcripts from Cufflinks and the pipeline to Ensembl
-transcripts to compare genes and transcripts detected by both methods.
-We selected only transcripts that match with more than 90\% identity and
-cover more than 80\% of an Ensembl transcript.
-Our pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of
-total 17,934 Ensembl genes.
-6,798 of those genes are both detectd by our pipeline and Cufflinks.
-In addition, we wanted to compare the number of isoforms detected by our
-pipeline and Cufflinks within the same gene; however, a transcript can
-match to multiple isoforms in the same gene, especially when it is not complete.
-Therefore, we compare the number of splice junctions detected by both
-methods within the same gene instead.
-From 6,798 genes, our pipeline detects 56,332 splice junctions, whereas
-Cufflinks detects 62,150 junctions.
-
-However, because Ensembl gene models do not include all isoforms or
-splice junctions from ESTs, we investigated splice junctions that are
-not in Ensembl but are supported by ESTs in those 6,798 genes and
-found that our pipeline detects 1,997 splice junctions and Cufflinks
-only detects 1,860 splice junctions.
-Figure ~\ref{alt_splice_site} shows an example of splice junctions not
-included in Ensembl gene models, which are detected by our pipeline.
-
-To conclude, Cufflinks performs better than our pipeline for detecting
-both genes and splice junctions in Ensembl gene models.
-However, we found that our pipeline detects more splice junctions
-that are not in Ensembl but are supported by ESTs.
-
-\subsubsection*{isoforms and mouse}
-
-The results illustrate that isoforms constructed
-by our pipeline are translatable and may be functional.  However,
-genes or isoforms with no match to mouse proteins are not necessarily
-artifacts.  They can be genes or isoforms that are only found in
-non-mammal vertebrates or novel genes.
+% \section*{Leftover Results text}
+% 
+% The program can merge all transcripts from multiple-kmers and predict a
+% structure of full-length isoforms and genes.  Total of 20,439 genes were
+% obtained from global assembly and 21,963 genes from local assembly.  However,
+% the number of genes obtained from global + local assembly is only 20,922 genes.
+% The number of isoforms slightly increased after combining transcripts from
+% global and local assembly together (Table~\ref{genes_transcripts}). 
+% 
+% Most transcripts include a large part of untranslated regions (UTRs), especially
+% $3'$ UTR in our datasets.  These regions are challenging to predict correctly
+% solely from computational methods due to low degree of conservation [?].
+% RNA-Seq-derived gene models are useful for studying variations within UTR
+% regions, which may be involved in regulation of isoform expression\cite{}
+% (Fig.~\ref{long_utr}).
+% 
+% \subsection*{An overestimated number of genes and transcripts}
+% 
+% The number of genes from our pipeline might be overestimated due to
+% fragmented transcripts.  In our datasets, read coverage is lower at
+% the $5'$ end due to the method used to convert mRNA to cDNA in library
+% preparation.  A bias of read coverage toward $3'$ end and low
+% expression level result in fragmented transcripts as shown in
+% Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
+% junctions due to duplication in genome sequence can lead to a large
+% number of dubious isoforms.  This seems to occur in a few genome
+% sequences such as chromosome E64\_random.
+% Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
+% gene models have only a few isoforms.
+% 
+% In addition, parameters used in building gene models such as minimum
+% exon/intron size determine how transcripts are merged together, which
+% can increases or decreases the number of isoforms.  In this study, we
+% adjust the parameters to enhance sensitivity of splice junctions
+% detection that may result in overestimation of the number of isoforms.
+% 
+% Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
+% especially $3'$ UTR in our datasets.
+% These regions are challenging to predict correctly solely from computational
+% methods due to low degree of conservation (cite).
+% RNA-Seq-derived gene models are useful for studying variations within UTR regions,
+% which may be involved in regulation of isoform expression (cite).
+% Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
+% $5'$UTRs and extended $3'$UTR respectively.
+% 
+% \subsubsection*{Cufflinks}
+% 
+% To evaluate an efficiency of our pipeline in detecting splice junctions,
+% we compared the number of splice junctions detected by our pipeline to
+% that from , ESTs and Ensembl gene models version 64.
+% Cufflinks detected 88,289 (87.93\%) splice junctions that are also detected
+% by our pipeline.
+% However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported
+% by ESTs or mRNAs.
+% This suggests that the assembly detects some splice junctions that are
+% not detected by Cufflinks and they are genuine splice junctions.
+% 
+% We also aligned transcripts from Cufflinks and the pipeline to Ensembl
+% transcripts to compare genes and transcripts detected by both methods.
+% We selected only transcripts that match with more than 90\% identity and
+% cover more than 80\% of an Ensembl transcript.
+% Our pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of
+% total 17,934 Ensembl genes.
+% 6,798 of those genes are both detectd by our pipeline and Cufflinks.
+% In addition, we wanted to compare the number of isoforms detected by our
+% pipeline and Cufflinks within the same gene; however, a transcript can
+% match to multiple isoforms in the same gene, especially when it is not complete.
+% Therefore, we compare the number of splice junctions detected by both
+% methods within the same gene instead.
+% From 6,798 genes, our pipeline detects 56,332 splice junctions, whereas
+% Cufflinks detects 62,150 junctions.
+% 
+% However, because Ensembl gene models do not include all isoforms or
+% splice junctions from ESTs, we investigated splice junctions that are
+% not in Ensembl but are supported by ESTs in those 6,798 genes and
+% found that our pipeline detects 1,997 splice junctions and Cufflinks
+% only detects 1,860 splice junctions.
+% Figure ~\ref{alt_splice_site} shows an example of splice junctions not
+% included in Ensembl gene models, which are detected by our pipeline.
+% 
+% To conclude, Cufflinks performs better than our pipeline for detecting
+% both genes and splice junctions in Ensembl gene models.
+% However, we found that our pipeline detects more splice junctions
+% that are not in Ensembl but are supported by ESTs.
+% 
+% \subsubsection*{isoforms and mouse}
+% 
+% The results illustrate that isoforms constructed
+% by our pipeline are translatable and may be functional.  However,
+% genes or isoforms with no match to mouse proteins are not necessarily
+% artifacts.  They can be genes or isoforms that are only found in
+% non-mammal vertebrates or novel genes.
 
@@
 
-Choice of k-mers greatly affects sensitivity of splice variants detection in RNA-Seq assembly using Velvet/Oases.
-Short k-mers increase sensitivity, but also introduce errors from misassembly.
-However, for a study focusing on alternative splicing, it is desirable to detect as many splice variant as possible.
-We have presented the local assembly technique that enhances the ability to detect splice variants of Velvet/Oases from the same
-k-mer.
-The mechanism behind this technique is to be investigated.
-Importantly, understanding of the mechanisms may lead to a filtering technique that can be applied to organisms that lack a reference genome.
+For organisms with a reference genome and Ensembl annotation, building gene
+models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
+reference guide seem to be a preferable method for gene and isoform expression
+analysis.  However, we have shown that Ensembl models in chicken are missing a
+substantial number of splice variants found in ESTs and mRNAs and Cufflinks does
+not recover all splice variants.  Previous study by Schulz \emph{et
+al.}\cite{Schulz:2012je} demonstrates that Cufflinks outperformed \emph{de novo}
+assembly in detecting genes and isoforms in Ensembl annotations in mouse.
+However, we found that Cufflinks almost exclusively finds splice junctions in
+mouse Ensembl gene models when our gene models include splice junctions not in
+Ensembl gene models but supported by ESTs (Figure~\ref{mus_venn}).
 
-We have also developed a pipeline that assembles reconstructed transcripts from multiple k-mers to build putative gene models.
-Overall, a majority of splice variants detected by our pipeline are also detected by Cufflinks.
-Both our pipeline and Cufflinks also detect some splice variants that are not overlapped.
-Combination of both methods results in more complete gene models,
-which suggests that combination of \emph{de novo} assembly and reference-based assembly is preferred to study alternative splicing in chicken.
+To build mouse Cufflinks gene models, we used Tophat default parameters claimed
+to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
+detected splice junctions not in Ensembl.  This suggests that the parameters may
+have been set based on mammalian Ensembl models, which could limit the
+efficiency of finding novel splice variants in other non-mammal organisms.  In
+contrast, {\em de novo} assembly does not rely on species-specific parameters.
+It is, therefore, recommended that a combination of Cufflinks models and
+transcripts from de novo assembly should be used to build gene models that
+include more splice variants.
 
-Interestingly, Cufflinks detects more genes and isoforms included in Ensembl annotations than our pipeline.
-However, our pipeline detects many more splice junctions not included in Ensembl but supported by ESTs.
-As observed in previous study by Schulz \emph{et al.}\cite{Schulz:2012je}, Cufflinks also outperforms \emph{de novo} assembly in
-detecting genes and isoforms in Ensembl annotations in mouse.
-We speculate that Cufflinks possesses a mechanism that is tailored to detect Ensembl gene models.
+We have also shown that the local assembly could greatly increase sensitivity
+of splice variant detection. However, how the local assembly affect the assembly
+is not clearly understood.  We speculated that read with multiple alignments may
+play a significant role in enhancing the assembly of splice variants.
+Understanding of the mechanisms may lead to a technique that can be applied to
+find splice variants in organisms that lack a reference genome.
 
-In this study, gene models are built from chicken genome version 2.1 (galGal3),
-which contains a considerable number of sequence duplications and
-missassemblies that supposed to be eliminated in the latest version of genome assembly (galGal4).
-Duplications and misassemblies lead to false splice junctions, which in turn produce a large number of splice variants as
-observed in some chromosomes.
-Conversely, this suggests that one might be able to use the pipeline to evaluate the quality of reference genome.
+Both Cufflinks and our pipeline relies heavily on a reference genome, therefore,
+the quality of the genome will greatly affect the quality of the gene models.
+In this study gene models were built from chicken genome version 2.1 (galGal3),
+which contains a considerable number of sequence duplications and missassemblies
+that supposed to be eliminated in the latest version of genome assembly
+(galGal4).  Duplications and misassemblies lead to false splice junctions, which
+in turn produce a large number of splice variants as observed in some
+chromosomes.
 
@@
 \subsection*{Quality trimming of reads}
-Both single- and paired-end reads in this study were trimmed using Condetri version 2.1 with default parameters.
-In addition, the first 10 bases of each reads were trimmed off due to an inconsistency of base-calling as shown in supplementary figure S?.
+Both single- and paired-end reads in this study were trimmed using Condetri
+version 2.1 with default parameters.  In addition, the first 10 bases of each
+reads were trimmed off due to an inconsistency of base-calling.
+
+% (as shown in supplementary figure S?.)
 
@@
 Mouse RNA-Seq dataset (SRX062280) is downloaded from Short Read Archives (SRA).
-Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from spleen of chicken line 6 and 7.
+Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from spleen of
+chicken line 6 and 7.
 
@@
 
-Single and paired-end reads were mapped to chicken genome by Tophat2\cite{Trapnell:2009dp} release 2.0.0 using
-default parameters without annotations.
-All reads were mapped to cDNA sequences derived from gene models by Bowtie2\cite{Langmead:2009fv} with
-default parameters \texttt{(n=2, l=28, e=70, k=1)}.
-Reads from mouse dataset were mapped to mouse genome (mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
+Single and paired-end reads were mapped to chicken genome by
+Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without
+annotations.  All reads were mapped to cDNA sequences derived from gene models
+by Bowtie2\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28,
+e=70, k=1)}.  Reads from mouse dataset were mapped to mouse genome (mm9)
+downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
 
@@
 
-Reads from each dataset were first assembled separately in global assembly without using a reference genome.
-In contrast, reads from each dataset were first mapped to the chicken genome using Tophat2.
-Then only reads mapped to the genome were assembled by chromosomes in local assembly (Fig.~\ref{local_assembly}).
-Global and local assembly was performed using Velvet version 1.2.03\cite{Zerbino:2008vu}
-with default parameters except for hash length (k-mer).
-A range of k-mer length from 21-31 was used to assemble reads in both global and local assembly.
-Lastly, transcripts from both methods were assembled by Oases version 0.2.06\cite{Schulz:2012je}.
+Reads from each dataset were first assembled separately in global assembly
+without using a reference genome.  In contrast, reads from each dataset were
+first mapped to the chicken genome using Tophat2.  Then only reads mapped to the
+genome were assembled by chromosomes in local assembly
+(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
+Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
+hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
+reads from chicken data and k-mer length 27 was used to assemble reads from
+mouse data.  Lastly, transcripts from both methods were assembled by Oases
+version 0.2.06\cite{Schulz:2012je}.
+A poly-A tail, short transcripts and transcripts with low complexity are removed
+by seqclean~\cite{seqclean} with default parameters.  Redundant transcripts are
+removed by cd-hit-est from CD-HIT suite\cite{Li:2006hr}.  A substantial number
+of transcripts are removed at this step, which facilitates gene models
+construction process.
 
-A poly-A tail, short transcripts and transcripts with low complexity are removed by
-seqclean\cite{seqclean} with default parameters.
-Redundant transcripts are removed by cd-hit-est from CD-HIT suite\cite{Li:2006hr}.
-A large number of transcripts are removed at this step, which facilitates gene models construction process.
-
-We obtained 334,475 transcripts, of which 295,588 transcripts (88.37\%) mapped to chicken genome.
-Only transcripts mapped to chicken genome are used to build gene models.
+We obtained 339,199 transcripts, of which 315,998 transcripts (93.2\%) mapped
+to chicken genome.  Only transcripts mapped to chicken genome are used to build
+gene models.
 
@@
 
-Figure~\ref{overall_pipeline} depicts an overall gene model construction pipeline.
-Transcripts of all datasets from local and global assembly were mapped to the chicken genome using
-BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna -noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.
-Alignments and gaps from BLAT outputs are considered exons and introns respectively.
-Optionally, data from other sources (ESTs, RefGenes, etc.) can be incorporated with transcripts from assembly to improve gene models.
-All transcripts are then assembled using Gimme, a program that assembles transcripts based on their alignments to the reference genome.
-An algorithm for assembling transcripts is described below.
-A maximum set of transcripts obtained from Gimme are then reduced to only a minimum set of transcripts that contain
-all splice junctions and untranslated regions (UTRs).
-After that, transcripts that are highly similar ($>99\%$) are clustered and removed by CD-HIT version 4.5.6\cite{Li:2006hr}.
-Parameters for clustering are \texttt{-n 10, -r 0 and -c 0.99}.
-Only a representative of each cluster is kept in gene models.
+Figure~\ref{overall_pipeline} depicts an overall gene model construction
+pipeline.  Transcripts of all datasets from local and global assembly were
+mapped to the chicken genome using BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna
+-noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.  Alignments and gaps
+from BLAT outputs are considered exons and introns respectively.  Optionally,
+data from other sources (ESTs, RefGenes, etc.) can be incorporated with
+transcripts from assembly to improve gene models.  All transcripts are then
+assembled using Gimme, a program that assembles transcripts based on their
+alignments to the reference genome.  An algorithm for assembling transcripts is
+described below.  A maximum set of transcripts obtained from Gimme are then
+reduced to only a minimum set of transcripts that contain all splice junctions
+and untranslated regions (UTRs).  After that, transcripts that are highly
+similar ($>99\%$) are clustered and removed by CD-HIT version
+4.5.6\cite{Li:2006hr}.  Parameters for clustering are \texttt{-n 10, -r 0 and -c
+0.99}.  Only a representative of each cluster is kept in gene models.
 
@@
 
-A gene model can be represented as a splice graph composed of exons as nodes and introns as edges.
-However, transcripts of the same gene vary in size and structure depending on the expression level
-and a hash length number used in assembly.
-Furthermore, incomplete exons and fragmented transcripts complicate the construction of a splice graph.
-In this study, we developed an algorithm that handles incomplete exons and fragmented transcripts
-and constructs a maximum assembly of gene models.
+A gene model can be represented as a splice graph composed of exons as nodes and
+introns as edges.  However, transcripts of the same gene vary in size and
+structure depending on the expression level and a hash length number used in
+assembly.  Furthermore, incomplete exons and fragmented transcripts complicate
+the construction of a splice graph.  In this study, we developed an algorithm
+that handles incomplete exons and fragmented transcripts and constructs a
+maximum assembly of gene models.
 
-The algorithm first builds an intron graph using introns as nodes.
-Each intron contains exons whose one of their splice sites perfectly match intron boundaries.
-Exons are considered incomplete and eliminated if they locate at the $3'$ or $5'$ end of the transcripts and
-they are not the largest exons (exon 3a and 3b in Fig.~\ref{algorithm}).
-Transcripts were then grouped into the same gene if they have at least one intron or exon in common.
-Then, a splice graph composed of exons is created and structures of isoforms are derived from traversing
-paths in the splice graph.
-Gimme is open-source and available at \texttt{https://github.com/ged-lab/gimme}.
+The algorithm first builds an intron graph using introns as nodes.  Each intron
+contains exons whose one of their splice sites perfectly match intron
+boundaries.  Exons are considered incomplete and eliminated if they locate at
+the $3'$ or $5'$ end of the transcripts and they are not the largest exons (exon
+3a and 3b in Fig.~\ref{algorithm}).  Transcripts were then grouped into the same
+gene if they have at least one intron or exon in common.  Then, a splice graph
+composed of exons is created and structures of isoforms are derived from
+traversing paths in the splice graph.  Gimme is open-source and available at
+\texttt{https://github.com/ged-lab/gimme}.
 
@@
 
-We employed ESTScan version 2.1 to translate protein sequences from our gene models.
-The matrix used for building Hidden Markov model was built from chicken reference cDNA sequences using tools from ESTScan.
+We employed ESTScan version 3.0.3 to translate protein sequences from our gene
+models.  The matrix used for building Hidden Markov model was built from chicken
+reference cDNA sequences using tools from ESTScan. Only protein sequences longer
+than 50 bp are included in the analysis.
 
@@
 
-To identify unique sequences from two datasets, a set of 20-mers is created for both datasets using
-khmer\cite{khmer}.
-Then, 20-mers from a query dataset are compared with 20-mers from the target dataset.
-The sequence is considered unique if more than 90\% of 20-mers in the query is unique.
-Any unique region shorter than 100 bp is ignored.
+To identify unique sequences from two datasets, a set of 20-mers is created for
+both datasets using khmer\cite{khmer}.  Then, 20-mers from a query dataset are
+compared with 20-mers from the target dataset.  The sequence is considered
+unique if more than 90\% of 20-mers in the query is unique.  Any unique region
+shorter than 300 bp is ignored.
 
@@
 
-Protein sequences translated from each isoform using ESTScan were searched against mouse reference proteins by BLAST 2.2.25+\cite{Tatusova:1999tz}.
-A bit score to a length ratio was calculated for each hit that had an e-value $\le 10^{-20}$.
-Only the highest value of all isoforms from each gene was shown in the gene plot; whereas, values of all isoforms were shown in the isoform plot.
+Protein sequences translated from each isoform using ESTScan were searched
+against mouse reference proteins by BLAST 2.2.25+\cite{Tatusova:1999tz}.  A bit
+score to a length ratio was calculated for each hit that had an e-value $\le$
+10\textsuperscript{-20}.  Only the highest value of all isoforms from each gene
+was shown in the gene plot; whereas, values of all isoforms were shown in the
+isoform plot.
 
@@
 
-Reads from each dataset were mapped to transcripts from the gene models using Bowtie version 2.0-beta 5 with default parameters.
-Reads mapped across exon junctions from all datasets were counted using Samtools\cite{Li:2009vz} and
-Pysam\cite{pysam}
+Reads from each dataset were mapped to transcripts from the gene models using
+Bowtie version 1.0.0. The parameter is set for Bowtie to report up to 100
+alignments per read.  Reads mapped across exon junctions from all datasets were
+counted using Samtools\cite{Li:2009vz} and Pysam\cite{pysam}.
 
 \subsection*{Sequence assembly using Cufflinks}
-Reads are mapped to a genome sequence using Tophat2.
-Gene models are built from each datasets by Cufflinks 2.0.0\cite{Trapnell:2010kd}.
-All gene models are then merged together using Cuffmerge.
+Reads are mapped to a genome sequence using Tophat.  Gene models are built from
+each datasets by Cufflinks 2.0.0\cite{Trapnell:2010kd}.  All gene models are
+then merged together using Cuffmerge.
 
 \subsection*{Expressed sequence tags and Genbank mRNA}
-Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome website.
-The database was updated from GENBANK on 1 January 2014.
-Sequences were aligned to a chicken genome using BLAT.
+Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome
+website.  The database was updated from GENBANK on 1 January 2014.  Sequences
+were aligned to a chicken genome using BLAT.
 
@@
 \begin{table}[!ht]
+
 \caption{
-\bf{Unique sequences from global and local assembly}}
-% @CTB do we want to make this base pairs instead of #s? Same for table 2.
+\bf{Total unique sequences from global and local assembly (k=$21-31$)}}
+\begin{tabular}{ccc}
+\hline
+Dataset & Total Sequence & \\
+        & Global & Local \\
+\hline
+Line 6 uninfected & 90,705 & 68,845\\
+Line 6 infected & 104,785 & 70,191 \\
+Line 7 uninfected & 90,125 & 63,302\\
+Line 7 infected &  92,192 & 67,097 \\
+\hline
+%table information
+\end{tabular}
+%\begin{flushleft}Table Caption
+%\end{flushleft}
+\label{total_sequences}
+
+\caption{
+\bf{Unique sequences between global and local assembly}}
 \begin{tabular}{ccccc}
 \hline
-Dataset & \multicolumn{2}{c}{Total Sequence} & \multicolumn{2}{c}{Unique Sequence}\\
+Dataset & Total size (bp) & & Unique Sequence (bp) & \\
         & Global & Local & Global & Local\\
 \hline
-Line 6 uninfected & 338,353 & 335,180 & 10,068 (2.97\%) & 1,258 (0.37\%)\\
-Line 6 infected & 353,485 & 319,920 & 11,639 (3.29\%)& 2,270 (0.70\%)\\
-Line 7 uninfected & 335,876 & 302,443 & 10,649 (3.17\%) & 1,570 (0.51\%)\\
-Line 7 infected & 371,404 & 327,491 & 11,859 (3.19?\%)& 1,199 (0.36\%)\\
+Line 6 uninfected & 77,454,439 & 36,662,830 & 3,686,835 (4.8\%) & 307,975 (0.8\%)\\
+Line 6 infected & 86,622,623 & 37,877,766 & 4,157,541 (4.8\%)& 400,702 (1.0\%)\\
+Line 7 uninfected & 76,566,717 & 33,571,348 & 4,180,202 (5.4\%) & 365,850 (1.1\%)\\
+Line 7 infected & 74,957,624 & 33,824,849 & 4,242,922 (5.7\%)& 326,169 (9.6\%)\\
 \hline
@@
 \hline
-Line 6 uninfected & 2,132 & 104 & 1,322 (62.01\%) & 39 (37.50\%)\\
-Line 6 infected & 2,499 & 104 & 1,514 (60.58\%)& 40 (38.46\%)\\
-Line 7 uninfected & 2,633 & 136 & 1,560 (59.25\%) & 52 (38.24\%)\\
-Line 7 infected & 2,409 & 152 & 1,390 (57.70\%)& 50 (32.89\%)\\
+Line 6 uninfected & 1285929 & 96830 & 260321 (20.0\%) & 9413 (9.7\%)\\
+Line 6 infected & 1631356 & 59813 & 312849 (19.2\%)& 5132 (8.6\%)\\
+Line 7 uninfected & 1800634 & 104229 & 349346 (19.4\%) & 9883 (9.5\%)\\
+Line 7 infected & 1611354 & 125640 & 296915 (18.4\%)& 9381 (7.5\%)\\
 \hline
@@
 \hline
-Oases-M assembly & 111,237 & 6,860 & 421 (6.1\%) \\
-Unmerged assembly & 112,708 & 8,295 & 1,607 (19.4\%) \\
+Oases-M assembly & 111,273 & 6,860 & 420 (6.1\%) \\
+Global assembly & 112,708 & 8,295 & 1,608 (19.4\%) \\
 \hline
@@
 \hline
-Global & 20,439 & 26,035 \\
-Local & 21,963 & 25,859 \\
-Global + Local & 20,922 & 27,732 \\
-Cufflinks & 25,318 & 34,959 \\
-Global + Local + Cufflinks & 21,734 & 32,855 \\
-Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\
 Ensembl & 17,934 & 23,392 \\
+Global & 21,492 & 31,418 \\
+Local & 24,928 & 29,776 \\
+Global + Local & 22,587 & 34,800 \\
+Cufflinks & 31,073 & 38,307 \\
+Global + Local + Cufflinks & 25044 & 45793 \\
+Global + Local + Cufflinks (w/Ensembl) & 24915 & 60976 \\
 \hline
@@
 
-\caption{
-\bf{Single-end reads mapped to gene models (maximum)}}
-\begin{tabular}{cccccc}
-\hline
-Dataset & Mapped & Unmapped \\
-\hline
-Line 6 uninfected & 20,169,993 (85.54\%) & 3,409,559 (14.46\%) \\
-Line 6 infected & 18,790,831 (80.13\%) & 4,658,150 (19.87\%) \\
-Line 7 uninfected & 19,844,293 (82.94\%) & 4,081,516 (17.06\%) \\
-Line 7 infected & 21,772,102 (85.84\%) & 3,590,465 (14.16\%) \\
-\hline
-%table information
-\end{tabular}
-%\begin{flushleft}Table Caption
-%\end{flushleft}
-\label{single-end_map}
-
-\caption{
-\bf{Paired-end reads mapped to gene models (maximum)}}
-\begin{tabular}{cccccc}
-\hline
-Dataset & Mapped & Unmapped \\
-\hline
-Line 6 uninfected & 28,696,112 (85.36\%) & 4,967,765 (14.76\%) \\
-Line 6 infected & 20,514,438 (85.93\%) & 3,394,055 (14.20\%) \\
-Line 7 uninfected & 28,159,776 (85.65\%) & 4,761,556 (14.46\%) \\
-Line 7 infected & 29,564,592 (85.88\%) & 4,913,243 (14.25\%) \\
-\hline
-%table information
-\end{tabular}
-%\begin{flushleft}Table Caption
-%\end{flushleft}
-\label{paired-end_map}
 \end{table}
@@
 \caption{
-{\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
-Transcripts are aligned to a chicken genome by BLAT\@. Gimme then constructs gene models based on alignments of transcripts.
+{\bf Gene model construction pipeline.} Transcripts are obtained from two
+assembly methods -- global and local assembly.  Transcripts are aligned to a
+chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
+of transcripts.
 }
@@
 {\bf Local Assembly Pipeline.}
-Reads are first mapped to a chicken genome.
-Then only mapped reads are assembled by Velvet and Oases.
+Reads are first mapped to a chicken genome.  Then only mapped reads are
+assembled by Velvet and Oases.
 }
@@
 \caption{
-{\bf Intron and exon graphs.}
-Each intron connects to exons whose splice junctions match it boundary.
-Some exons are excluded from the final gene model if they are incomplete (exon 3a,b).
-Introns sharing at least one exon are grouped together.
-Then an exon graph is made using exons as nodes.
+{\bf Intron and exon graphs.} Each intron connects to exons whose splice
+junctions match it boundary.  Some exons are excluded from the final gene model
+if they are incomplete (exon 3a,b).  Introns sharing at least one exon are
+grouped together.  Then an exon graph is made using exons as nodes.
 }
 \label{algorithm}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{kmers-variance.pdf}
-\end{center}
-\caption{
-{\bf Different isoforms are detected by different k-mer lengths.}
-K-mer=21 detects a skipped exon which is not detected by k-mer=31.
-The skipped exon is also annotated in Ensembl gene models.
-}
-\label{kmer-variance}
 \end{figure}
@@
 
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{kmers-variance.pdf}
+% \end{center}
+% \caption{
+% {\bf Different isoforms are detected by different k-mer lengths.}
+% K-mer=21 detects a skipped exon which is not detected by k-mer=31.
+% The skipped exon is also annotated in Ensembl gene models.
+% }
+% \label{kmer-variance}
+% \end{figure}
+
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{fragmented_transcripts.pdf}
+% \end{center}
+% \caption{
+% {\bf Example of fragmented transcripts near $5'$ end of a long transcript.}
+% }
+% \label{fragmented_transcripts}
+% \end{figure}
+
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{model_comparisons.pdf}
+% \end{center}
+% \caption{
+% {\bf Comparison of gene models from the \emph{de novo} assembly pipeline with
+% reference and Ensembl gene models on UCSC genome browser.}
+% }
+% \label{model_comparisons}
+% \end{figure}
+
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=6in]{alter_5utr.pdf}
+% \end{center}
+% \caption{
+% {\bf Examples of alternative $5'$ UTRs in RNA-Seq gene models.}
+% }
+% \label{alter_5utr}
+% \end{figure}
+% 
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{long_utr.pdf}
+% \end{center}
+% \caption{
+% {\bf Examples of an extended $3'$ UTR in RNA-Seq gene models.}
+% }
+% \label{long_utr}
+% \end{figure}
+
 \begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=5in]{fragmented_transcripts.pdf}
+\includegraphics[width=5in]{mapped-reads.pdf}
 \end{center}
 \caption{
-{\bf Example of fragmented transcripts near $5'$ end of a long transcript.}
+{\bf Cumulative counts of splice junctions with spliced reads from both single-
+and paired-end data from Cufflinks and Gimme models.}
 }
-\label{fragmented_transcripts}
+\label{mapped-reads}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=5in]{model_comparisons.pdf}
+\includegraphics[width=5in]{cdf_splice.pdf}
 \end{center}
 \caption{
-{\bf Comparison of gene models from the \emph{de novo} assembly pipeline with reference and Ensembl gene models on UCSC genome browser.}
+{\bf Cumulative counts of splice junctions with spliced reads from both single-
+and paired-end data.}
 }
-\label{model_comparisons}
+\label{cdf_splice}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=6in]{alter_5utr.pdf}
+\includegraphics[width=5in]{chick_est_venn.pdf}
 \end{center}
 \caption{
-{\bf Examples of alternative $5'$ UTRs in RNA-Seq gene models.}
+    {\bf Splice junctions supported by ESTs or mRNAs.}
 }
-\label{alter_5utr}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{long_utr.pdf}
-\end{center}
-\caption{
-{\bf Examples of an extended $3'$ UTR in RNA-Seq gene models.}
-}
-\label{long_utr}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{bitscore.pdf}
-\end{center}
-\caption{
-{\bf Histogram of bit score/length ratio of isoforms and genes that match mouse proteins.}
-}
-\label{bitscore}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{cdf_single_splice.pdf}
-\end{center}
-\caption{
-{\bf Cumulative counts of splice junctions with spliced reads up to 1000 reads.}
-}
-\label{cdf_single_splice}
+\label{chick_est_venn}
 \end{figure}
@@
 \caption{
-{\bf Splice sites in chicken Ensembl gene models detected by Cufflinks and the \emph{de novo} assembly pipeline.}
-Cufflinks detects many annotated isoforms that are not detected by the pipeline.
-The figure also shows that both methods detect a large number of unannotated splice junctions, which suggests
-that those junctions may be genuine.
-}
+{\bf Splice sites in chicken Ensembl gene models detected by Cufflinks and the
+{\em de novo} assembly pipeline.} Cufflinks detects many annotated isoforms
+that are not detected by the pipeline.  The figure also shows that both methods
+detect a large number of unannotated splice junctions, which suggests that those
+junctions may be genuine.  }
 \label{chick_venn}
@@
 \begin{center}
-\includegraphics[width=5in]{chick_est_venn.pdf}
+\includegraphics[width=5in]{mouse_venn.pdf}
 \end{center}
 \caption{
-{\bf Splice sites in Chicken ESTs detected by Cufflinks and the \emph{de novo} assembly pipeline.}
-In contrast to Ensembl gene models, the pipeline and Cufflinks detects the similar number of splice junctions
-from ESTs and mRNAs. This indicates that the pipeline is as efficient as Cufflinks at detecting
-non-annotated splice junctions.
-}
-\label{chick_est_venn}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{mus_venn.pdf}
-\end{center}
-\caption{
-{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de novo} assembly pipeline.}
-In contrast to chicken datasets, Cufflinks and the pipeline detects the similar amount of non-overlapped annotated junctions.
-This may due to a higher quality of gene annotations in mouse.
-}
+{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de novo}
+assembly pipeline.} In contrast to chicken datasets, Cufflinks and the pipeline
+detects the similar amount of non-overlapped annotated junctions.  This may due
+to a higher quality of gene annotations in mouse.  }
 \label{mus_venn}
@@
 \begin{center}
-\includegraphics[width=5in]{alt_splice_site.pdf}
+\includegraphics[width=5in]{cuff_gimme_junctions_venn.pdf}
 \end{center}
 \caption{
-{\bf Unannotated alternative splice site.}
-The pipeline detects alternative splices site not annotated in Ensembl and Cufflinks but are supported by ESTs.
+    {\bf Splice junctions found in merged models.} Merged models (B) find more
+    splice junctions in ENSEMBL and ESTs + mRNAs than Cufflinks models only (A).
 }
-\label{alt_splice_site}
+\label{combined_venn}
 \end{figure}
+
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{chick_est_mrna_venn.pdf}
+% \end{center}
+% \caption{
+% {\bf Splice sites in Chicken ESTs and mRNAs not included in Ensembl detected by
+% Cufflinks and the \emph{de novo} assembly pipeline.}
+% Both Cufflinks and our pipeline find splice junctions not included in Ensembl
+% but supported by EST and mRNA sequences.
+% }
+% \label{chick_est_mrna_venn}
+% \end{figure}
 
@@
 \begin{center}
-\includegraphics[width=5in]{mrna_cuff_gimme.pdf}
+\includegraphics[width=5in]{mouse-homolog-boxplot.pdf}
 \end{center}
 \caption{
-{\bf mRNAs + Cufflinks + Assembly gene models.}
-Gimme can combine transcripts from different sources to build gene models.
-In this figure, the final gene model includes several isoforms not annotated in the reference gene model.
+{\bf Box plots of bit score/length ratio of isoforms and genes that match mouse
+proteins.}
 }
-\label{mrna_cuff_gimme}
+\label{bitscore}
 \end{figure}
 
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{alt_splice_site.pdf}
+% \end{center}
+% \caption{
+% {\bf Unannotated alternative splice site.} The pipeline detects alternative
+% splices site not annotated in Ensembl and Cufflinks but are supported by ESTs.
+% }
+% \label{alt_splice_site}
+% \end{figure}
+% 
+% \begin{figure}[!ht]
+% \begin{center}
+% \includegraphics[width=5in]{mrna_cuff_gimme.pdf}
+% \end{center}
+% \caption{
+% {\bf mRNAs + Cufflinks + Assembly gene models.} Gimme can combine transcripts
+% from different sources to build gene models.  In this figure, the final gene
+% model includes several isoforms not annotated in the reference gene model.  }
+% \label{mrna_cuff_gimme}
+% \end{figure}
+
 \end{document}

### 2014-09-10 paper.tex
revised based on comments from Hans and Jerry

@@
 \section*{Abstract}
-Splice variants play an important role in biological systems especially in the
+Splice variants play an important role in biological systems, especially in the
 immune system and the central nervous system. An expression profile of splice
@@
 models from RNA-Seq reads. However, we found that both types of method recovered
-distinct set of splice variants.  In this article, we introduce a pipeline that
-can combined transcripts from {\em de novo} assembly and other gene models to
+distinct sets of splice variants.  In this article, we introduce a pipeline that
+can combine transcripts from {\em de novo} assembly and other gene models to
 constructs gene models that include more splice variants. We also described a
@@
 detected novel splice junctions based on alignment of reads spanning
-across putative exon junctions.  To map reads across exon junctions,
+putative exon junctions.  To map reads across exon junctions,
 reads are split into two parts and each part is mapped to the genome
 independently (@cite).  A splice junction is then identified based on
-alignments of each half of a read that fall between two exons at
+alignments of each half of a read that falls between two exons at
 exon-intron boundaries.  With this approach, Wang \emph{et al} have
@@
 
-Using a similar approach, Pickrell {\em et al} identified more than
-150,000 novel canonical splice junctions in lymphoblastoid cells.  The
-study also shows that the number of unannotated splice junctions
-varies among cells from different human tissues, which suggests
-tissue-specific expression of isoforms\cite{Pickrell:2010gt}. 
+Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
+identified more than 150,000 novel canonical splice junctions in lymphoblastoid
+cells.  The study also shows that the number of unannotated splice junctions
+varies among cells from different human tissues, which suggests tissue-specific
+expression of isoforms\cite{Pickrell:2010gt}.
 % A majority of unannotated isoforms found in both studies are from alternative
@@
 that can align reads across putative exon junctions, to reconstruct a
-full-length transcript.  Cufflinks identified 12,712 novel isoforms,
+full-length transcript.  Cufflinks~\cite{} identified 12,712 novel isoforms,
 of which 7,395 ($58\%$) contain novel splice junctions in mouse
 myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
-employing similar mapping-based approach, to reconstruct a full-length
+employing a similar mapping-based approach, to reconstruct a full-length
 transcripts from mouse RNA-Seq data and discovered approximately 490
@@
 transcripts from RNA-Seq data in many studies.
-Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct
+Trinity\cite{Grabherr:2011jb} was successfully used to reconstruct
 transcripts from yeast and mouse datasets.  It was also shown that
@@
 Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
-approach is capabale of increasing sensitivity of detecting
-alternative isoforms of a mapping-based method.
+approach is capable of increasing sensitivity of detecting
+alternative isoforms over a mapping-based method.
 Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
@@
 different sensitivity and specificity.  However, Oases with Oases-M
-has been shown to be superior than other {\em de novo} assemblers at
+has been shown to be superior to other {\em de novo} assemblers at
 discovering isoforms in human and mouse\cite{Schulz:2012je}.
@@
 not found by Cufflinks.  We also showed that transcripts reconstructed
-from {\em de novo} assembly and mapping-based approach can be merged
+from {\em de novo} assembly and mapping-based approaches can be merged
 to build more complete gene models.
@@
 highly erroneous or contained low-complexity artifactual sequence that probably
-originated from sample processing and reverse transcription (@cite).
+originated from sample processing and reverse transcription (@cite).  Thus the
+merged gene models produced by Gimme represent the significant majority of the
+assemblable data.
 
@@
 % @LP I'll search for it.
-
-Thus the merged gene models produced by Gimme represent the significant majority
-of the assemblable data.
 
@@
 spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
-(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in random chromosomes,
+(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
 which may have a great number of genome missassemblies. The number of junctions
 outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
-
 % @CTB how do you get junctions with no spliced reads!?
@@
 % bowtie report all alignments instead.
-
 More than 95\% of our predicted splice junctions have a coverage
@@
 real splice junctions.
-
 % @CTB I'm slightly bugged by this statement.  Are we right?!
@@
 Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
-supported by ESTs or mRNAs from Genbank~\ref{chick_est_venn}.  This is
+supported by ESTs or mRNAs from Genbank (Fig.~\ref{chick_est_venn}).  This is
 especially surprising since our mRNAseq data is from spleen, and most of the
@@
 predictions.  We therefore used Gimme to incorporate the Cufflinks models to
-global and local assembly Table.~\ref{genes_transcripts}.  This resulted in a
+global and local assembly, Table.~\ref{genes_transcripts}.  This resulted in a
 decreased number of total genes, suggesting that some fragmented genes were
@@
 models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
-reference guide seem to be a preferable method for gene and isoform expression
+reference guide seems to be a preferable method for gene and isoform expression
 analysis.  However, we have shown that Ensembl models in chicken are missing a
 substantial number of splice variants found in ESTs and mRNAs and Cufflinks does
-not recover all splice variants.  Previous study by Schulz \emph{et
-al.}\cite{Schulz:2012je} demonstrates that Cufflinks outperformed \emph{de novo}
+not recover all splice variants.  A previous study by Schulz \emph{et
+al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
 assembly in detecting genes and isoforms in Ensembl annotations in mouse.
 However, we found that Cufflinks almost exclusively finds splice junctions in
-mouse Ensembl gene models when our gene models include splice junctions not in
+mouse Ensembl gene models whereas our gene models include splice junctions not in
 Ensembl gene models but supported by ESTs (Figure~\ref{mus_venn}).
@@
 It is, therefore, recommended that a combination of Cufflinks models and
-transcripts from de novo assembly should be used to build gene models that
+transcripts from {\em de novo assembly} should be used to build gene models that
 include more splice variants.
@@
 We have also shown that the local assembly could greatly increase sensitivity
-of splice variant detection. However, how the local assembly affect the assembly
-is not clearly understood.  We speculated that read with multiple alignments may
-play a significant role in enhancing the assembly of splice variants.
-Understanding of the mechanisms may lead to a technique that can be applied to
-find splice variants in organisms that lack a reference genome.
-
-Both Cufflinks and our pipeline relies heavily on a reference genome, therefore,
+of splice variant detection. However, how the local assembly affects the
+assembly process is not clearly understood.  We speculate that reads with
+multiple alignments may play a significant role in enhancing the assembly of
+splice variants.  Understanding of the mechanisms may lead to a technique that
+can be applied to find splice variants in organisms that lack a reference
+genome.
+
+Both Cufflinks and our pipeline rely heavily on a reference genome; therefore,
 the quality of the genome will greatly affect the quality of the gene models.
 In this study gene models were built from chicken genome version 2.1 (galGal3),
-which contains a considerable number of sequence duplications and missassemblies
-that supposed to be eliminated in the latest version of genome assembly
+which contains ~17 Mb of sequence duplications and missassemblies
+that were eliminated in the latest version of genome assembly
 (galGal4).  Duplications and misassemblies lead to false splice junctions, which
@@
 
-\subsection*{Mapping reads to The Reference Genome and Gene Models}
-
-Single and paired-end reads were mapped to chicken genome by
+\subsection*{Mapping reads to The Genome and Gene Models}
+
+Single and paired-end reads were mapped to the chicken genome by
 Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without
@@
 by Bowtie2\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28,
-e=70, k=1)}.  Reads from mouse dataset were mapped to mouse genome (mm9)
-downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
+e=70, k=1)}.  Reads from the mouse dataset were mapped to the mouse genome
+(mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
 
@@
 
-Reads from each dataset were first assembled separately in global assembly
+Reads from each dataset were first assembled separately by global assembly
 without using a reference genome.  In contrast, reads from each dataset were
-first mapped to the chicken genome using Tophat2.  Then only reads mapped to the
-genome were assembled by chromosomes in local assembly
+first mapped to the chicken genome using Tophat2.  Then only reads mapped to
+the genome were assembled by chromosome in the local assembly
 (Fig.~\ref{local_assembly}).  Global and local assembly was performed using
@@
 mouse data.  Lastly, transcripts from both methods were assembled by Oases
-version 0.2.06\cite{Schulz:2012je}.
-A poly-A tail, short transcripts and transcripts with low complexity are removed
-by seqclean~\cite{seqclean} with default parameters.  Redundant transcripts are
-removed by cd-hit-est from CD-HIT suite\cite{Li:2006hr}.  A substantial number
-of transcripts are removed at this step, which facilitates gene models
-construction process.
+version 0.2.06\cite{Schulz:2012je}.  A poly-A tail, short transcripts and
+transcripts with low complexity are removed by seqclean~\cite{seqclean} with
+default parameters.  Redundant transcripts are removed by cd-hit-est from the
+CD-HIT suite\cite{Li:2006hr}.  A substantial number of transcripts are removed
+at this step, which facilitates gene model construction process.
 
@@
 data from other sources (ESTs, RefGenes, etc.) can be incorporated with
-transcripts from assembly to improve gene models.  All transcripts are then
+transcripts from the assembly to improve gene models.  All transcripts are then
 assembled using Gimme, a program that assembles transcripts based on their
@@
 structure depending on the expression level and a hash length number used in
-assembly.  Furthermore, incomplete exons and fragmented transcripts complicate
+the assembly.  Furthermore, incomplete exons and fragmented transcripts complicate
 the construction of a splice graph.  In this study, we developed an algorithm
@@
 The algorithm first builds an intron graph using introns as nodes.  Each intron
-contains exons whose one of their splice sites perfectly match intron
-boundaries.  Exons are considered incomplete and eliminated if they locate at
-the $3'$ or $5'$ end of the transcripts and they are not the largest exons (exon
-3a and 3b in Fig.~\ref{algorithm}).  Transcripts were then grouped into the same
-gene if they have at least one intron or exon in common.  Then, a splice graph
-composed of exons is created and structures of isoforms are derived from
-traversing paths in the splice graph.  Gimme is open-source and available at
+contains exons one of whose splice sites perfectly match intron boundaries.
+Exons are considered incomplete and eliminated if they locate at the $3'$ or
+$5'$ end of the transcripts and they are not the largest exons.  ,
+Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
+they have at least one intron or exon in common.  Then, a splice graph composed
+of exons is created and structures of isoforms are derived from traversing
+paths in the splice graph.  Gimme is open-source and available at
 \texttt{https://github.com/ged-lab/gimme}.
@@
 We employed ESTScan version 3.0.3 to translate protein sequences from our gene
-models.  The matrix used for building Hidden Markov model was built from chicken
+models.  The matrix used for building Hidden Markov models was built from chicken
 reference cDNA sequences using tools from ESTScan. Only protein sequences longer
@@
 Reads are mapped to a genome sequence using Tophat.  Gene models are built from
-each datasets by Cufflinks 2.0.0\cite{Trapnell:2010kd}.  All gene models are
+each dataset by Cufflinks 2.0.0\cite{Trapnell:2010kd}.  All gene models are
 then merged together using Cuffmerge.
@@
 \subsection*{Expressed sequence tags and Genbank mRNA}
-Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome
-website.  The database was updated from GENBANK on 1 January 2014.  Sequences
-were aligned to a chicken genome using BLAT.
+Expressed sequence tags (ESTs) and mRNAs were downloaded from the UCSC genome
+website.  The database was loaded from GENBANK on 1 January 2014.  Sequences
+were aligned to the chicken genome using BLAT.
 
@@
 {\bf Gene model construction pipeline.} Transcripts are obtained from two
-assembly methods -- global and local assembly.  Transcripts are aligned to a
+assembly methods -- global and local assembly.  Transcripts are aligned to the
 chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
-of transcripts.
-}
+of transcripts.  }
 \label{overall_pipeline}

### 2014-09-16 paper.tex
minor update
misc fixes

@@
 variants to recover splice variants not found by Cufflinks and {\em de novo}
-assembly.  Combination of Cufflinks gene models and transcripts from assembly
-include more splice variants, increasing sensitivity of alternative splicing
-study.
+assembly.  Combining Cufflinks gene models and transcripts from assembly
+yields more splice variants, increasing sensitivity of alternative splicing
+studies.
 % Please keep the Author Summary between 150 and 200 words
@@
 Until recently, studies of alternative splicing have been limited to a
-small number of genes and isoforms due to high-cost and low-throughput
-sequencing of expressed sequence tags (ESTs) and full-length cDNA
+small number of genes and isoforms due to the high cost and low throughput
+of sequencing expressed sequence tags (ESTs) and full-length cDNA
 libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
 has been used successfully in many studies to gain unprecedented
-insight into a complexity of transcriptomes.
-
+insight into the complexity of transcriptomes (@cite).
 It has been estimated that, in human, $92-94$\% of multiexon genes
@@
 splice sites include both canonical and non-canonical splice sites.
-Approximately, $46\%-75\%$ of canonical splice sites are supported by
+Approximately $46\%-75\%$ of canonical splice sites are supported by
 ESTs.  Novel splice junctions have different levels of read coverage
 suggesting that both high- and low-expressed isoforms are unannotated.
-
 Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
@@
 
-Several tools have been developed not only to detect novel splice
-junctions but also to reconstruct full-length isoforms from short
-reads without using prior gene annotations.  These tools are especially
-useful for transcriptome analysis of organims with incomplete gene
-annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
-junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
-that can align reads across putative exon junctions, to reconstruct a
-full-length transcript.  Cufflinks~\cite{} identified 12,712 novel isoforms,
-of which 7,395 ($58\%$) contain novel splice junctions in mouse
-myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
-employing a similar mapping-based approach, to reconstruct a full-length
-transcripts from mouse RNA-Seq data and discovered approximately 490
-novel alternative isoforms in lincRNA loci, which are expressed in any
-of the three different cell types\cite{Guttman:2010io}.
-
-Although these mapping-based methods have been useful in detecting
-both splice junctions and isoforms, they rely heavily on a reference
-genome.  Hence, it is not necessarily practical to apply these methods
-to organisms lacking a high-quality reference genome.  This limitation
+Several reference-based tools have been developed not only to detect
+novel splice junctions but also to reconstruct full-length isoforms
+from short reads without using prior gene annotations.  These tools
+are especially useful for transcriptome analysis of organims with
+incomplete gene annotations.  Cufflinks\cite{Trapnell:2010kd} relies
+on splice junctions detected from Tophat\cite{Trapnell:2009dp}, a read
+aligner that can align reads across putative exon junctions, to
+reconstruct a full-length transcript.  In mRNAseq from mouse myoblast
+cell lines, Cufflinks identified 12,712 novel isoforms, of which 7,395
+($58\%$) contain novel splice junctions.  Guttman {\em et al} used
+Scripture, a tool employing a similar mapping-based approach, to
+reconstruct full-length transcripts from mouse RNA-Seq data and
+discovered approximately 490 novel alternative isoforms in lincRNA
+loci, which are expressed in any of the three different cell
+types\cite{Guttman:2010io}.  Although these mapping-based methods have
+been useful in detecting both splice junctions and isoforms, they rely
+heavily on a high-quality reference genome.  Hence, it is not
+necessarily practical to apply these methods to organisms lacking such
+a reference.
+
+The requirement for a high-quality reference genome
 can be overcome by {\em de novo} assembly of short reads.
-
 A number of {\em de novo} assemblers have been used to reconstruct
@@
 
-We next used Tophat to align mRNAseq reads to the genome and partition
+We separately used Tophat to align mRNAseq reads to the genome and partition
 reads by chromosome; we then assembled the partitioned reads with
 Velvet and Oases using the same range of parameters as the global
-assembly, above.  While the local assemblies were considerably more
+assembly, above.  While these ``local'' assemblies were considerably more
 computationally efficient, they lacked several thousand unique regions
@@
 publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
-are found by either Cufflinks or our gene models suggesting a variation in
+are found by either Cufflinks or our gene models, but not both, suggesting a variation in
 sensitivity between two methods. Note that this cross-validation suggests that
@@
 reference guide seems to be a preferable method for gene and isoform expression
-analysis.  However, we have shown that Ensembl models in chicken are missing a
-substantial number of splice variants found in ESTs and mRNAs and Cufflinks does
-not recover all splice variants.  A previous study by Schulz \emph{et
+analysis.  However, we find that Ensembl models in chicken are missing a
+substantial number of splice variants found in ESTs and mRNAs, and Cufflinks does
+not recover a number of them.  A previous study by Schulz \emph{et
 al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
 assembly in detecting genes and isoforms in Ensembl annotations in mouse.
-However, we found that Cufflinks almost exclusively finds splice junctions in
-mouse Ensembl gene models whereas our gene models include splice junctions not in
-Ensembl gene models but supported by ESTs (Figure~\ref{mus_venn}).
+We find that Cufflinks almost exclusively finds splice junctions in
+mouse Ensembl gene models, whereas our gene models include splice junctions that are not in
+Ensembl gene models but that {\em are} well supported by ESTs (Figure~\ref{mus_venn}).
 
@@
 contrast, {\em de novo} assembly does not rely on species-specific parameters.
-It is, therefore, recommended that a combination of Cufflinks models and
-transcripts from {\em de novo assembly} should be used to build gene models that
+We therefore recommend that a combination of Cufflinks models and
+transcripts from {\em de novo assembly} be used to build gene models that
 include more splice variants.
@@
 
-Both Cufflinks and our pipeline rely heavily on a reference genome; therefore,
+Both Cufflinks and our pipeline rely on a reference genome; therefore,
 the quality of the genome will greatly affect the quality of the gene models.
@@
 \subsection*{Data}
-Mouse RNA-Seq dataset (SRX062280) is downloaded from Short Read Archives (SRA).
-Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from spleen of
-chicken line 6 and 7.
+
+The mouse RNA-Seq dataset (SRX062280) was downloaded from the Short Read
+Archive.  Chicken RNA-Seq datasets were obtained from
+sequencing of mRNAs from spleen of chicken line 6 and 7.
 
@@
 Exons are considered incomplete and eliminated if they locate at the $3'$ or
-$5'$ end of the transcripts and they are not the largest exons.  ,
+$5'$ end of the transcripts and they are not the largest exons.
+% @CTB ???
+  ,
 Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
@@
 To identify unique sequences from two datasets, a set of 20-mers is created for
-both datasets using khmer\cite{khmer}.  Then, 20-mers from a query dataset are
+both datasets using khmer\cite{khmer} (@CTB what script? I may have just removed it :).  Then, 20-mers from a query dataset are
 compared with 20-mers from the target dataset.  The sequence is considered

### 2014-09-18 paper.tex
removed reads_error_profiles_legend.pdf
minor revised paper.tex

updated cdf figure

@@
 \hline
-Line 6 uninfected & 1285929 & 96830 & 260321 (20.0\%) & 9413 (9.7\%)\\
-Line 6 infected & 1631356 & 59813 & 312849 (19.2\%)& 5132 (8.6\%)\\
-Line 7 uninfected & 1800634 & 104229 & 349346 (19.4\%) & 9883 (9.5\%)\\
-Line 7 infected & 1611354 & 125640 & 296915 (18.4\%)& 9381 (7.5\%)\\
+Line 6 uninfected & 1,285,929 & 96,830 & 260,321 (20.0\%) & 9,413 (9.7\%)\\
+Line 6 infected & 1,631,356 & 59,813 & 312,849 (19.2\%)& 5,132 (8.6\%)\\
+Line 7 uninfected & 1,800,634 & 104,229 & 349,346 (19.4\%) & 9,883 (9.5\%)\\
+Line 7 infected & 1,611,354 & 125,640 & 296,915 (18.4\%)& 9,381 (7.5\%)\\
 \hline
@@
 Cufflinks & 31,073 & 38,307 \\
-Global + Local + Cufflinks & 25044 & 45793 \\
-Global + Local + Cufflinks (w/Ensembl) & 24915 & 60976 \\
+Global + Local + Cufflinks & 25,044 & 45,793 \\
+Global + Local + Cufflinks (w/Ensembl) & 24,915 & 60,976 \\
 \hline
@@
 \caption{
-{\bf Cumulative counts of splice junctions with spliced reads from both single-
-and paired-end data.}
+{\bf Cumulative counts of splice junctions with spliced reads from single- and
+paired-end data.}
 }
@@
 that are not detected by the pipeline.  The figure also shows that both methods
-detect a large number of unannotated splice junctions, which suggests that those
-junctions may be genuine.  }
+detect a large number of unannotated splice junctions.}
 \label{chick_venn}
@@
 assembly pipeline.} In contrast to chicken datasets, Cufflinks and the pipeline
-detects the similar amount of non-overlapped annotated junctions.  This may due
-to a higher quality of gene annotations in mouse.  }
+detects the similar amount of non-overlapped annotated junctions.}
 \label{mus_venn}

### 2014-09-20 paper.tex
updated discussion, bitscores part and others.

@@
 % general merging tool?
-% @LP I think it's kind of a general merging tool.
+% @LP I think it's kind of a general merging tool based on de Bruijn graph.
 
@@
 % @CTB I'm slightly bugged by this statement.  Are we right?!
-% @LP the reason I use 4 as a cutoff because there are four samples and if at
-% least one read from each sample map to the junction, it could be real.
-% Apparently, it is not always the case, but that is my simple logic.
+% @LP the main reason I use four reads as a cutoff because there are four
+% samples and if at least one read from each sample map to the junction, it
+% will likely be real.  Apparently, it is not always the case, but that is just
+% my simple logic.
 
@@
 splice junctions from ESTs and ENSEMBL respectively
-(Figure~\ref{combined_venn}), which is about 10\% greater than that from
+(Fig.~\ref{combined_venn}), which is about 10\% greater than that from
 corresponding unmerged gene models.
@@
 
+% @LP Maybe I should add bitscores from Cufflinks so that we can compare with
+% those from Gimme. If the results are similar, we can say that extending
+% Cufflinks with assembly does not affect gene structures in a bad way.
 To validate our predicted isoforms, we extracted putative coding sequences from
@@
 ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
-genes match mouse proteins at a bit score $\ge1.0$.
-These matches have a bit-score/length ratio greater than 1, which indicates a
-good agreement between chicken and mouse proteins.
+genes match mouse proteins at a bit-score/length $\ge1.0$
+(Fig.~\ref{bitscore}).  A bit-score/length ratio greater than 1 indicates that
+approximately more than half of the protein sequence matches mouse proteins
+well. The results suggest that merging gene models from Cufflinks and {\em de
+novo} aseembly did not disrupt the structure of protein sequences.
 
@@
 % splice variants, etc?  Note we need to go over pickrell paper carefully.
+% @LP I discuss about it below. Hope it helps.
 
@@
 
-For organisms with a reference genome and Ensembl annotation, building gene
-models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
-reference guide seems to be a preferable method for gene and isoform expression
-analysis.  However, we find that Ensembl models in chicken are missing a
-substantial number of splice variants found in ESTs and mRNAs, and Cufflinks does
-not recover a number of them.  A previous study by Schulz \emph{et
-al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
-assembly in detecting genes and isoforms in Ensembl annotations in mouse.
-We find that Cufflinks almost exclusively finds splice junctions in
-mouse Ensembl gene models, whereas our gene models include splice junctions that are not in
-Ensembl gene models but that {\em are} well supported by ESTs (Figure~\ref{mus_venn}).
-
-To build mouse Cufflinks gene models, we used Tophat default parameters claimed
-to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
-detected splice junctions not in Ensembl.  This suggests that the parameters may
-have been set based on mammalian Ensembl models, which could limit the
-efficiency of finding novel splice variants in other non-mammal organisms.  In
-contrast, {\em de novo} assembly does not rely on species-specific parameters.
-We therefore recommend that a combination of Cufflinks models and
-transcripts from {\em de novo assembly} be used to build gene models that
-include more splice variants.
-
-We have also shown that the local assembly could greatly increase sensitivity
-of splice variant detection. However, how the local assembly affects the
-assembly process is not clearly understood.  We speculate that reads with
-multiple alignments may play a significant role in enhancing the assembly of
-splice variants.  Understanding of the mechanisms may lead to a technique that
-can be applied to find splice variants in organisms that lack a reference
-genome.
-
-Both Cufflinks and our pipeline rely on a reference genome; therefore,
-the quality of the genome will greatly affect the quality of the gene models.
-In this study gene models were built from chicken genome version 2.1 (galGal3),
-which contains ~17 Mb of sequence duplications and missassemblies
-that were eliminated in the latest version of genome assembly
-(galGal4).  Duplications and misassemblies lead to false splice junctions, which
-in turn produce a large number of splice variants as observed in some
-chromosomes.
+\subsection*{{\em De novo} assembly should be used to extend gene models from
+Cufflinks}
+For organisms with a reference genome and ENSEMBL annotation, Cufflinks with
+ENSEMBL gene models as a reference guide seems to be a preferable method for
+building gene models. However, we have shown that ENSEMBL models in chicken are
+missing a substantial number of splice variants found in ESTs and mRNAs that
+are also not recovered by Cufflinks. Moreover, results from mouse data suggest
+that Tophat's default parameters may have been tailored to maximize the finding
+of isoforms in ENSEMBL models (Fig.~\ref{mus_venn}), which could lead to
+restricted detection of novel splice variants.
+% 85,512 (97.2\%) of splice
+% junctions found by Cufflinks are also canonical compared to 92.6\% in mouse
+% ENSEMBL gene models. Assuming that mouse data were not selected for splice
+% variants with canonical splice junctions, the percentage of canonical splice
+% sites found by Cufflinks is rather high. 
+Therefore, gene models built from RNA-Seq reads using Cufflinks + ENSEMBL gene
+models alone may not be suitable for genes and isoforms analyses that include
+finding novel genes and splice variants.  {\em De novo} assembly is not tied to
+any specific gene model parameters and does not only detect canonical splice
+sites, thus we recommend that it is used to extend gene models built from
+Cufflinks + ENSEMBL gene models, especially in organisms with relatively
+incomplete ENSEMBL gene models.  Besides {\em de novo} assembly,
+annotation from other sources can also be incorporated to extend Cufflinks models
+using our pipeline.
+
+% Moreover, efficiency of finding minor and non-canonical splice sites
+% of Cufflinks is also low, which could cause a problem when Cufflinks is used to
+% construct gene models for organisms with more isoforms with minor and
+% non-canonicla splice sites.
+
+\subsection*{Reads with multiple alignments may be key to detection of unique
+splice variants from local assembly}
+We have shown that local assembly greatly increases sensitivity of splice
+variant detection; however, the mechanism is still not clearly understood.  In
+local assembly, reads mapped to each chromosome are assembled separately and
+because Tophat reports multiple alignments for spliced reads, a single spliced
+read from an exon junction present in multiple chromosomes gets assembled
+multiple times. We speculate that this increases the coverage of exon junctions
+found in multiple chromosomes, which help improve the assembly of splice
+variants with low coverage. Further investigation of an inherent mechanism of
+local assembly may lead to a method that increases sensitivity of splice
+variants in organisms that lack a reference genome.
+
+\subsection*{Spurious predicted isoforms are due to poor genome assembly}
+Both Cufflinks and our pipeline rely on a reference genome; therefore, the
+quality of the genome greatly affects the quality of the gene models.  In this
+study, gene models were built from chicken genome version 2.1 (galGal3), which
+contains ~17 Mb of sequence duplications and missassemblies that were
+eliminated in the latest version of genome assembly (galGal4).  Duplications
+and misassemblies lead to spurious splice junctions, which in turn produce
+spurious splice variants. Figure~\ref{bitscore} shows that the median of
+bitscore/length ratios of isoforms is lower than that of genes, suggesting
+poorer alignments of isoforms to mouse protein sequences.
 
@@
 version 2.1 with default parameters.  In addition, the first 10 bases of each
-reads were trimmed off due to an inconsistency of base-calling.
-
-% (as shown in supplementary figure S?.)
+reads were trimmed off due to an inconsistency of base-calling as shown in
+supplementary Figure 1.
 
@@
 The mouse RNA-Seq dataset (SRX062280) was downloaded from the Short Read
-Archive.  Chicken RNA-Seq datasets were obtained from
-sequencing of mRNAs from spleen of chicken line 6 and 7.
+Archive.  Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from
+spleens of control and infected chicken (4 days post infection) lines 6 and 7
+using standard Illumina protocol for non-strand specific single-end reads.
 
@@
 Single and paired-end reads were mapped to the chicken genome by
-Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without
+Tophat\cite{Trapnell:2009dp} release 1.3.1 using default parameters without
 annotations.  All reads were mapped to cDNA sequences derived from gene models
-by Bowtie2\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28,
-e=70, k=1)}.  Reads from the mouse dataset were mapped to the mouse genome
-(mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.
+by Bowtie\cite{Langmead:2009fv} release 1.0.0 with default parameters Reads from the
+mouse dataset were mapped to the mouse genome (mm9) downloaded from Tophat
+website \texttt{http://tophat.cbcb.umd.edu}.
 
@@
 
+% @LP check the version of tophat
 Reads from each dataset were first assembled separately by global assembly
@@
 mouse data.  Lastly, transcripts from both methods were assembled by Oases
-version 0.2.06\cite{Schulz:2012je}.  A poly-A tail, short transcripts and
-transcripts with low complexity are removed by seqclean~\cite{seqclean} with
-default parameters.  Redundant transcripts are removed by cd-hit-est from the
-CD-HIT suite\cite{Li:2006hr}.  A substantial number of transcripts are removed
-at this step, which facilitates gene model construction process.
-
-We obtained 339,199 transcripts, of which 315,998 transcripts (93.2\%) mapped
-to chicken genome.  Only transcripts mapped to chicken genome are used to build
+version 0.2.06\cite{Schulz:2012je}.
+
+A poly-A tail, short transcripts and transcripts with low complexity are
+removed by SeqClean~\cite{seqclean} with default parameters.  Redundant
+transcripts are removed by \texttt{cd-hit-est} from the CD-HIT
+suite\cite{Li:2006hr} release 4.5.4.  A substantial number of transcripts are
+removed at this step, which facilitates gene model construction process.  We
+obtained 339,199 transcripts, of which 315,998 transcripts (93.2\%) mapped to
+chicken genome.  Only transcripts mapped to chicken genome are used to build
 gene models.
@@
 pipeline.  Transcripts of all datasets from local and global assembly were
-mapped to the chicken genome using BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna
--noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.  Alignments and gaps
+mapped to the chicken genome using BLAT\cite{Kent:2002tv}. Alignments and gaps
 from BLAT outputs are considered exons and introns respectively.  Optionally,
@@
 reduced to only a minimum set of transcripts that contain all splice junctions
-and untranslated regions (UTRs).  After that, transcripts that are highly
-similar ($>99\%$) are clustered and removed by CD-HIT version
-4.5.6\cite{Li:2006hr}.  Parameters for clustering are \texttt{-n 10, -r 0 and -c
-0.99}.  Only a representative of each cluster is kept in gene models.
+and untranslated regions (UTRs).
 
@@
 The algorithm first builds an intron graph using introns as nodes.  Each intron
-contains exons one of whose splice sites perfectly match intron boundaries.
-Exons are considered incomplete and eliminated if they locate at the $3'$ or
-$5'$ end of the transcripts and they are not the largest exons.
-% @CTB ???
-  ,
-Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
-they have at least one intron or exon in common.  Then, a splice graph composed
-of exons is created and structures of isoforms are derived from traversing
-paths in the splice graph.  Gimme is open-source and available at
+contains exons whose splice sites perfectly match intron boundaries. At the
+$5\prime$ and the $3\prime$ ends, some exons may be smaller than others due to
+incomplete assembly (supp. Fig.??). Only the largest exons are kept to ensure
+that the transcript has the most complete exons (Fig.~\ref{algorithm}).
+Transcripts were then grouped into the same gene if they have at least one
+intron or exon in common.  Then, a splice graph composed of exons is created
+and structures of isoforms are derived from traversing paths in the splice
+graph.  Gimme is open-source and available at
 \texttt{https://github.com/ged-lab/gimme}.
@@
 To identify unique sequences from two datasets, a set of 20-mers is created for
-both datasets using khmer\cite{khmer} (@CTB what script? I may have just removed it :).  Then, 20-mers from a query dataset are
+both datasets using khmer\cite{} . Then, 20-mers from a query dataset are
 compared with 20-mers from the target dataset.  The sequence is considered
@@
 shorter than 300 bp is ignored.
+% @CTB what script? I may have just removed it.
+% @LP The script is named assembly-diff.py. I have it in my repo. It is a
+% modified version of yours.
+
+% @LP how do I cite khmer? Should I cite QP paper?
 
@@
 were aligned to the chicken genome using BLAT.
+
+\subsection*{Pipeline for the study}
+The pipeline, scripts and instructions for reproducing this study are open
+source and available at \texttt{http://github.com/likit/gimme-paper}.
 

### 2014-09-20 paper.tex
updated some citations and figure description

@@
 alignments per read.  Reads mapped across exon junctions from all datasets were
-counted using Samtools\cite{Li:2009vz} and Pysam\cite{pysam}.
+counted using Samtools\cite{li2009sequence} and Pysam\cite{pysam}.
 
@@
 \caption{
-{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de novo}
-assembly pipeline.} In contrast to chicken datasets, Cufflinks and the pipeline
-detects the similar amount of non-overlapped annotated junctions.}
+{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de
+novo} assembly pipeline.} In contrast to chicken datasets, 85,160 (96.8\%) of
+splice junctions found by Cufflinks are annotated in mouse ENSEMBL models. The
+higher percentage of ENSEMBL splice junctions found by Cufflinks may be a
+result of more complete ENSEMBL gene models. While Gimme can detect the similar
+number of ENSEMBL splice junctions, it detects many more non-ENSEMBL splice
+junctions in mouse.}
 \label{mus_venn}

### 2014-09-20 paper.tex
updated description

@@
 study, gene models were built from chicken genome version 2.1 (galGal3), which
-contains ~17 Mb of sequence duplications and missassemblies that were
+contains $\sim17$ Mb of sequence duplications and missassemblies that were
 eliminated in the latest version of genome assembly (galGal4).  Duplications
 and misassemblies lead to spurious splice junctions, which in turn produce
-spurious splice variants. Figure~\ref{bitscore} shows that the median of
-bitscore/length ratios of isoforms is lower than that of genes, suggesting
-poorer alignments of isoforms to mouse protein sequences.
+spurious splice variants. Spurious junctions are non-canonical and a majority
+of them (17.8\%) are in chrUn\_random (sup.Fig.2). Figure~\ref{bitscore} also
+shows that the median of bitscore/length ratios of isoforms is lower than that
+of genes, suggesting poorer alignments of isoforms to mouse protein sequences.
 
@@
 {\bf Box plots of bit score/length ratio of isoforms and genes that match mouse
-proteins.}
-}
+proteins.} Only the greatest ratio of isoforms from the same gene is plotted for
+the gene plot. Ratios from every isoform are plotted for the isoform plot.}
 \label{bitscore}

### 2014-09-30 paper.tex
updated some figure descriptions

@@
 
-% @CTB why are the global ones not included in the final gene models?
-% @LP they are included in the final models if they align to the genome.
-
 and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
@@
 the two assembly methods.
-
-% @CTB how many reads are discarded in the process of mapping?
 
@@
 
-% @LP Maybe I should add bitscores from Cufflinks so that we can compare with
-% those from Gimme. If the results are similar, we can say that extending
-% Cufflinks with assembly does not affect gene structures in a bad way.
 To validate our predicted isoforms, we extracted putative coding sequences from
 our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
-translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
-50 or more amino acids.  We then searched for homologous sequences in mouse
-ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
-genes match mouse proteins at a bit-score/length $\ge1.0$
-(Fig.~\ref{bitscore}).  A bit-score/length ratio greater than 1 indicates that
-approximately more than half of the protein sequence matches mouse proteins
-well. The results suggest that merging gene models from Cufflinks and {\em de
-novo} aseembly did not disrupt the structure of protein sequences.
-
-% @CTB for discussion: what's surprising about all of this? so many
-% splice variants, etc?  Note we need to go over pickrell paper carefully.
-% @LP I discuss about it below. Hope it helps.
+translated 28,772 of 34,800 (82.6\%) of isoforms from Gimme models to protein
+sequences with 50 or more amino acids.  We then searched for homologous
+sequences in mouse ENSEMBL, and found that 22,460 (77.7\%) and 22,991 (79.9\%)
+of isoforms from Cufflinks and Gimme match mouse proteins at a bit-score/length
+$\ge1.0$(Fig.~\ref{bitscore}).  The number of distinct genes are 16,731 and
+12,945 from Cufflinks and Gimme respectively.  A bit-score/length ratio greater
+than 1 indicates that approximately more than half of the protein sequence
+matches mouse proteins well. The sizes of genes and isoforms from Gimme are
+slightly greater than those from Cufflinks while the bitscore/length ratios are
+slightly lower suggesting that transcripts from assembly are more complete.
+
+% @CTB for discussion: what's surprising about all of this? so many splice
+% variants, etc?  Note we need to go over pickrell paper carefully.  @LP I
+% discuss about it below. Hope it helps.
 
@@
 
-\subsection*{Mapping reads to The Genome and Gene Models}
+\subsection*{Mapping Reads to The Genome and Gene Models}
 
@@
 
-Figure~\ref{overall_pipeline} depicts an overall gene model construction
+Figure~\ref{img:pipeline} depicts an overall gene model construction
 pipeline.  Transcripts of all datasets from local and global assembly were
@@
 \section*{Figure Legends}
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{overall_pipeline.pdf}
-\end{center}
-\caption{
-{\bf Gene model construction pipeline.} Transcripts are obtained from two
-assembly methods -- global and local assembly.  Transcripts are aligned to the
-chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
-of transcripts.  }
-\label{overall_pipeline}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{local_assembly}
-\end{center}
-\caption{
-{\bf Local Assembly Pipeline.}
-Reads are first mapped to a chicken genome.  Then only mapped reads are
-assembled by Velvet and Oases.
-}
-\label{local_assembly}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{algorithm.pdf}
-\end{center}
-\caption{
-{\bf Intron and exon graphs.} Each intron connects to exons whose splice
-junctions match it boundary.  Some exons are excluded from the final gene model
-if they are incomplete (exon 3a,b).  Introns sharing at least one exon are
-grouped together.  Then an exon graph is made using exons as nodes.
-}
-\label{algorithm}
-\end{figure}
 
@@
 
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{pipeline.pdf}
+\end{center}
+\caption{
+{\bf Gene model construction pipeline.}
+Transcripts are obtained from two assembly methods -- global and
+local assembly. Transcripts with low complexity are removed and
+poly(A) tails are trimmed by SeqClean. Transcripts are aligned to
+the chicken genome using BLAT and alignments are then used by
+Gimme to construct gene models. Gene models from Cufflinks and
+other sources can also be incorporated to build the gene models.
+}
+\label{img:pipeline}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{local_assembly.pdf}
+\end{center}
+\caption{
+{\bf Local assembly pipeline.}
+Reads are first mapped to a chicken genome by Tophat.  Then reads
+mapped to each chromosome are assembled by Velvet and Oases
+separately.
+}
+\label{local_assembly}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{algorithm.pdf}
+\end{center}
+\caption{
+{\bf Intron and exon graph.}
+Introns sharing at least one exon are grouped together, then an
+exon graph is constructed with exons from all introns as nodes.
+Exons are excluded from the final gene model if they are smaller
+than other exons sharing the same intron (exon 1a, 3a and 3b).
+Edges of the exon graph (red arrow) are implied from the
+connections of introns (blue arrow).  A transcript is constructed
+from concatenation of all unique exons in the graph.
+}
+\label{algorithm}
+\end{figure}
 \end{document}

### 2014-10-01 paper.tex
updated results and figure descriptions
edited bitscore/length ratio figure

@@
 spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
-(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
+(Figure~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
 which may have a great number of genome missassemblies. The number of junctions
@@
 Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
-supported by ESTs or mRNAs from Genbank (Fig.~\ref{chick_est_venn}).  This is
+supported by ESTs or mRNAs from Genbank (Figure~\ref{chick_est_venn}).  This is
 especially surprising since our mRNAseq data is from spleen, and most of the
@@
 set, Gimme and Cufflinks recover approximately the same number of splice
-junctions already known from ENSEMBL (Figure ~\ref{mus_venn}).  However, Gimme
+junctions already known from ENSEMBL (Figure~\ref{mus_venn}).  However, Gimme
 recovers a substantial number of additional splice variants beyond Cufflinks and
@@
 merged together to form more complete gene structures (e.g. see
-Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
+Figure~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
 splice junctions from ESTs and ENSEMBL respectively
-(Fig.~\ref{combined_venn}), which is about 10\% greater than that from
+(Figure~\ref{combined_venn}), which is about 10\% greater than that from
 corresponding unmerged gene models.
@@
 
-To validate our predicted isoforms, we extracted putative coding sequences from
-our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
-translated 28,772 of 34,800 (82.6\%) of isoforms from Gimme models to protein
-sequences with 50 or more amino acids.  We then searched for homologous
-sequences in mouse ENSEMBL, and found that 22,460 (77.7\%) and 22,991 (79.9\%)
-of isoforms from Cufflinks and Gimme match mouse proteins at a bit-score/length
-$\ge1.0$(Fig.~\ref{bitscore}).  The number of distinct genes are 16,731 and
-12,945 from Cufflinks and Gimme respectively.  A bit-score/length ratio greater
-than 1 indicates that approximately more than half of the protein sequence
-matches mouse proteins well. The sizes of genes and isoforms from Gimme are
-slightly greater than those from Cufflinks while the bitscore/length ratios are
-slightly lower suggesting that transcripts from assembly are more complete.
+To validate our predicted isoforms, we extracted putative coding
+sequences from our gene models with ESTScan\cite{Iseli:1999vd}.
+ESTScan successfully translated 28,772 of 34,800 (82.6\%) of
+isoforms from Gimme models and 28,918 of 38,307 (75.5\%) of
+isoforms from Cufflinks to protein sequences with 50 or more
+amino acids.  We then searched for homologous sequences in mouse
+ENSEMBL, and found that 22,991 (79.9\%) and 22,460 (77.6\%) of
+isoforms from Gimme and Cufflinks match mouse proteins at a
+bit-score/length $\ge1.0$ (Figure~\ref{bitscore}A).  The number of
+distinct genes are 12,945 and 16,731 from Gimme and Cufflinks
+respectively.  A bit-score/length ratio greater than 1.0 indicates
+that approximately more than half of the protein sequence matches
+mouse proteins well.
+
+\subsubsection*{Merged gene models have more complete
+transcripts}
+
+The size of genes and isoforms from Gimme are slightly greater
+than those from Cufflinks (Figure~\ref{bitscore}B) while the
+bit-score/length ratios are slightly lower. To validate that
+Gimme software can be used to combine different sets of gene
+models to create an improved set of gene models, we used our
+software to combine Gimme and Cufflinks models to create merged
+gene models and found that the size of transcripts from merged
+gene models are greater than those from either Cufflinks or Gimme
+models.  Total of 38,535 of 45,793 (84.2\%) isoforms were
+successfully translated to protein sequences with 50 or more
+amino acids and 31,765 (82.4\%) from 13,958 genes matched mouse
+proteins. A majority of genes and isoforms also have a
+bit-score/length ratio greater than 1.0.  This suggests that
+merging Cufflinks and Gimme models creates more complete gene
+models.
 
@@
 that Tophat's default parameters may have been tailored to maximize the finding
-of isoforms in ENSEMBL models (Fig.~\ref{mus_venn}), which could lead to
+of isoforms in ENSEMBL models (Figure~\ref{mus_venn}), which could lead to
 restricted detection of novel splice variants.
@@
 the genome were assembled by chromosome in the local assembly
-(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
+(Figure~\ref{local_assembly}).  Global and local assembly was performed using
 Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
@@
 $5\prime$ and the $3\prime$ ends, some exons may be smaller than others due to
-incomplete assembly (supp. Fig.??). Only the largest exons are kept to ensure
-that the transcript has the most complete exons (Fig.~\ref{algorithm}).
+incomplete assembly (supp.Fig.??). Only the largest exons are kept to ensure
+that the transcript has the most complete exons (Figure~\ref{algorithm}).
 Transcripts were then grouped into the same gene if they have at least one
@@
 \caption{
-{\bf Cumulative counts of splice junctions with spliced reads from both single-
-and paired-end data from Cufflinks and Gimme models.}
+{\bf Percentages of reads mapped to the gene models from Ensembl
+and Gimme.}
+Reads were mapped to transcripts derived from the gene models
+using Bowtie.
 }
@@
     {\bf Splice junctions supported by ESTs or mRNAs.}
+    Both Cufflinks and Gimme detect unique splice junctions
+    supported by ESTs + mRNAs. A large number of splice junctions
+    in ESTs + mRNAs are not found in either Gimme or Cufflinks
+    because ESTs + mRNAs contain transcripts for other tissues.
 }
@@
 \caption{
-{\bf Splice sites in chicken Ensembl gene models detected by Cufflinks and the
-{\em de novo} assembly pipeline.} Cufflinks detects many annotated isoforms
-that are not detected by the pipeline.  The figure also shows that both methods
-detect a large number of unannotated splice junctions.}
+{\bf Splice sites in chicken Ensembl gene models detected by
+Cufflinks and Gimme}
+Cufflinks detects many annotated isoforms that are not detected
+by the pipeline.  The figure also shows that both methods detect
+a large number of unannotated splice junctions.}
 \label{chick_venn}
@@
 \caption{
-{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de
-novo} assembly pipeline.} In contrast to chicken datasets, 85,160 (96.8\%) of
-splice junctions found by Cufflinks are annotated in mouse ENSEMBL models. The
-higher percentage of ENSEMBL splice junctions found by Cufflinks may be a
-result of more complete ENSEMBL gene models. While Gimme can detect the similar
-number of ENSEMBL splice junctions, it detects many more non-ENSEMBL splice
+{\bf Splice sites in mouse Ensembl gene models, Cufflinks and
+Gimme pipeline.}
+In contrast to chicken datasets, 85,160 (96.8\%) of splice
+junctions found by Cufflinks are annotated in mouse ENSEMBL
+models. While Gimme can detect the similar number of ENSEMBL
+splice junctions, it detects many more non-ENSEMBL splice
 junctions in mouse.}
@@
 \caption{
-    {\bf Splice junctions found in merged models.} Merged models (B) find more
-    splice junctions in ENSEMBL and ESTs + mRNAs than Cufflinks models only (A).
-}
+{\bf Splice junctions found in merged models.} Merged models (B)
+find more splice junctions in ENSEMBL and ESTs + mRNAs than
+Cufflinks models only (A).}
 \label{combined_venn}

### 2014-10-02 paper.tex
added some citations

@@
 has been used successfully in many studies to gain unprecedented
-insight into the complexity of transcriptomes (@cite).
+insight into the complexity of transcriptomes~\cite{nagalakshmi2008transcriptional,mortazavi2008mapping,pan2008deep,Wang:2008ea}.
 It has been estimated that, in human, $92-94$\% of multiexon genes
@@
 reads are split into two parts and each part is mapped to the genome
-independently (@cite).  A splice junction is then identified based on
+independently.  A splice junction is then identified based on
 alignments of each half of a read that falls between two exons at
@@
 incomplete nature of the current chicken genome assembly, which is lacking
-approximately 5\% of its true gene content (@cite).  Interestingly, over a
-hundred regions were present {\em only} in the local assemblies,
-suggesting that the local assemblies might be recovering additional exons.
-Significant numbers of unique regions from both global and local assemblies
-showed homology to the mouse genome, indicating that at least some of these
-unique sequences represented real sequence content.
+approximately 5\% of its true gene content~\cite{hillier2004sequence}.
+Interestingly, over a hundred regions were present {\em only} in the local
+assemblies, suggesting that the local assemblies might be recovering additional
+exons.  Significant numbers of unique regions from both global and local
+assemblies showed homology to the mouse genome, indicating that at least some
+of these unique sequences represented real sequence content.
 Figure~\ref{global_vs_local} shows an example of different isoforms detected by
@@
 highly erroneous or contained low-complexity artifactual sequence that probably
-originated from sample processing and reverse transcription (@cite).  Thus the
-merged gene models produced by Gimme represent the significant majority of the
-assemblable data.
+originated from sample processing and reverse
+transcription~\cite{martin2011next}.  Thus the merged gene models produced by
+Gimme represent the significant majority of the assemblable data.
 
@@
 pipeline that incorporates de novo gene prediction and homology-based
-approaches as well as expression data (@cite).
+approaches as well as expression data~\cite{flicek2013ensembl}.
 

### 2015-01-22 2014-gimme.tex
renamed to 2014-gimme

--- 2014-09-30 paper.tex
@@
 
+% @LP Maybe I should add bitscores from Cufflinks so that we can compare with
+% those from Gimme. If the results are similar, we can say that extending
+% Cufflinks with assembly does not affect gene structures in a bad way.
 To validate our predicted isoforms, we extracted putative coding sequences from
 our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
-translated 28,772 of 34,800 (82.6\%) of isoforms from Gimme models to protein
-sequences with 50 or more amino acids.  We then searched for homologous
-sequences in mouse ENSEMBL, and found that 22,460 (77.7\%) and 22,991 (79.9\%)
-of isoforms from Cufflinks and Gimme match mouse proteins at a bit-score/length
-$\ge1.0$(Fig.~\ref{bitscore}).  The number of distinct genes are 16,731 and
-12,945 from Cufflinks and Gimme respectively.  A bit-score/length ratio greater
-than 1 indicates that approximately more than half of the protein sequence
-matches mouse proteins well. The sizes of genes and isoforms from Gimme are
-slightly greater than those from Cufflinks while the bitscore/length ratios are
-slightly lower suggesting that transcripts from assembly are more complete.
-
-% @CTB for discussion: what's surprising about all of this? so many splice
-% variants, etc?  Note we need to go over pickrell paper carefully.  @LP I
-% discuss about it below. Hope it helps.
+translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
+50 or more amino acids.  We then searched for homologous sequences in mouse
+ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
+genes match mouse proteins at a bit-score/length $\ge1.0$
+(Fig.~\ref{bitscore}).  A bit-score/length ratio greater than 1 indicates that
+approximately more than half of the protein sequence matches mouse proteins
+well. The results suggest that merging gene models from Cufflinks and {\em de
+novo} aseembly did not disrupt the structure of protein sequences.
+
+% @CTB for discussion: what's surprising about all of this? so many
+% splice variants, etc?  Note we need to go over pickrell paper carefully.
+% @LP I discuss about it below. Hope it helps.
 
@@
 
-\subsection*{Mapping Reads to The Genome and Gene Models}
+\subsection*{Mapping reads to The Genome and Gene Models}
 
@@
 
-Figure~\ref{img:pipeline} depicts an overall gene model construction
+Figure~\ref{overall_pipeline} depicts an overall gene model construction
 pipeline.  Transcripts of all datasets from local and global assembly were
@@
 \section*{Figure Legends}
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{overall_pipeline.pdf}
+\end{center}
+\caption{
+{\bf Gene model construction pipeline.} Transcripts are obtained from two
+assembly methods -- global and local assembly.  Transcripts are aligned to the
+chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
+of transcripts.  }
+\label{overall_pipeline}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{local_assembly}
+\end{center}
+\caption{
+{\bf Local Assembly Pipeline.}
+Reads are first mapped to a chicken genome.  Then only mapped reads are
+assembled by Velvet and Oases.
+}
+\label{local_assembly}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=5in]{algorithm.pdf}
+\end{center}
+\caption{
+{\bf Intron and exon graphs.} Each intron connects to exons whose splice
+junctions match it boundary.  Some exons are excluded from the final gene model
+if they are incomplete (exon 3a,b).  Introns sharing at least one exon are
+grouped together.  Then an exon graph is made using exons as nodes.
+}
+\label{algorithm}
+\end{figure}
 
@@
 
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{pipeline.pdf}
-\end{center}
-\caption{
-{\bf Gene model construction pipeline.}
-Transcripts are obtained from two assembly methods -- global and
-local assembly. Transcripts with low complexity are removed and
-poly(A) tails are trimmed by SeqClean. Transcripts are aligned to
-the chicken genome using BLAT and alignments are then used by
-Gimme to construct gene models. Gene models from Cufflinks and
-other sources can also be incorporated to build the gene models.
-}
-\label{img:pipeline}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{local_assembly.pdf}
-\end{center}
-\caption{
-{\bf Local assembly pipeline.}
-Reads are first mapped to a chicken genome by Tophat.  Then reads
-mapped to each chromosome are assembled by Velvet and Oases
-separately.
-}
-\label{local_assembly}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=5in]{algorithm.pdf}
-\end{center}
-\caption{
-{\bf Intron and exon graph.}
-Introns sharing at least one exon are grouped together, then an
-exon graph is constructed with exons from all introns as nodes.
-Exons are excluded from the final gene model if they are smaller
-than other exons sharing the same intron (exon 1a, 3a and 3b).
-Edges of the exon graph (red arrow) are implied from the
-connections of introns (blue arrow).  A transcript is constructed
-from concatenation of all unique exons in the graph.
-}
-\label{algorithm}
-\end{figure}
 \end{document}
